### 2017-03-17 Tex Style Files/check.tex
First Commit

%% Usage: Latex check.tex
%%
%% Utility to test if all the files required for wspc review volume project
%% are available in your present LaTeX path.
%% - ws-rv9x6.cls
%% - ws-crop9x6.sty
%% - crop.sty
%% - rotating.sty
%% - url.sty
%% - makeidx.sty
%% - chapterbib.sty
%% - cite.sty
%% - ws-minitoc.sty
%% - ws-index.sty

\newif\onerror

\typeout{*}
\typeout{1. Searching for LaTeX2e ... }
\ifx\documentclass\undefined
 \typeout{*}
 \typeout{* Fatal error occurred:}
 \typeout{*}
 \typeout{* The WS Review Volume class can only be used with LaTeX2e}
 \typeout{*}
 \typeout{* Stopping...}
 \expandafter\stop
\else
 \typeout{   ... ok}
\fi

\def\next#1/#2/#3\next{#1#2}
\typeout{ }
\typeout{2.  Searching for LaTeX2e version ... }
\ifnum\expandafter\next\fmtversion\next<199612 \relax
 \typeout{*}
 \typeout{* Fatal error occurred:}
 \typeout{*}
 \typeout{* Your LaTeX version is quite outdated, \fmtversion.}
 \typeout{* The WS Review Volume class can only be used with}
 \typeout{* a recent version of LaTeX2e, please upgrade your installation!}
 \typeout{*}
 \expandafter\stop
\else
 \ifnum\expandafter\next\fmtversion\next<199806 \relax
   \typeout{* Warning: }
   \typeout{*}
   \typeout{* Your LaTeX version is outdated, \fmtversion.}
   \typeout{* We believe that it will probably work, but if you encounter}
   \typeout{* problems you will need upgrade your installation.}
   \typeout{*}
   \typein{* Type <return> to continue ...}
   \errortrue
 \else
   \typeout{   ... ok}
 \fi
\fi

\typeout{ }
\typeout{3. Searching for ws-rv9x6.cls ... }
\IfFileExists{ws-rv9x6.cls}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{* ... ws-rv9x6.cls, file not found! }
     \typeout{*}
     \typeout{* Fatal error occurred:}
     \typeout{*}
     \typeout{* The ws-rv9x6 class file which is part of the WS Review Volume distribution}
     \typeout{* must be installed in a directory which is searched by LaTeX.}
     \typeout{*}
     \typeout{* Stopping...}
     \typeout{*}
     \stop
}

\typeout{ }
\typeout{4. Searching for ws-crop9x6.sty ... }
\IfFileExists{ws-crop9x6.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... ws-crop9x6.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The ws-crop9x6 file which is part of the WS Books distribution}
     \typeout{* must be installed in a directory which is searched by LaTeX.}
     \typeout{*}
     \typeout{* Without this package you will not be able to typeset }
     \typeout{* your book in 9x6 trim size.}
     \typeout{*}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{5. Searching for crop.sty ... }
\IfFileExists{crop.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... crop.sty, file not found! }
     \typeout{*}
     \typeout{*}
     \typeout{* The crop package, which should be part of a good LaTeX}
     \typeout{* distribution, can not be found.}
     \typeout{*}
     \typeout{* This package is required for printing the crop marks.}
     \typeout{* Try to download this package from a CTAN host.}
     \typeout{* Go to http://www.ctan.org/search.html and search for "crop".}
     \typeout{*}
     \typeout{* Stopping...}
     \typeout{*}
     \stop
}

\typeout{ }
\typeout{6. Searching for rotating.sty ... }
\IfFileExists{rotating.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... rotating.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The rotating package, which should be part of a good LaTeX}
     \typeout{* distribution, can not be found.}
     \typeout{*}
     \typeout{* Without this package you will not be able to use}
     \typeout{* sidewaysfigure and sidewaystable environments.}
     \typeout{* Try to download this package from a CTAN host.}
     \typeout{* Go to http://www.ctan.org/search.html and search for "rotating".}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{7. Searching for url.sty ... }
\IfFileExists{url.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... url.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The url package, which should be part of a good LaTeX}
     \typeout{* distribution, can not be found.}
     \typeout{*}
     \typeout{* Without this package you will not be able to use the \string\url}
     \typeout{* command. Try to download this package from a CTAN host.}
     \typeout{* Go to http://www.ctan.org/search.html and search for "url".}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{8. Searching for makeidx.sty ... }
\IfFileExists{makeidx.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... makeidx.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The makeidx package, which should be part of a good LaTeX}
     \typeout{* distribution, can not be found.}
     \typeout{*}
     \typeout{* Without this package you will not be able to generate index.}
     \typeout{* Try to download this package from a CTAN host.}
     \typeout{* Go to http://www.ctan.org/search.html and search for "makeidx".}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{9. Searching for chapterbib.sty ... }
\IfFileExists{chapterbib.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... chapterbib.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The chapterbib package, which should be part of a good LaTeX}
     \typeout{* distribution, can not be found.}
     \typeout{*}
     \typeout{* Without this package you will not be able to include separate bibliography for each chapter.}
     \typeout{* Try to download this package from a CTAN host.}
     \typeout{* Go to http://www.ctan.org/search.html and search for "chapterbib".}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{10. Searching for cite.sty ... }
\IfFileExists{cite.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... cite.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The cite package, which should be part of a good LaTeX}
     \typeout{* distribution, can not be found.}
     \typeout{*}
     \typeout{* Without this package you will not be able to generate citations in superscript.}
     \typeout{* Try to download this package from a CTAN host.}
     \typeout{* Go to http://www.ctan.org/search.html and search for "cite".}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{11. Searching for ws-minitoc.sty ... }
\IfFileExists{ws-minitoc.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... ws-minitoc.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The ws-minitoc package which is part of the WS Review Volume distribution}
     \typeout{* must be installed in a directory which is searched by LaTeX.}
     \typeout{*}
     \typeout{* You can use the ws-rv9x6 class without this package but }
     \typeout{* you can not have chapter level tocs.}
     \typeout{*}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{12. Searching for ws-rv-har.bst ... }
\IfFileExists{ws-rv-har.bst}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... ws-rv-har.bst, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The ws-rv-har bibliography style file which is part of the WS Review Volume distribution}
     \typeout{* must be installed in a directory which is searched by LaTeX.}
     \typeout{*}
     \typeout{* You can use the ws-rv9x6 class without this package but }
     \typeout{* you can not use BiBTeX features.}
     \typeout{*}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{13. Searching for ws-rv-van.bst ... }
\IfFileExists{ws-rv-van.bst}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... ws-rv-van.bst, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The ws-rv-van bibliography style file which is part of the WS Review Volume distribution}
     \typeout{* must be installed in a directory which is searched by LaTeX.}
     \typeout{*}
     \typeout{* You can use the ws-rv9x6 class without this package but }
     \typeout{* you can not use BiBTeX features.}
     \typeout{*}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{14. Searching for ws-index.sty ... }
\IfFileExists{ws-index.sty}
    {
     \typeout{  ... ok}
    }
    {
     \typeout{  ... ws-index.sty, file not found! }
     \typeout{ }
     \typeout{* Warning:}
     \typeout{*}
     \typeout{* The ws-index package which is part of the WS RV distribution}
     \typeout{* must be installed in a directory which is searched by LaTeX.}
     \typeout{*}
     \typeout{* You can use the ws-rv9x6 class without this package but }
     \typeout{* you can not use multiple indexes.}
     \typeout{*}
     \typeout{*}
     \typein{* Type <return> to continue ...}
     \errortrue
    }

\typeout{ }
\typeout{* Finished Testing *}
\typeout{ }
\onerror
\typeout{* Some of the files/packages required by WS was not found,}
\typeout{*}
\typeout{* Please review the above comments carefully and update your system.}
\typeout{*}
\typeout{*****************************************************************}
\typein{* Type <return> to continue ...} \else
\typeout{******************************************************}
\typeout{*}
\typeout{* All the files and packages required by WS was found,}
\typeout{* you may now proceed LaTeXing your files.}
\typeout{*}
\typeout{******************************************************} \fi
\stop
\endinput

### 2017-04-19 Paper Drafts/Draft 0/Draft0.tex
Draft 0: Intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Review Volume (last updated on 20-4-2015)                            %%
%% Trim Size: 9in x 6in                                                 %%
%% Text Area: 7.35in (include runningheads) x 4.5in                     %%
%% Main Text: 10 on 13pt                                                %%
%% For support: Yolande Koh, <ykoh@wspc.com.sg>                         %%
%%              D. Rajesh Babu, <rajesh@wspc.com.sg>                    %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
\documentclass{ws-rv9x6}
\usepackage{subfigure}   % required only when side-by-side / subfigures are used
\usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
\usepackage{ws-rv-van}   % numbered citation & references (default)
%\usepackage{ws-index}   % to produce multiple indexes
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}       % author index
%\renewindex{default}{idx}{ind}{Subject Index}  % subject index

\begin{document}

\chapter[Ancestral Reconstruction of Bat Echolocation Calls]{Ancestral Reconstruction of Bat Echolocation Calls}\label{ra_ch1}

\author[J.P. Meagher et al.]{J.P. Meagher\footnote{Author footnote.}}
%\index[aindx]{Author, F.} % or \aindx{Author, F.}
%\index[aindx]{Author, S.} % or \aindx{Author, S.}

\address{Department of Statistics,\\
University of Warwick, \\
J.Meagher@Warwick.ac.uk\footnote{Affiliation footnote.}}

\begin{abstract}
The stated aim of the Statistical Data Science workshop jointly organised by the Department of Mathematics and Data Science Institute at Imperial College London, and Winton Global investment management, is ``exploring the nature of the relationship between statistics and data science''. 
\end{abstract}
%\markright{Customized Running Head for Odd Page} % default is Chapter Title.
\body

%\tableofcontents

\section{Introduction}

\subsection{Motivation}
Data science, the `` study of the generalisable extraction of knowledge from data'' \cite{dhar2013data} is motivated by availability of interesting datasets. One field producing such datasets is Bioacoustics, where data is collected, often through citizen science \cite{kullenberg2016citizen} initiatives, for monitoring and conservation purposes.\cite{allen2006citizen} Bats (order \textit{Chiroptera}) have been of particular interest.\cite{pettorelli2013indicator}

Bats form the second most speciose order of mammals, behind rodents, with over 1200 species. \cite{simmons2005order} Bats have been identified as ideal bioindicators for monitoring climate change and habitat quality.\cite{jones2009carpe} Good bioindicators are easy to identify and sample, well distributed geographically, respond to changes in habitat in a manner correlated with other taxa, and have a well understood natural history.\cite{moreno2007shortcuts} Bats have the potential to satisfy these characteristics excellently.

Bats are, for the most part, nocturnal, flying echolocators.\cite{kunz1994bats} Echolocation refers to the bats use of, typically ultrasonic\cite{corso1963bone}, calls and their echoes to forage and navigate in the night sky.\cite{griffin1944echolocation} Thus, bats leak information about themselves into the environment, allowing acoustic monitoring.\cite{pettorelli2013indicator}. Work towards the development of automatic acoustic monitoring algorithms for bats \cite{stathopoulos2017bat} \cite{walters2012continental} is ongoing. Similar algorithms have been developed for insects\cite{chesmore2004automated} and birds\cite{briggs2012acoustic}, reflecting the level of interest in bioacoustic monitoring.

The growing database of bat call recordings \cite{collen2012evolution} also opens up other avenues for research. Bat echolocation calls are diverse and can generally be sorted into categories according to the duration, bandwidth and use of harmonics in the call.\cite{maltby20104} It has been observed that closely related species have similar call structures indicating that some of the variation is due to a shared evolutionary history.\cite{jones2006evolution} Comparative analysis of bat echolocation call parameters for ancestral reconstruction\cite{joy2016ancestral} have been performed.\cite{collen2012evolution} However, an approach based Statistical Data Science principles may shed further light on the evolution of echolocation in bats, without relying heavily on domain specific knowledge.

\subsection{Literature review}

Echolocation, or biosonar, is a process whereby sound is produced and the echoes that return from objects are used to perceive the environment.\cite{jones2005echolocation} Animals that use echolocation include bats, toothed whales, some birds, and some shrews and rats.\cite{jones2005echolocation} Bat echolocation calls, ranging from 9 to 212 kHz, differ from those of other echolocators in that they are often laryngeal, complex, and of a relatively long duration.\cite{thomas2004echolocation} The evolutionary dynamics that resulted in bats being nocturnal, flying echolocators have resulted in a wide variety of call structures across species.\cite{jones2006evolution} Echolocation in bats demonstrates convergent evolution, and bat species have been arranged in `guilds' based on habitat and foraging strategy, with similar call structures observed within a `guild'. \cite{aldridge1987morphology} \cite{neuweiler1990auditory} \cite{schnitzler2001echolocation} However, it has been acknowledged that ``echolocation signals reflect a phylogenetically determined basic call structure shaped by specific ecological conditions.''\cite{schnitzler2004evolution} Ancestral Reconstruction is a key aspect of understanding the variation is call structure that is due to phylogenetic relationships between species.

Ancestral Reconstruction involves the extrapolation back in time from measured characteristics of current populations to the ancestral state, the estimate of the same characteristic in common ancestors.\cite{joy2016ancestral} Effective ancestral reconstruction relies on an accurate model for evolution, that is, both the dynamic evolutionary process and the phylogenetic relationships between species. \cite{joy2016ancestral} There have been maximum parsimony\cite{fitch1971toward}, maximum likelihood\cite{pupko2000fast}, and Bayesian\cite{pagel2004bayesian} approaches taken to ancestral reconstruction.

Other studies have considered bat echolocation calls for ancestral reconstruction. \cite{fenton1995signal} \cite{collen2012evolution}. Fenton\cite{fenton1995signal} hypothesised that early bats used short, broadband clicks for echolocation based on the observation that this is the method of echolocating employed by all animals except bats. Collen\cite{collen2012evolution} proposed instead that early bats used short, multi-harmonic, narrowband laryngeal calls for echolocation, based on a comparative analysis of echolocation call parameters. An approach to the reconstruction of early bat echolocation calls which is not so selective in the information it employs may shed further light on this topic. Gaussian Process Regression on Phylogenies for function valued traits provide a promising path to explore.\cite{jones2013evolutionary} Jones \& Moriarty proposed an extension to Gaussian Process Regression\cite{rasmussen2006gaussian} which allowed the generalisation of Ornstein-Uhlenbeck\cite{uhlenbeck1930theory} models of continuous-time character evolution for traits considered as functional data objects,\cite{ramsay2006functional} through a phylogeny. This study represents an effort to apply these methods to the ancestral reconstruction of bat echolocation calls.

\section{Methods}

\subsection{Recordings}

Describe the Data. Get started with notation. Worth looking at vassilos stuff here.

\subsection{Power spectral Density}

Define PSD, get references, Fourier transform of the data.

\subsection{Functional Data Objects}

Justification of PSD as a functional data object, smoothing, regularisation etc.

\subsection{Component Analysis}

PCA and ICA execution. Tuning of components.

\subsection{Phylogenetic Gaussian Processes}

Gaussian Processes, Matern / OU kernel, interpretation of parameters

proceedings

\section{Results}

\subsection{Simulation Study}

given hypers, what happens along the given tree

\subsection{Mexican Bat Echolocation call Dataset}

What happens when the method is applied to the Mexican bat calls?

\section{Discussion}

\subsection{What do we learn from this?}

Phylogenetic / Non-phylogenetic Signal. Not just noise. Although speculative and probably limited going back to root may be useful for more recent ancestors.

\subsection{Future Work in this Area}

Implement for Spectrograms, produce some recordings.

\bibliographystyle{ws-rv-van}
\bibliography{../BatBiblio}

%\blankpage
%\printindex[aindx]                 % to print author index
%\printindex                         % to print subject index

\end{document}

### 2017-04-19 Paper Drafts/Draft 0/Draft0.tex
Spectrogram figure and Data Description

@@
 %%
-%\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
-\documentclass{ws-rv9x6}
+\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
+%\documentclass{ws-rv9x6}
 \usepackage{subfigure}   % required only when side-by-side / subfigures are used
@@
 %\tableofcontents
-
 \section{Introduction}
@@
 
+\begin{figure}[ht]
+	\centerline{
+		\subfigure[Broadband sweep, multiharmonic - \textit{Antrozous pallidus}]
+		{\includegraphics[width=1.5in]{../Figures/anpaSpectrogram.eps}\label{fig::anpa}}
+		\hspace*{4pt}
+		\subfigure[Narrowband, multiharmonic - \textit{Balantiopteryx plicata
+			}]
+		{\includegraphics[width=1.5in]{../Figures/baplSpectrogram.eps}\label{fig::bapl}}
+	}
+	\centerline{
+		\subfigure[Broadband sweep, single harmonic - \textit{Myotis yumanensis}]
+		{\includegraphics[width=1.5in]{../Figures/myyuSpectrogram.eps}\label{fig::myyu}}
+		\hspace*{4pt}
+		\subfigure[Constant frequency with broadband sweep - \textit{Pteronotus parnellii}]
+		{\includegraphics[width=1.5in]{../Figures/ptpaSpectrogram.eps}\label{fig::ptpa}}
+	}
+	\caption{Representative call spectrograms for some bat species, demonstrating echolocation call structures.}\label{fig::spectrogram}
+\end{figure}
+
 \section{Methods}
+
+\subsection{Data Description}
+
+The post processed echolocation call data accompanying Stathopoulos et al.
+\cite{stathopoulos2017bat} was used in this analysis.
+Echolocation calls were recorded across north and central Mexico with a Pettersson 1000x bat detector (Pettersson Elektronik AB, Uppsala, Sweden).
+Live trapped bats were measured and identified to species level using field keys.\cite{ceballos2005mamiferos} \cite{medellin2sanchez} Bats were recorded either while released from the hand or while tied to a zip line.
+The bat detector was set to record calls manually in real time, full spectrum at 500 kHz.
+In total the dataset consists of 22 species in five families, 449 individual bats and 1814 individual echolocation call recordings.
+
+\begin{table}[ht]
+	\tbl{Echolocation Call Dataset Statistics}
+	{\begin{tabular}{@{}lcc@{}} \toprule
+			Species & Individuals & Calls \\ \colrule
+			Family: Emballonuridae & & \\
+			1 \textit{Balantiopteryx plicata} & 16 & 100 \\
+			\\
+			Family: Molossidae && \\
+			2 \textit{Nyctinomops femorosaccus} & 16 & 100 \\
+			3 \textit{Tadarida brasiliensis} & 49 & 100  \\
+			\\
+			Family: Vespertilionidae && \\
+			4 \textit{Antrozous pallidus} & 58 & 100 \\
+			5 \textit{Eptesicus fuscus} & 74 & 100 \\
+			6 \textit{Idionycteris phyllotis} & 6 & 100 \\
+			7 \textit{Lasiurus blossevillii} & 10 & 90 \\
+			8 \textit{Lasiurus cinereus} & 5 & 42 \\
+			9 \textit{Lasiurus xanthinus} & 8 & 100 \\
+			10 \textit{Myotis volans} & 8 & 100 \\
+			11 \textit{Myotis yumanensis} & 5 & 89 \\
+			12 \textit{Pipistrellus hesperus} & 85 & 100 \\
+			\\
+			Family: Mormoopidae && \\
+			13 \textit{Mormoops megalophylla} & 10 & 100 \\
+			14 \textit{Pteronotus davyi} & 8 & 100 \\
+			15 \textit{Pteronotus parnellii} & 23 & 100 \\
+			16 \textit{Pteronotus personatus} & 7 & 51 \\
+			\\
+			Family:Phyllostomidae && \\
+			17 \textit{Artibeus jamaicensis} & 11 & 82 \\
+			18 \textit{Desmodus rotundus} & 6 & 38 \\
+			19 \textit{Leptonycteris yerbabuenae} & 26 & 100 \\
+			20 \textit{Macrotus californicus} & 6 & 53 \\
+			21 \textit{Sturnira ludovici} & 8 & 51 \\
+			22 \textit{Sturnira lilium} & 4 & 20 \\
+			\botrule
+		\end{tabular}
+	}
+	\begin{tabnote}
+		$^{\text a}$Sample table footnote.
+	\end{tabnote}
+	\label{tab::dataset}
+\end{table}
+
+The 
+The echolocation call dataset used for this study is the
 

### 2017-04-19 R Code/Figures/Spectrograms.R

@@ -1,0 +1,52 @@
+## SDS: Figure 1
+library(ancestralbat)
+library(signal)
+
+# Get example spectrograms
+w <- 512
+anpa <- specgram(mexicancalls$chirp1[[2]], w, Fs = 500000, window = hamming(w),
+                 overlap = 15*(w/16))
+
+bapl <- specgram(mexicancalls$chirp1[[72]], w, Fs = 500000, window = hamming(w),
+                 overlap = 15*(w/16))
+
+myyu <- specgram(mexicancalls$chirp1[[248]], w, Fs = 500000, window = hamming(w),
+                 overlap = 15*(w/16))
+
+ptpa <- specgram(mexicancalls$chirp1[[366]], w, Fs = 500000, window = hamming(w),
+                 overlap = 15*(w/16))
+
+# Produce Image for each and save as pdf
+((23 - max(bapl$t*1000) )/ 2)
+
+x11()
+anpa_s <- 10*log10(abs(t(anpa$S^2)))
+anpa_s[anpa_s < 0] <- 0
+image(anpa$t*1000, anpa$f/1000, anpa_s, col = grey(512:0 /512), ylim = c(9, 212), 
+      xlim = c(0 - ((23 - max(anpa$t*1000) )/ 2), max(anpa$t*1000) + ((23 - max(anpa$t*1000) )/ 2)),
+      xlab = "Time (ms)", ylab = "Frequency (kHz)", cex.lab = 1.5, cex.axis = 1.5)
+box()
+
+x11()
+bapl_s <- 10*log10(abs(t(bapl$S^2)))
+bapl_s[bapl_s < 0] <- 0
+image(bapl$t*1000, bapl$f/1000, bapl_s, col = grey(512:0 /512), ylim = c(9, 212), 
+      xlim = c(0 - ((23 - max(bapl$t*1000) )/ 2), max(bapl$t*1000) + ((23 - max(bapl$t*1000) )/ 2)),
+      xlab = "Time (ms)", ylab = "Frequency (kHz)", cex.lab = 1.5, cex.axis = 1.5)
+box()
+
+x11()
+myyu_s <- 10*log10(abs(t(myyu$S^2)))
+myyu_s[myyu_s < 0] <- 0
+image(myyu$t*1000, myyu$f/1000, myyu_s, col = grey(512:0 /512), ylim = c(9, 212), 
[…]

### 2017-04-20 Paper Drafts/Draft 0/Draft0.tex
Added evolutionary tree figure

@@
 
-\subsection{Literature review}
-
-Echolocation, or biosonar, is a process whereby sound is produced and the echoes that return from objects are used to perceive the environment.\cite{jones2005echolocation} Animals that use echolocation include bats, toothed whales, some birds, and some shrews and rats.\cite{jones2005echolocation} Bat echolocation calls, ranging from 9 to 212 kHz, differ from those of other echolocators in that they are often laryngeal, complex, and of a relatively long duration.\cite{thomas2004echolocation} The evolutionary dynamics that resulted in bats being nocturnal, flying echolocators have resulted in a wide variety of call structures across species.\cite{jones2006evolution} Echolocation in bats demonstrates convergent evolution, and bat species have been arranged in `guilds' based on habitat and foraging strategy, with similar call structures observed within a `guild'.
-\cite{aldridge1987morphology} \cite{neuweiler1990auditory} \cite{schnitzler2001echolocation} However, it has been acknowledged that ``echolocation signals reflect a phylogenetically determined basic call structure shaped by specific ecological conditions.''\cite{schnitzler2004evolution} Ancestral Reconstruction is a key aspect of understanding the variation is call structure that is due to phylogenetic relationships between species.
-
-Ancestral Reconstruction involves the extrapolation back in time from measured characteristics of current populations to the ancestral state, the estimate of the same characteristic in common ancestors.\cite{joy2016ancestral} Effective ancestral reconstruction relies on an accurate model for evolution, that is, both the dynamic evolutionary process and the phylogenetic relationships between species.
-\cite{joy2016ancestral} There have been maximum parsimony\cite{fitch1971toward}, maximum likelihood\cite{pupko2000fast}, and Bayesian\cite{pagel2004bayesian} approaches taken to ancestral reconstruction.
-
-Other studies have considered bat echolocation calls for ancestral reconstruction.
-\cite{fenton1995signal} \cite{collen2012evolution}.
-Fenton\cite{fenton1995signal} hypothesised that early bats used short, broadband clicks for echolocation based on the observation that this is the method of echolocating employed by all animals except bats.
-Collen\cite{collen2012evolution} proposed instead that early bats used short, multi-harmonic, narrowband laryngeal calls for echolocation, based on a comparative analysis of echolocation call parameters.
-An approach to the reconstruction of early bat echolocation calls which is not so selective in the information it employs may shed further light on this topic.
-Gaussian Process Regression on Phylogenies for function valued traits provide a promising path to explore.\cite{jones2013evolutionary} Jones \& Moriarty proposed an extension to Gaussian Process Regression\cite{rasmussen2006gaussian} which allowed the generalisation of Ornstein-Uhlenbeck\cite{uhlenbeck1930theory} models of continuous-time character evolution for traits considered as functional data objects,\cite{ramsay2006functional} through a phylogeny.
-This study represents an effort to apply these methods to the ancestral reconstruction of bat echolocation calls.
-
 \begin{figure}[ht]
@@
 
+\subsection{Literature review}
+
+Echolocation, or biosonar, is a process whereby sound is produced and the echoes that return from objects are used to perceive the environment.\cite{jones2005echolocation} Animals that use echolocation include bats, toothed whales, some birds, and some shrews and rats.\cite{jones2005echolocation} Bat echolocation calls, ranging from 9 to 212 kHz, differ from those of other echolocators in that they are often laryngeal, complex, and of a relatively long duration.\cite{thomas2004echolocation} The evolutionary dynamics that resulted in bats being nocturnal, flying echolocators have resulted in a wide variety of call structures across species.\cite{jones2006evolution} Echolocation in bats demonstrates convergent evolution, and bat species have been arranged in `guilds' based on habitat and foraging strategy, with similar call structures observed within a `guild'.
+\cite{aldridge1987morphology} \cite{neuweiler1990auditory} \cite{schnitzler2001echolocation} However, it has been acknowledged that ``echolocation signals reflect a phylogenetically determined basic call structure shaped by specific ecological conditions.''\cite{schnitzler2004evolution} Ancestral Reconstruction is a key aspect of understanding the variation is call structure that is due to phylogenetic relationships between species.
+
+Ancestral Reconstruction involves the extrapolation back in time from measured characteristics of current populations to the ancestral state, the estimate of the same characteristic in common ancestors.\cite{joy2016ancestral} Effective ancestral reconstruction relies on an accurate model for evolution, that is, both the dynamic evolutionary process and the phylogenetic relationships between species.
+\cite{joy2016ancestral} There have been maximum parsimony\cite{fitch1971toward}, maximum likelihood\cite{pupko2000fast}, and Bayesian\cite{pagel2004bayesian} approaches taken to ancestral reconstruction.
+
+Other studies have considered bat echolocation calls for ancestral reconstruction.
+\cite{fenton1995signal} \cite{collen2012evolution}.
+Fenton\cite{fenton1995signal} hypothesised that early bats used short, broadband clicks for echolocation based on the observation that this is the method of echolocating employed by all animals except bats.
+Collen\cite{collen2012evolution} proposed instead that early bats used short, multi-harmonic, narrowband laryngeal calls for echolocation, based on a comparative analysis of echolocation call parameters.
+An approach to the reconstruction of early bat echolocation calls which is not so selective in the information it employs may shed further light on this topic.
+Gaussian Process Regression on Phylogenies for function valued traits provide a promising path to explore.\cite{jones2013evolutionary} Jones \& Moriarty proposed an extension to Gaussian Process Regression\cite{rasmussen2006gaussian} which allowed the generalisation of Ornstein-Uhlenbeck\cite{uhlenbeck1930theory} models of continuous-time character evolution for traits considered as functional data objects,\cite{ramsay2006functional} through a phylogeny.
+This study represents an effort to apply these methods to the ancestral reconstruction of bat echolocation calls.
+
 \section{Methods}
@@
 The bat detector was set to record calls manually in real time, full spectrum at 500 kHz.
-In total the dataset consists of 22 species in five families, 449 individual bats and 1814 individual echolocation call recordings.
+In total the dataset consists of 22 species in five families, 449 individual bats and 1816 individual echolocation call recordings.
 
@@
 	\tbl{Echolocation Call Dataset Statistics}
-	{\begin{tabular}{@{}lcc@{}} \toprule
-			Species & Individuals & Calls \\ \colrule
-			Family: Emballonuridae & & \\
-			1 \textit{Balantiopteryx plicata} & 16 & 100 \\
+	{\begin{tabular}{@{}lccc@{}} \toprule
+			Species & Key & Individuals & Calls \\ \colrule
+			Family: Emballonuridae &&& \\
+			1 \textit{Balantiopteryx plicata} & Bapl & 16 & 100 \\
 			\\
-			Family: Molossidae && \\
-			2 \textit{Nyctinomops femorosaccus} & 16 & 100 \\
-			3 \textit{Tadarida brasiliensis} & 49 & 100  \\
+			Family: Molossidae &&& \\
+			2 \textit{Nyctinomops femorosaccus} & Nyfe & 16 & 100 \\
+			3 \textit{Tadarida brasiliensis} & Tabr & 49 & 100  \\
 			\\
-			Family: Vespertilionidae && \\
-			4 \textit{Antrozous pallidus} & 58 & 100 \\
-			5 \textit{Eptesicus fuscus} & 74 & 100 \\
-			6 \textit{Idionycteris phyllotis} & 6 & 100 \\
-			7 \textit{Lasiurus blossevillii} & 10 & 90 \\
-			8 \textit{Lasiurus cinereus} & 5 & 42 \\
-			9 \textit{Lasiurus xanthinus} & 8 & 100 \\
-			10 \textit{Myotis volans} & 8 & 100 \\
-			11 \textit{Myotis yumanensis} & 5 & 89 \\
-			12 \textit{Pipistrellus hesperus} & 85 & 100 \\
+			Family: Vespertilionidae &&& \\
+			4 \textit{Antrozous pallidus} & Anpa & 58 & 100 \\
+			5 \textit{Eptesicus fuscus} & Epfu & 74 & 100 \\
+			6 \textit{Idionycteris phyllotis} & Idph & 6 & 100 \\
+			7 \textit{Lasiurus blossevillii} & Labl & 10 & 90 \\
+			8 \textit{Lasiurus cinereus} & Laci & 5 & 42 \\
+			9 \textit{Lasiurus xanthinus} & Laxa & 8 & 100 \\
+			10 \textit{Myotis volans} & Myvo & 8 & 100 \\
+			11 \textit{Myotis yumanensis} & Myyu & 5 & 89 \\
+			12 \textit{Pipistrellus hesperus} & Pihe & 85 & 100 \\
 			\\
-			Family: Mormoopidae && \\
-			13 \textit{Mormoops megalophylla} & 10 & 100 \\
-			14 \textit{Pteronotus davyi} & 8 & 100 \\
-			15 \textit{Pteronotus parnellii} & 23 & 100 \\
-			16 \textit{Pteronotus personatus} & 7 & 51 \\
+			Family: Mormoopidae &&& \\
+			13 \textit{Mormoops megalophylla} & Mome & 10 & 100 \\
+			14 \textit{Pteronotus davyi} & Ptda & 8 & 100 \\
+			15 \textit{Pteronotus parnellii} & Ptpa & 23 & 100 \\
+			16 \textit{Pteronotus personatus} & Ptpe & 7 & 51 \\
 			\\
-			Family:Phyllostomidae && \\
-			17 \textit{Artibeus jamaicensis} & 11 & 82 \\
-			18 \textit{Desmodus rotundus} & 6 & 38 \\
-			19 \textit{Leptonycteris yerbabuenae} & 26 & 100 \\
-			20 \textit{Macrotus californicus} & 6 & 53 \\
-			21 \textit{Sturnira ludovici} & 8 & 51 \\
-			22 \textit{Sturnira lilium} & 4 & 20 \\
+			Family:Phyllostomidae &&& \\
+			17 \textit{Artibeus jamaicensis} & Arja & 11 & 82 \\
+			18 \textit{Desmodus rotundus} & Dero & 6 & 38 \\
+			19 \textit{Leptonycteris yerbabuenae} & Leye & 26 & 100 \\
+			20 \textit{Macrotus californicus} & Maca & 6 & 53 \\
+			21 \textit{Sturnira ludovici} & Stlu & 8 & 51 \\
+			22 \textit{Sturnira lilium} & Stli & 4 & 20 \\
 			\botrule
@@
 	}
-	\begin{tabnote}
-		$^{\text a}$Sample table footnote.
-	\end{tabnote}
 	\label{tab::dataset}
@@
 
-The 
-The echolocation call dataset used for this study is the
+Phylogenetic relationships between species are described and dated according to the Bat super-tree produced by Collen.\cite{collen2012evolution}
 
+\begin{figure}
+	\centering
+	\includegraphics[width= 0.9\textwidth]{../Figures/Phylogeny.eps}
+\end{figure}
 \subsection{Recordings}

### 2017-04-20 R Code/Figures/Phylogeny.R

@@ -1,0 +1,7 @@
+library(ancestralbat)
+library(ape)
+
+x11(height = 5)
+plot.phylo(mexicantree)
+axisPhylo()
+title(xlab = 'Time (millions of years)')

### 2017-04-24 Paper Drafts/Draft 0/Draft0.tex
function Valued traits Update

@@
 
+\subsection{The evolution of function valued traits}
+
+Bat echolocation calls can be described as 'function valued' traits.\cite{meyer2005up} In this context, 'function valued' refers to traits measured along a continuous scale, which can then be represented by a continuous mathematical function.
+Data of this nature can be viewed as functional data objects.\cite{ramsay2006functional} The Functional Phylogenies Group\cite{group2012phylogenetic} argue the case for performing evolutionary inference, including ancestral reconstruction, on function valued traits, with a particular focus on the evolution of linguistic speech sounds.
+They propose performing inference on a functional representation of the speech sound itself, by extending Gaussian Process Regression\cite{rasmussen2006gaussian} to take advantage of the tree structure of phylogenetic relationships.\cite{jones2013evolutionary}.
+
+Phylogenetic Gaussian Process Regression (PGPR), as developed by Jones \& Moriarty,\cite{jones2013evolutionary} extends Ornstein-Uhlenbeck\cite{uhlenbeck1930theory} models for continuous-time character evolution to function valued traits.
+In comparative studies of genetic data the O-U model is used to describe stabilising selection with random genetic drift, the notion that traits have some optimal value, around which they vary.\cite{hansen1997stabilizing} The PGPR is based on a kernel with 3 hyperparameters.
+Phylogenetic noise and length scale hyperparameters govern the correlation between nodes on a tree due to the phylogeny, while non-phylogenetic noise accounts for other sources of variation.
+Estimating these hyperparameters allows posterior distributions for traits at internal nodes to be constructed.
+In this way estimates for the ancestral state can be made.
+This model does not require that variance of the trait is homogenous across the whole function space.
+By defining a set of deterministic basis functions, combinations of which can be used to describe the observed trait functions, the variance of traits can be modelled to vary over the function space.
+An implementation of this method on simulated data is given by Hajipantelis.\cite{hadjipantelis2013function}
+
+\subsection{Paper Description}
+
+In a similar manner to the representation of speech sounds as spectrograms,\cite{pigoli2015analysis} bat echolocation calls will be represented by their Energy Spectral Density\cite{antoniou2006digital} (ESD).
+The ESD, the energy of a signal as a function of a frequency, provides a straightforward characterisation from which a PGPR can be implemented.
+
+Using a synthetic dataset based on the bat phylogeny the conditions for an effective implementation of PGPR will be demonstrated.
+
+PGPR will then be implemented for the set of bat calls and posterior distributions of ESD estimated.
+
+Spectral representations of the calls will be described, the implementation of PGPR.
+Independent Component Analysis.
+Simulations study to demonstrate the PGPR can work for this dataset.
+Application to Bats.
+
 \section{Methods}
@@
 \begin{figure}
-	\centering
+	\centering	q
 	\includegraphics[width= 0.9\textwidth]{../Figures/Phylogeny.eps}
 \end{figure}
-\subsection{Recordings}
 
-Describe the Data. Get started with notation. Worth looking at vassilos stuff here.
+\subsection{From Echolocation Call recordings to Spectral Curves}
 
-\subsection{Power spectral Density}
+Echolocation calls can be considered to be continuous, non-periodic, pulse signals which are a function of time \(x(t)\).
 
-Define PSD, get references, Fourier transform of the data.
+The Fourier Transform of a signal is an expression of the signal in the frequency domain, that is, as a function of frequency. It is given by
 
-\subsection{Functional Data Objects}
+\[
+X(f) = \int_{-\infty}^{\infty} x(t) e^{-2\pi ft} dt.
+\]
 
-Justification of PSD as a functional data object, smoothing, regularisation etc.
+The energy of the signal is given by
+
+\[E = \int_{-\infty}^{\infty} |x(t)|^2 dt = \int_{-\infty}^{\infty} |X(f)|^2 df,\]
+
+by Parseval's Theorem.
+
+The ESD is then \(|X(f)|^2\).
+This is a representation of a sound and so it is appropriate to scale this to a decibel scale.
+Not also that \(|X(f)|^2\) is a periodic function and symmetric around 0.
+Given that bat echolocation calls range from \(9 - 212\)kHz we are interested in \(S(f) = log_{10} |X(f)|^2, f \in [9, 212] \).
+This can then be treated as a functional data object.
+
+By treating \(S(f)\) as a functional data object it is treated as a set of noisy realisations taken of a smooth underlying curve.
+Estimating this underlying curve is performed by smoothing.
+Another aspect of functional data analysis is registration, in this case it would consist of separating out phase and amplitude variation.
+As the energy at specific frequencies is of interest here, registration along the frequency axis is not appropriate.
+However the level of the amplitude depends on factors such as the distance of the bat from the microphone while being recorded.
+This variation is not interesting.
+Thus registration on the amplitude axis is appropriate.
+thus all curves are registered such that the amplitude lies on the interval \([0,1]\).
+
+Once smoothed and registered the curves are sampled from at every kHz. These Spectral curves are the functions to be analysed in the comparative analysis.
+
+\subsection{Phylogenetic Gaussian Process Regression for spatially inhomogenous data}
+
+Some stuff about Gaussian Process Regression
+
+Extension of Gaussian Processes to Phylogenetic Data
+
+Extension to spatially inhomogenous data.
+
+The theoretical results presented by Jones \& Moriarty\cite{jones2013evolutionary} rely on two assumptions about the data
 

### 2017-04-25 Paper Drafts/Draft 0/Draft0.tex
Phylogenetic Gaussian Processes Update

@@
 
-\subsection{Phylogenetic Gaussian Process Regression for spatially inhomogenous data}
+\subsection{Phylogenetic Gaussian Processes}
+
+Given a dataset, \(X\), of \(N\) observations of an input vector \(\mathbf{x}\) associated with a vector of response variables \(\mathbf{y}\), a Gaussian Process model assumes that the relationship between \(\mathbf{x}\) and \(y\) is governed by a multivariate Gaussian distribution.
+Formally, a Gaussian Process is a collection of random variables, any finite number of which have a joint Gaussian distribution.\cite{rasmussen2006gaussian}
+
+In a function-space view of Gaussian Processes, a Gaussian Process \(f(\cdot)\) is completely characterised by its mean and covariance functions.
+For simplicity, the mean function can be set such that \(m(\mathbf{x}) = \mathbf{E} [f(\mathbf{x})] = 0\).
+The covariance function, defined by a kernel \(k(\mathbf{x}, \mathbf{x}') = \mathbf{E}[(f(\mathbf{x}) - m(\mathbf{x}))(f(\mathbf{x'}) - m(\mathbf{x'}))] \), describes the relationship between coordinates within the input space.
+Thus, setting \(y_i = f(\mathbf{x}_i)\), models the vector of response variables as a multivariate Gaussian distribution, \(\mathbf{y} \sim \mathcal{N}(0, K)\), where \(K = K(X, X)\) is the covariance matrix given by \(K_{ij} = K(X, X)_{ij} = k(\mathbf{x}_i, \mathbf{x}_j)\).
+
+Once a Gaussian Process has been fitted for the observed data, a conditional distribution for unobserved points, \(X_*\), on the input space can be estimated.
+The key predictive equations for Gaussian Process Regression are
+\[
+\mathbf{f}_* | X, \mathbf{y}, X_* \sim \mathcal{N}(\bar{\mathbf{f}_*}, cov(\mathbf{f}_*))
+\]
+where
+\[
+\bar{\mathbf{f}_*} \equiv \mathbf{E} [\mathbf{f}_* | X, \mathbf{y}, X_*] = K(X_*, X) (K)^{-1} \mathbf{y}
+\]
+and
+\[
+cov(\mathbf{f}_*) = K(X_*, K_*) - K(X_*, X) (K)^{-1} K(X, X_*).
+\]
+
+Thus the key aspect of Gaussian Process Regression is the definition of an appropriate kernel, which is context dependant.
+
+The Gaussian Process framework provides a flexible general approach to regression problems, a framework which Jones \& Moriarty extended to phylogenteic inference for function valued traits.
+Suppose the phylogenetic tree, \(\mathbf{T}\), denotes the relationships between the observed data in evolutionary time.
+The trait itself exists in some function space \(\mathcal{X}\) and so each observation is indexed by \((\mathbf{x}, \mathbf{t})\), where \(\mathbf{x}\) describes the observed functional trait, and \(\mathbf{t}\) the position of the observation in \(\mathbf{T}\), evolutionary time.
+A Phylogenetic Gaussian process can then be defined by a suitable kernel.
+The form of this kernel is given by Jones \& Moriarty.\cite{jones2013evolutionary}
+
+The results are presented for a Phylogenetic Gaussian Process subject to a number of simplifying assumptions.
+It is assumed that conditional on common ancestors in \(\mathbf{T}\) any two traits are independent.
+It is also assumed that marginal process of evolution, that is the path on \(\mathbf{T}\) from the root to any tip, is identical along all paths.
+Finally, it is also assumed that the marginal process is separable by evolutionary time and the function valued trait space.
+
+In order to fit a Phylogenetic Gaussian Process model with spatially inhomogenous variance, a set of basis functions must be defined.
+Intuitively, these basis functions should define `components' of the trait, each of which are subject to an independent univariate Phylogenetic Gaussian Process, which can be additively combined to model the trait at various points in evolutionary time.
+The cubICA\cite{blaschke2004cubica} implementation of Independent Components Analysis offers a method of estimating such a set of basis functions, referred to as components hereafter.
+
+given by the Ornstein-Uhelenbeck process,
 

### 2017-05-02 Paper Drafts/Draft 0/Draft0.tex
Simulation Study

@@
 
-given by the Ornstein-Uhelenbeck process,
+The implementation here models the Phylogenetic Gaussian process with an Ornstein-Uhlenbeck kernel, a special case of the Matern class of kernels, with additive gaussian noise.
 
-Some stuff about Gaussian Process Regression
-
-Extension of Gaussian Processes to Phylogenetic Data
-
-Extension to spatially inhomogenous data.
-
-The theoretical results presented by Jones \& Moriarty\cite{jones2013evolutionary} rely on two assumptions about the data
-
-\subsection{Component Analysis}
-
-PCA and ICA execution. Tuning of components.
-
-\subsection{Phylogenetic Gaussian Processes}
-
-Gaussian Processes, Matern / OU kernel, interpretation of parameters
-
-proceedings
+\[
+k(x, x) = phylogenetic + nonphylogenetic
+\]
+In this way the PGP can account for both  phylogenetic and non-phylogenetic aspects of the evolutionary process.
 
@@
 \subsection{Simulation Study}
+
+a simulation study examining PGP for a random tree was performed for a random tree with 128 tips \cite{hadjipantelis2013function}.
+However the phylogenetic tree presented here is very different from that, having only 22 tips.
+In order to understand how effectively hyperparameters can be estimated for this OU process a simulation study of my own was performed.
+
+Each observation is associated with some tip on the phylogenetic tree, however we are not limited to a single observation at each tip.
+we are interested in the variation between species and so we can place multiple observations at each tip.
+Care was taken to ensure the dataset was balanced in some sense and so the same number of observations were taken from each tip.
+
+The object of this simulation study is to demonstrate the conditions required for type II maximum likelihood estimation to successfully estimate hyperparameters for a Phylogenetic Ornstein-Uhlenbeck Process.
+
+The Phylogenetic Ornsetin-Uhlenbeck Process kernel is
+\[
+k(x,x') = \sigma_p \exp \left( \frac{|x - x'|}{\ell} \right)  + \sigma_n
+\]
+where \(\sigma_p\) is the phylogenetic noise, \(\ell\) the phylogenetic length-scale, and \(sigma_n\) the non-phylogenetic noise.
+
+The key parameter here is \(\ell\).
+It can be easily shown that when \(\ell >> |x - x'|\) there is a high degree of correlation between process observations at \(x\) and \(x'\).
+When \(\ell << |x - x'|\) there is little correlation and the process becomes indistinguishable from white noise.
+With this in mind fix \(\sigma_p = 1\) and \(\sigma_n = 1\).
+
+Investigate the accuracy for the process over the tree for \(\ell = \{10, 25, 50, 100, 200\}\) considering \(n = \{1, 2, 4, 8\}\) observations at each tip of the tree.
+
+Create a dataframe storing the results of 1000 simulations for the vlaues of \(\ell, n \) of interest.
+
+In order to demonstrate the accuracy of type 2 MLE estimation  the mean estimatior(sample variance) of each hyperparameter is given below.
+
+\begin{table}[ht]
+	\tbl{Echolocation Call Dataset Statistics}
+	{\begin{tabular}{@{}rccccc@{}} \toprule
+			& \multicolumn{5}{c}{\(\ell\)} \\
+			& 10 & 20 & 50 & 100 & 200 \\ 
+			\colrule
+			\(n = 1\)  &&&& \\
+			\(2\)  &&&& \\
+			\(4\)  &&&& \\
+			\(8\)  &&&& \\
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::simulation}
+\end{table}
+
+More observations better than less.
+
+Exponential distribution of length scale.
+
+Some pictures, pairs plot maybe. Compare different sample sizes, distributions. Note plots cannot use colour.
 

### 2017-05-02 R Code/Simulation Study/writeUp.Rmd
Simulation Study

--- 2017-04-24 Paper Drafts/Draft 0/Draft0.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Review Volume (last updated on 20-4-2015)                            %%
-%% Trim Size: 9in x 6in                                                 %%
-%% Text Area: 7.35in (include runningheads) x 4.5in                     %%
-%% Main Text: 10 on 13pt                                                %%
-%% For support: Yolande Koh, <ykoh@wspc.com.sg>                         %%
-%%              D. Rajesh Babu, <rajesh@wspc.com.sg>                    %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%
-\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
-%\documentclass{ws-rv9x6}
-\usepackage{subfigure}   % required only when side-by-side / subfigures are used
-\usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
-\usepackage{ws-rv-van}   % numbered citation & references (default)
-%\usepackage{ws-index}   % to produce multiple indexes
-\makeindex
-%\newindex{aindx}{adx}{and}{Author Index}       % author index
-%\renewindex{default}{idx}{ind}{Subject Index}  % subject index
+---
+title: "Simulation Study for Statistical Data Science Conference"
+author: "J.P. Meagher"
+date: "1 May 2017"
+output: html_document
+---
 
-\begin{document}
+The object of this simulation study is to demonstrate the conditions required for type II maximum likelihood estimation to successfully estimate hyperparameters for a Phylogenetic Ornstein-Uhlenbeck Process.
 
-\chapter[Ancestral Reconstruction of Bat Echolocation Calls]{Ancestral Reconstruction of Bat Echolocation Calls}\label{ra_ch1}
+The Phylogenetic Ornsetin-Uhlenbeck Process kernel is
+\[
+k(x,x') = \sigma_p \exp \left( \frac{|x - x'|}{\ell} \right)  + \sigma_n
+\]
+where \(\sigma_p\) is the phylogenetic noise, \(\ell\) the phylogenetic length-scale, and \(sigma_n\) the non-phylogenetic noise.
 
-\author[J.P. Meagher et al.]{J.P. Meagher\footnote{Author footnote.}}
-%\index[aindx]{Author, F.} % or \aindx{Author, F.}
-%\index[aindx]{Author, S.} % or \aindx{Author, S.}
+The key parameter here is \(\ell\).
+It can be easily shown that when \(\ell >> |x - x'|\) there is a high degree of correlation between process observations at \(x\) and \(x'\).
+When \(\ell << |x - x'|\) there is little correlation and the process becomes indistinguishable from white noise.
+With this in mind fix \(\sigma_p = 1\) and \(\sigma_n = 1\).
 
-\address{Department of Statistics,\\
-University of Warwick, \\
-J.Meagher@Warwick.ac.uk\footnote{Affiliation footnote.}}
+Investigate the accuracy for the process over the tree for \(\ell = \{10, 25, 50, 100, 200\}\) considering \(n = \{1, 2, 4, 8\}\) observations at each tip of the tree
 
-\begin{abstract}
-The stated aim of the Statistical Data Science workshop jointly organised by the Department of Mathematics and Data Science Institute at Imperial College London, and Winton Global investment management, is ``exploring the nature of the relationship between statistics and data science''.
-\end{abstract}
-%\markright{Customized Running Head for Odd Page} % default is Chapter Title.
-\body
+```{r}
+library(sdsBAT)
 
-%\tableofcontents
-\section{Introduction}
+n <- 1
+l <- 10
 
-\subsection{Motivation}
-Data science, the `` study of the generalisable extraction of knowledge from data'' \cite{dhar2013data} is motivated by availability of interesting datasets.
-One field producing such datasets is Bioacoustics, where data is collected, often through citizen science \cite{kullenberg2016citizen} initiatives, for monitoring and conservation purposes.\cite{allen2006citizen} Bats (order \textit{Chiroptera}) have been of particular interest.\cite{pettorelli2013indicator}
+tree <- expand_tree(phylogeny, n)
+sim_10_1 <- pou_simulation_study(log(c(1,l,1)), tree, 100, logl_function = pou_logl_fast, optim_function = 'uobyqa', upper_initialisation = c(2, 150, 2), n_restarts = 5)
 
-Bats form the second most speciose order of mammals, behind rodents, with over 1200 species.
-\cite{simmons2005order} Bats have been identified as ideal bioindicators for monitoring climate change and habitat quality.\cite{jones2009carpe} Good bioindicators are easy to identify and sample, well distributed geographically, respond to changes in habitat in a manner correlated with other taxa, and have a well understood natural history.\cite{moreno2007shortcuts} Bats have the potential to satisfy these characteristics excellently.
+pou_simulation_study
 
-Bats are, for the most part, nocturnal, flying echolocators.\cite{kunz1994bats} Echolocation refers to the bats use of, typically ultrasonic\cite{corso1963bone}, calls and their echoes to forage and navigate in the night sky.\cite{griffin1944echolocation} Thus, bats leak information about themselves into the environment, allowing acoustic monitoring.\cite{pettorelli2013indicator}.
-Work towards the development of automatic acoustic monitoring algorithms for bats \cite{stathopoulos2017bat} \cite{walters2012continental} is ongoing.
-Similar algorithms have been developed for insects\cite{chesmore2004automated} and birds\cite{briggs2012acoustic}, reflecting the level of interest in bioacoustic monitoring.
+pou_simulation_study(log(c(1,10,1)), phylogeny, 10, n_restarts = 5)
 
-The growing database of bat call recordings \cite{collen2012evolution} also opens up other avenues for research.
-Bat echolocation calls are diverse and can generally be sorted into categories according to the duration, bandwidth and use of harmonics in the call.\cite{maltby20104} It has been observed that closely related species have similar call structures indicating that some of the variation is due to a shared evolutionary history.\cite{jones2006evolution} Comparative analysis of bat echolocation call parameters for ancestral reconstruction\cite{joy2016ancestral} have been performed.\cite{collen2012evolution} However, an approach based Statistical Data Science principles may shed further light on the evolution of echolocation in bats, without relying heavily on domain specific knowledge.
-
-\begin{figure}[ht]
-	\centerline{
-		\subfigure[Broadband sweep, multiharmonic - \textit{Antrozous pallidus}]
-		{\includegraphics[width=1.5in]{../Figures/anpaSpectrogram.eps}\label{fig::anpa}}
-		\hspace*{4pt}
-		\subfigure[Narrowband, multiharmonic - \textit{Balantiopteryx plicata
-		}]
-		{\includegraphics[width=1.5in]{../Figures/baplSpectrogram.eps}\label{fig::bapl}}
-	}
-	\centerline{
-		\subfigure[Broadband sweep, single harmonic - \textit{Myotis yumanensis}]
-		{\includegraphics[width=1.5in]{../Figures/myyuSpectrogram.eps}\label{fig::myyu}}
-		\hspace*{4pt}
-		\subfigure[Constant frequency with broadband sweep - \textit{Pteronotus parnellii}]
-		{\includegraphics[width=1.5in]{../Figures/ptpaSpectrogram.eps}\label{fig::ptpa}}
-	}
-	\caption{Representative call spectrograms for some bat species, demonstrating echolocation call structures.}\label{fig::spectrogram}
-\end{figure}
-
-\subsection{Literature review}
-
-Echolocation, or biosonar, is a process whereby sound is produced and the echoes that return from objects are used to perceive the environment.\cite{jones2005echolocation} Animals that use echolocation include bats, toothed whales, some birds, and some shrews and rats.\cite{jones2005echolocation} Bat echolocation calls, ranging from 9 to 212 kHz, differ from those of other echolocators in that they are often laryngeal, complex, and of a relatively long duration.\cite{thomas2004echolocation} The evolutionary dynamics that resulted in bats being nocturnal, flying echolocators have resulted in a wide variety of call structures across species.\cite{jones2006evolution} Echolocation in bats demonstrates convergent evolution, and bat species have been arranged in `guilds' based on habitat and foraging strategy, with similar call structures observed within a `guild'.
-\cite{aldridge1987morphology} \cite{neuweiler1990auditory} \cite{schnitzler2001echolocation} However, it has been acknowledged that ``echolocation signals reflect a phylogenetically determined basic call structure shaped by specific ecological conditions.''\cite{schnitzler2004evolution} Ancestral Reconstruction is a key aspect of understanding the variation is call structure that is due to phylogenetic relationships between species.
-
-Ancestral Reconstruction involves the extrapolation back in time from measured characteristics of current populations to the ancestral state, the estimate of the same characteristic in common ancestors.\cite{joy2016ancestral} Effective ancestral reconstruction relies on an accurate model for evolution, that is, both the dynamic evolutionary process and the phylogenetic relationships between species.
-\cite{joy2016ancestral} There have been maximum parsimony\cite{fitch1971toward}, maximum likelihood\cite{pupko2000fast}, and Bayesian\cite{pagel2004bayesian} approaches taken to ancestral reconstruction.
-
-Other studies have considered bat echolocation calls for ancestral reconstruction.
-\cite{fenton1995signal} \cite{collen2012evolution}.
-Fenton\cite{fenton1995signal} hypothesised that early bats used short, broadband clicks for echolocation based on the observation that this is the method of echolocating employed by all animals except bats.
-Collen\cite{collen2012evolution} proposed instead that early bats used short, multi-harmonic, narrowband laryngeal calls for echolocation, based on a comparative analysis of echolocation call parameters.
-An approach to the reconstruction of early bat echolocation calls which is not so selective in the information it employs may shed further light on this topic.
-Gaussian Process Regression on Phylogenies for function valued traits provide a promising path to explore.\cite{jones2013evolutionary} Jones \& Moriarty proposed an extension to Gaussian Process Regression\cite{rasmussen2006gaussian} which allowed the generalisation of Ornstein-Uhlenbeck\cite{uhlenbeck1930theory} models of continuous-time character evolution for traits considered as functional data objects,\cite{ramsay2006functional} through a phylogeny.
-This study represents an effort to apply these methods to the ancestral reconstruction of bat echolocation calls.
-
-\subsection{The evolution of function valued traits}
-
-Bat echolocation calls can be described as 'function valued' traits.\cite{meyer2005up} In this context, 'function valued' refers to traits measured along a continuous scale, which can then be represented by a continuous mathematical function.
-Data of this nature can be viewed as functional data objects.\cite{ramsay2006functional} The Functional Phylogenies Group\cite{group2012phylogenetic} argue the case for performing evolutionary inference, including ancestral reconstruction, on function valued traits, with a particular focus on the evolution of linguistic speech sounds.
-They propose performing inference on a functional representation of the speech sound itself, by extending Gaussian Process Regression\cite{rasmussen2006gaussian} to take advantage of the tree structure of phylogenetic relationships.\cite{jones2013evolutionary}.
-
-Phylogenetic Gaussian Process Regression (PGPR), as developed by Jones \& Moriarty,\cite{jones2013evolutionary} extends Ornstein-Uhlenbeck\cite{uhlenbeck1930theory} models for continuous-time character evolution to function valued traits.
-In comparative studies of genetic data the O-U model is used to describe stabilising selection with random genetic drift, the notion that traits have some optimal value, around which they vary.\cite{hansen1997stabilizing} The PGPR is based on a kernel with 3 hyperparameters.
-Phylogenetic noise and length scale hyperparameters govern the correlation between nodes on a tree due to the phylogeny, while non-phylogenetic noise accounts for other sources of variation.
-Estimating these hyperparameters allows posterior distributions for traits at internal nodes to be constructed.
-In this way estimates for the ancestral state can be made.
-This model does not require that variance of the trait is homogenous across the whole function space.
-By defining a set of deterministic basis functions, combinations of which can be used to describe the observed trait functions, the variance of traits can be modelled to vary over the function space.
-An implementation of this method on simulated data is given by Hajipantelis.\cite{hadjipantelis2013function}
-
-\subsection{Paper Description}
-
-In a similar manner to the representation of speech sounds as spectrograms,\cite{pigoli2015analysis} bat echolocation calls will be represented by their Energy Spectral Density\cite{antoniou2006digital} (ESD).
-The ESD, the energy of a signal as a function of a frequency, provides a straightforward characterisation from which a PGPR can be implemented.
-
-Using a synthetic dataset based on the bat phylogeny the conditions for an effective implementation of PGPR will be demonstrated.
-
-PGPR will then be implemented for the set of bat calls and posterior distributions of ESD estimated.
-
-Spectral representations of the calls will be described, the implementation of PGPR.
-Independent Component Analysis.
-Simulations study to demonstrate the PGPR can work for this dataset.
-Application to Bats.
-
-\section{Methods}
-
-\subsection{Data Description}
-
-The post processed echolocation call data accompanying Stathopoulos et al.
-\cite{stathopoulos2017bat} was used in this analysis.
-Echolocation calls were recorded across north and central Mexico with a Pettersson 1000x bat detector (Pettersson Elektronik AB, Uppsala, Sweden).
-Live trapped bats were measured and identified to species level using field keys.\cite{ceballos2005mamiferos} \cite{medellin2sanchez} Bats were recorded either while released from the hand or while tied to a zip line.
-The bat detector was set to record calls manually in real time, full spectrum at 500 kHz.
-In total the dataset consists of 22 species in five families, 449 individual bats and 1816 individual echolocation call recordings.
-
-\begin{table}[ht]
-	\tbl{Echolocation Call Dataset Statistics}
-	{\begin{tabular}{@{}lccc@{}} \toprule
-			Species & Key & Individuals & Calls \\ \colrule
-			Family: Emballonuridae &&& \\
-			1 \textit{Balantiopteryx plicata} & Bapl & 16 & 100 \\
-			\\
-			Family: Molossidae &&& \\
-			2 \textit{Nyctinomops femorosaccus} & Nyfe & 16 & 100 \\
-			3 \textit{Tadarida brasiliensis} & Tabr & 49 & 100  \\
-			\\
-			Family: Vespertilionidae &&& \\
-			4 \textit{Antrozous pallidus} & Anpa & 58 & 100 \\
-			5 \textit{Eptesicus fuscus} & Epfu & 74 & 100 \\
-			6 \textit{Idionycteris phyllotis} & Idph & 6 & 100 \\
-			7 \textit{Lasiurus blossevillii} & Labl & 10 & 90 \\
-			8 \textit{Lasiurus cinereus} & Laci & 5 & 42 \\
-			9 \textit{Lasiurus xanthinus} & Laxa & 8 & 100 \\
-			10 \textit{Myotis volans} & Myvo & 8 & 100 \\
-			11 \textit{Myotis yumanensis} & Myyu & 5 & 89 \\
-			12 \textit{Pipistrellus hesperus} & Pihe & 85 & 100 \\
-			\\
-			Family: Mormoopidae &&& \\
-			13 \textit{Mormoops megalophylla} & Mome & 10 & 100 \\
-			14 \textit{Pteronotus davyi} & Ptda & 8 & 100 \\
-			15 \textit{Pteronotus parnellii} & Ptpa & 23 & 100 \\
-			16 \textit{Pteronotus personatus} & Ptpe & 7 & 51 \\
-			\\
-			Family:Phyllostomidae &&& \\
-			17 \textit{Artibeus jamaicensis} & Arja & 11 & 82 \\
-			18 \textit{Desmodus rotundus} & Dero & 6 & 38 \\
-			19 \textit{Leptonycteris yerbabuenae} & Leye & 26 & 100 \\
-			20 \textit{Macrotus californicus} & Maca & 6 & 53 \\
-			21 \textit{Sturnira ludovici} & Stlu & 8 & 51 \\
-			22 \textit{Sturnira lilium} & Stli & 4 & 20 \\
-			\botrule
-		\end{tabular}
-	}
-	\label{tab::dataset}
-\end{table}
-
-Phylogenetic relationships between species are described and dated according to the Bat super-tree produced by Collen.\cite{collen2012evolution}
-
-\begin{figure}
-	\centering	q
-	\includegraphics[width= 0.9\textwidth]{../Figures/Phylogeny.eps}
-\end{figure}
-
-\subsection{From Echolocation Call recordings to Spectral Curves}
-
-Echolocation calls can be considered to be continuous, non-periodic, pulse signals which are a function of time \(x(t)\).
-
-The Fourier Transform of a signal is an expression of the signal in the frequency domain, that is, as a function of frequency. It is given by
-
-\[
-X(f) = \int_{-\infty}^{\infty} x(t) e^{-2\pi ft} dt.
-\]
-
-The energy of the signal is given by
-
-\[E = \int_{-\infty}^{\infty} |x(t)|^2 dt = \int_{-\infty}^{\infty} |X(f)|^2 df,\]
-
-by Parseval's Theorem.
-
-The ESD is then \(|X(f)|^2\).
-This is a representation of a sound and so it is appropriate to scale this to a decibel scale.
-Not also that \(|X(f)|^2\) is a periodic function and symmetric around 0.
-Given that bat echolocation calls range from \(9 - 212\)kHz we are interested in \(S(f) = log_{10} |X(f)|^2, f \in [9, 212] \).
-This can then be treated as a functional data object.
-
-By treating \(S(f)\) as a functional data object it is treated as a set of noisy realisations taken of a smooth underlying curve.
-Estimating this underlying curve is performed by smoothing.
-Another aspect of functional data analysis is registration, in this case it would consist of separating out phase and amplitude variation.
-As the energy at specific frequencies is of interest here, registration along the frequency axis is not appropriate.
-However the level of the amplitude depends on factors such as the distance of the bat from the microphone while being recorded.
-This variation is not interesting.
-Thus registration on the amplitude axis is appropriate.
-thus all curves are registered such that the amplitude lies on the interval \([0,1]\).
-
-Once smoothed and registered the curves are sampled from at every kHz. These Spectral curves are the functions to be analysed in the comparative analysis.
-
-\subsection{Phylogenetic Gaussian Process Regression for spatially inhomogenous data}
-
-Some stuff about Gaussian Process Regression
-
-Extension of Gaussian Processes to Phylogenetic Data
-
-Extension to spatially inhomogenous data.
-
-The theoretical results presented by Jones \& Moriarty\cite{jones2013evolutionary} rely on two assumptions about the data
-
-\subsection{Component Analysis}
-
-PCA and ICA execution. Tuning of components.
-
-\subsection{Phylogenetic Gaussian Processes}
-
-Gaussian Processes, Matern / OU kernel, interpretation of parameters
-
-proceedings
-
-\section{Results}
-
-\subsection{Simulation Study}
-
-given hypers, what happens along the given tree
-
-\subsection{Mexican Bat Echolocation call Dataset}
-
-What happens when the method is applied to the Mexican bat calls?
-
-\section{Discussion}
-
-\subsection{What do we learn from this?}
-
-Phylogenetic / Non-phylogenetic Signal. Not just noise. Although speculative and probably limited going back to root may be useful for more recent ancestors.
-
-\subsection{Future Work in this Area}
-
-Implement for Spectrograms, produce some recordings.
-
-\bibliographystyle{ws-rv-van}
-\bibliography{../BatBiblio}
-
-%\blankpage
-%\printindex[aindx]                 % to print author index
-%\printindex                         % to print subject index
-
-\end{document}
+```

### 2017-05-02 R Code/Figures/Phylogeny.R

@@ -4,4 +4,4 @@ library(ape)
 x11(height = 5)
 plot.phylo(mexicantree)
 axisPhylo()
-title(xlab = 'Time (millions of years)')
+title(main = list('Mexican Bat Phylogeny', font = 1, cex = 1.5), xlab = 'Time (millions of years)')

### 2017-05-05 R Code/Simulation Study/writeUp.Rmd
Simulated Hyperparameters and Component code

@@
 
-Investigate the accuracy for the process over the tree for \(\ell = \{10, 25, 50, 100, 200\}\) considering \(n = \{1, 2, 4, 8\}\) observations at each tip of the tree
+Investigate the accuracy for the process over the tree for \(\ell = \{10, 25, 50, 100, 200\}\) considering \(n = \{1, 2, 4, 8\}\) observations at each tip of the tree.
 
-```{r}
+Create a dataframe storing the results of 1000 simulations for the vlaues of \(\ell, n \) of interest.
+
+Note that `uobyqa()` offers the fastest optimisation method, `'Nelder-Mead'` and `'BFGS'` are roughly comparable to each other and only slightly slower, while `'CG'` is far slower.
+I will use `uobyqa` with 5 restarts.
+
+5 samples at each valeue take about 23 seconds, should take just over an hour to get the 1000 samples I want.
+
+```{r, cache=TRUE}
 library(sdsBAT)
+set.seed(7)
 
-n <- 1
-l <- 10
+n <- c(1, 2, 4, 8)
+l <- c(10, 25, 50, 100, 200)
+N <- 1000L
 
-tree <- expand_tree(phylogeny, n)
-sim_10_1 <- pou_simulation_study(log(c(1,l,1)), tree, 100, logl_function = pou_logl_fast, optim_function = 'uobyqa', upper_initialisation = c(2, 150, 2), n_restarts = 5)
+dummy <- matrix(1:20, nrow = length(n), ncol = length(l))
+rownames(dummy) <- c('n.1', 'n.2', 'n.4', 'n.8'); colnames(dummy) <- c('l.10', 'l.25', 'l.50', 'l.100', 'l.200')
 
-pou_simulation_study
+simulations <- data.frame(dummy)
+for(i in 1:length(n)){
+  for(j in 1:length(l)){
+    tree <- expand_tree(phylogeny, n[i])
+    experiment <- suppressMessages(pou_simulation_study(log(c(1,l[j],1)), tree, N, logl_function = pou_logl_fast, optim_function = 'uobyqa', upper_initialisation = c(2, 150, 2), n_restarts = 5))
+    simulations[[j]][[i]] <- list(experiment)
+  }
+}
 
-pou_simulation_study(log(c(1,10,1)), phylogeny, 10, n_restarts = 5)
-
+panel.hist <- function(x, ...)
+{
+    usr <- par("usr"); on.exit(par(usr))
+    par(usr = c(usr[1:2], 0, 1.5) )
+    h <- hist(x, plot = FALSE)
+    breaks <- h$breaks; nB <- length(breaks)
+    y <- h$counts; y <- y/max(y)
+    rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
+}
 ```

### 2017-05-05 R Code/Components/Components.R

@@ -1,0 +1,111 @@
+# Set seed for Reproducability.
+set.seed(8)
+
+# Create the samples of calls balanced by species and individual.
+N <- 1000
+resampled_calls <- balanced_samples(rescale_smooth_densities(), n_samples = N)
+
+# Proportion of variance accounted for by each component.
+get_proportion_variance <- function(df_col){
+  single_sample <- matrix(unlist(df_col), nrow = length(df_col), ncol = length(df_col[[1]]), byrow = T)
+  pca <- prcomp(single_sample, scale. = T)
+  proportion_variance <- pca$sdev^2 /(sum(pca$sdev^2))
+  return(proportion_variance)
+}
+
+sample_proportion_variance <- apply(resampled_calls, 2, get_proportion_variance)
+matplot(sample_proportion_variance[1:20, 1:100])
+
+mean_proportion_variance <- apply(sample_proportion_variance, 1, mean)
+plot(mean_proportion_variance)
+mean_proportion_variance[1:10]
+plot(cumsum(mean_proportion_variance))
+
+sum(sample_proportion_variance[9,] > 0.03)
+proportion_mask <- (sample_proportion_variance > 0.03)
+sum(proportion_mask)
+# The centering / mean curves for the samples.
+get_centring_curve <- function(df_col){
+  single_sample <- matrix(unlist(df_col), nrow = length(df_col), ncol = length(df_col[[1]]), byrow = T)
+  pca <- prcomp(single_sample, scale. = T)
+
+  return(pca$center)
+}
+
+sample_centring_curve <- apply(resampled_calls, 2, get_centring_curve)
+matplot(sample_centring_curve[, 1:100], type = 'l')
+mean_centring_curve <- apply(sample_centring_curve, 1, mean)
+plot(mean_centring_curve, type = 'l')
+
+# Get 10 most descriptive components.
+get_components <-function(df_col, n_components = 10){
+  single_sample <- matrix(unlist(df_col), nrow = length(df_col), ncol = length(df_col[[1]]), byrow = T)
+  pca <- prcomp(single_sample, scale. = T)
+
+  return(pca$rotation[, 1:n_components])
+}
+
+examined_components <- 10
+sample_components <- apply(resampled_calls, 2, get_components, n_components = examined components)
+sample_components <- array(sample_components, dim = c(208, 10, N))
+matplot(sample_components[,6,], type = 'l')
+proportion_mask <- proportion_mask[1:10, ]
+sum(proportion_mask)
+
[…]

### 2017-05-08 R Code/Components/Components.R
Datasets

@@ -3,7 +3,7 @@ set.seed(8)
 
 # Create the samples of calls balanced by species and individual.
 N <- 1000
-resampled_calls <- balanced_samples(rescale_smooth_densities(), n_samples = N)
+resampled_calls <- balanced_samples(smooth_and_regularise_call_densities(), n_samples = N)
 
 # Proportion of variance accounted for by each component.
 get_proportion_variance <- function(df_col){
@@ -20,8 +20,9 @@ mean_proportion_variance <- apply(sample_proportion_variance, 1, mean)
 plot(mean_proportion_variance)
 mean_proportion_variance[1:10]
 plot(cumsum(mean_proportion_variance))
+cumsum(mean_proportion_variance)[1:10]
 
-sum(sample_proportion_variance[9,] > 0.03)
+sum(sample_proportion_variance[7,] > 0.03)
 proportion_mask <- (sample_proportion_variance > 0.03)
 sum(proportion_mask)
 # The centering / mean curves for the samples.
@@ -33,9 +34,9 @@ get_centring_curve <- function(df_col){
 }
 
 sample_centring_curve <- apply(resampled_calls, 2, get_centring_curve)
-matplot(sample_centring_curve[, 1:100], type = 'l')
+matplot(sample_centring_curve, type = 'l')
 mean_centring_curve <- apply(sample_centring_curve, 1, mean)
-plot(mean_centring_curve, type = 'l')
+lines(mean_centring_curve, lwd = 3)
 
 # Get 10 most descriptive components.
 get_components <-function(df_col, n_components = 10){
@@ -45,67 +46,97 @@ get_components <-function(df_col, n_components = 10){
   return(pca$rotation[, 1:n_components])
 }
 
-examined_components <- 10
-sample_components <- apply(resampled_calls, 2, get_components, n_components = examined components)
-sample_components <- array(sample_components, dim = c(208, 10, N))
-matplot(sample_components[,6,], type = 'l')
-proportion_mask <- proportion_mask[1:10, ]
+examined_components <- 7
+sample_components <- apply(resampled_calls, 2, get_components, n_components = examined_components)
+sample_components <- array(sample_components, dim = c(208, examined_components, N))
+
+proportion_mask <- proportion_mask[1:examined_components, ]
 sum(proportion_mask)
 
 # Identify component peak frequencies
[…]

### 2017-05-08 Paper Drafts/Draft1/Draft1.tex
First Draft: Introduction

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Review Volume (last updated on 20-4-2015)                            %%
%% Trim Size: 9in x 6in                                                 %%
%% Text Area: 7.35in (include runningheads) x 4.5in                     %%
%% Main Text: 10 on 13pt                                                %%
%% For support: Yolande Koh, <ykoh@wspc.com.sg>                         %%
%%              D. Rajesh Babu, <rajesh@wspc.com.sg>                    %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
%\documentclass{ws-rv9x6}
\usepackage{subfigure}   % required only when side-by-side / subfigures are used
\usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
\usepackage{ws-rv-van}   % numbered citation & references (default)
%\usepackage{ws-index}   % to produce multiple indexes
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}       %author index
%\renewindex{default}{idx}{ind}{Subject Index}  %subject index

\begin{document}

\chapter[Ancestral Reconstruction of Bat Echolocation Calls]{Ancestral Reconstruction of Bat Echolocation Calls}\label{ra_ch1}

\author[J.P. Meagher et al.]{J.P. Meagher\footnote{Author footnote.}}
%\index[aindx]{Author, F.} % or \aindx{Author, F.}
%\index[aindx]{Author, S.} % or \aindx{Author, S.}

\address{Department of Statistics,\\
University of Warwick, \\
J.Meagher@Warwick.ac.uk\footnote{Affiliation footnote.}}

\begin{abstract}

\end{abstract}
%\markright{Customized Running Head for Odd Page} % default is Chapter Title.
\body

%\tableofcontents
\section{Introduction}

Advances in technology allowing the precise quantification and storage of information about the world around us continues to drive the emergence of Data Science as a discipline distinct from both Statistics and Computer Science.

Bioacoustics is one area of research generating vast quantities of data which also captures the imagination of the public, as evidenced by successful citizen science initiatives. \cite{allen2006citizen} \cite{pettorelli2013indicator} Bioacoustic techniques for biodiversity monitoring \cite{stathopoulos2017bat} \cite{damoulas2010bayesian} have the potential to make real policy impacts, particularly with regard to sustainable economic development and nature conservation.

In the acoustic monitoring of biodiversity, bats (order \textit{Chiroptera}) are of particular interest. Bats have been identified as ideal bioindicators for monitoring climate change and habitat quality,\cite{jones2009carpe} largely because bats broadcast information about themselves into their environment in the form of echolocation calls.\cite{jones2005echolocation} The development of automatic acoustic monitoring algorithms for classifying species of bats \cite{stathopoulos2017bat} \cite{walters2012continental} means that large scale, non-invasive monitoring is becoming possible.

While monitoring bat populations provides useful information, understanding the root causes and effects of what is observed requires that the natural history of extant bat species is also well understood. Traits, such as call structure or body size, exhibited by particular bat species are linked to the bats interactions with its environment. \cite{aldridge1987morphology} Existing fossil records are of limited use in inferring the traits exhibited by ancestral bats, particularly with respect to echolocation calls for example. The reconstruction of ancestral traits relies heavily on the comparative analysis\cite{felsenstein2004inferring} of extant bat species. Thus, statistical data science techniques may be particularly useful for inferring the evolutionary dynamics and reconstructing ancestral states of echolocation in bats.

Previous studies of bat echolocation calls for both classification \cite{walters2012continental} and ancestral reconstruction \cite{collen2012evolution} examined features of the call extracted from the spectrogram of the call. These, somewhat arbitrary, call features relied on significant domain knowledge to ensure they were sensibly selected and used. More recently however, general techniques for the classification of acoustic signals have been developed. \cite{stathopoulos2014bat} \cite{damoulas2010bayesian} These methods do not require, but can be augmented by, domain knowledge. General techniques for ancestral reconstruction of function-valued traits, such as speech sounds or echolocation calls, have been proposed \cite{group2012phylogenetic}. The study of bat echolocation calls offers an opportunity to examine the efficacy of these techniques.

A function-valued trait is measured along some continuous scale, usually time, and can then be modelled as a continuous mathematical function using techniques for functional data analysis. \cite{ramsay2006functional} Jones \& Moriarty \cite{jones2013evolutionary} develop a method which extends Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits.

\bibliographystyle{ws-rv-van}
\bibliography{../BatBiblio}

%\blankpage
%\printindex[aindx]                 % to print author index
%\printindex                         % to print subject index

\end{document}

### 2017-05-08 Paper Drafts/Draft1/Draft1.tex
First Draft: Full Introduction

@@
 A function-valued trait is measured along some continuous scale, usually time, and can then be modelled as a continuous mathematical function using techniques for functional data analysis.
-\cite{ramsay2006functional} Jones \& Moriarty \cite{jones2013evolutionary} develop a method which extends Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits.
+\cite{ramsay2006functional} Jones \& Moriarty \cite{jones2013evolutionary} developed a method which extends Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits over a phylogeny.
+The model facilitates the implementation of two popular models for continuous character state evolution,\cite{joy2016ancestral} the Brownian Motion and Ornstein-Uhlenbeck models.\cite{lande1976natural} A full demonstration of ancestral reconstruction for synthetic data using the method was presented by Hajipantelis et al.\cite{hadjipantelis2013function}
+
+This general approach to evolutionary inference for function-valued traits is implemented here for a set of bat echolocation calls.
+Our goal in doing so is twofold.
+These techniques had previously been considered in the context of modelling the evolution of human speech sounds in language.
+\cite{group2012phylogenetic} It is hoped that by applying these methods in the simpler context of the evolution of bat echolocation calls that progress can be made towards resolving methodological problems with these methods.
+For example, what is the most appropriate method of representing the information contained in an acoustic signal?
+
+We are also interested in what specifically these models tell us about bats and the evolutionary dynamics driving the development of echolocation.
+What impact might these results have on our understanding of ancestral bats and their behaviour?
+
+This paper presents the early stages of our research and some preliminary results.
 

### 2017-05-08 Paper Drafts/Draft1/Draft1.tex
First Draft: Funtional Representation started

@@
 \begin{abstract}
-
+ Something something bats something evolution something ancestral reconstruction something magic.
 \end{abstract}
@@
 
+\section{Functional Representation of Bat Echolocation Calls}
+
+A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time.
+\cite{ramsay2006functional} These measurements can be thought of as representing points on a curve varies gradually and continuously.
+In the context of phylogenetics, these functional data objects are called function-valued traits.
+\cite{meyer2005up}
+
+Denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of species \(S\) by \(\{\tilde{x}_{lm}^S(n) : n = 0, \dots, N_{lm}^S - 1\}\).
+Thus \(\{\tilde{x}_{lm}^S(\cdot)\}\) is a noisy realisation of \(x^S(\cdot)\), the echolocation call generating process for species \(S\), observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the process sampling rate in samples per second (Hz).
+
+Call recordings themselves are in fact functional data objects, however modelling the phylogenetic relationships between \(\{\tilde{x}_{lm}^S(\cdot)\}\) and \(\{\tilde{x}_{l'm'}^{S'}(\cdot)\}\) directly implies that the processes are comparable at point \(n\).
+This is not the case for acoustic signals, which are sinusoidal and can vary in time without significantly altering the information carried.
+Thus some alternative functional representation of the signal is required.
+
+The discrete Fourier transform of the signal \(\{\tilde{x}_{l'm'}^{S'}(\cdot)\}\) is given by\[\tilde{X}_{lm}^S(k) = \sum_{n = 0}^{N_{lm}^S - 1} \tilde{x}_{lm}^{S}(n) e^{-i2\pi kn / N_{lm}^S }.\] The energy spectral density of this signal is the magnitude of the Fourier transform and so the log spectral density per second (in decibel) is therefore \[\tilde{\mathcal{E}}_{lm}^S(k) = 10 \log_{10} \left(\tilde{X}_{lm}^S(k)\right) / f_S.\]
+
+Bat echolocation calls can be considered as function-valued traits. A trait which is repeatedly
+
 \bibliographystyle{ws-rv-van}

### 2017-05-09 Paper Drafts/Draft1/Draft1.tex
First Draft: Functional Representation

@@
 Traits, such as call structure or body size, exhibited by particular bat species are linked to the bats interactions with its environment.
-\cite{aldridge1987morphology} Existing fossil records are of limited use in inferring the traits exhibited by ancestral bats, particularly with respect to echolocation calls for example.
+\cite{aldridge1987morphology} Existing fossil records are of limited use in inferring the traits exhibited by ancestral bats, particularly with respect to echolocation calls.
 The reconstruction of ancestral traits relies heavily on the comparative analysis\cite{felsenstein2004inferring} of extant bat species.
@@
 
-Previous studies of bat echolocation calls for both classification \cite{walters2012continental} and ancestral reconstruction \cite{collen2012evolution} examined features of the call extracted from the spectrogram of the call.
+Previous studies of bat echolocation calls for both classification \cite{walters2012continental} and ancestral reconstruction \cite{collen2012evolution} examined features of the call extracted from the call spectrogram.
 These, somewhat arbitrary, call features relied on significant domain knowledge to ensure they were sensibly selected and used.
@@
 These techniques had previously been considered in the context of modelling the evolution of human speech sounds in language.
-\cite{group2012phylogenetic} It is hoped that by applying these methods in the simpler context of the evolution of bat echolocation calls that progress can be made towards resolving methodological problems with these methods.
-For example, what is the most appropriate method of representing the information contained in an acoustic signal?
+\cite{group2012phylogenetic} It is hoped that by applying these methods in the simpler context of the evolution of bat echolocation calls that progress can be made towards resolving methodological problems.
+For example, how do we extend these methods to more realistic models of evolution?
 
@@
 
-A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time.
-\cite{ramsay2006functional} These measurements can be thought of as representing points on a curve varies gradually and continuously.
+A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time. \cite{ramsay2006functional} 
+These measurements can be thought of as representing points on a curve that varies gradually and continuously.
 In the context of phylogenetics, these functional data objects are called function-valued traits.
@@
 
-The discrete Fourier transform of the signal \(\{\tilde{x}_{l'm'}^{S'}(\cdot)\}\) is given by\[\tilde{X}_{lm}^S(k) = \sum_{n = 0}^{N_{lm}^S - 1} \tilde{x}_{lm}^{S}(n) e^{-i2\pi kn / N_{lm}^S }.\] The energy spectral density of this signal is the magnitude of the Fourier transform and so the log spectral density per second (in decibel) is therefore \[\tilde{\mathcal{E}}_{lm}^S(k) = 10 \log_{10} \left(\tilde{X}_{lm}^S(k)\right) / f_S.\]
+The discrete Fourier transform of the signal \(\{\tilde{x}_{l'm'}^{S'}(\cdot)\}\) is given by\[\tilde{X}_{lm}^S(k) = \sum_{n = 0}^{N_{lm}^S - 1} \tilde{x}_{lm}^{S}(n) e^{-i2\pi kn / N_{lm}^S }.\] 
+The energy spectral density of this signal is the magnitude of the Fourier transform and so the log energy spectral density per second (in decibel) is estimated by \[\tilde{\mathcal{E}}_{lm}^S(k) = 10 \log_{10} \left( \frac{|\tilde{X}_{lm}^S(k)|f_s}{N_{lm}^S}\right).\]
+The log energy spectral density estimate for the signal \(\tilde{x}_{lm}^S(\cdot)\), \(\tilde{\mathcal{E}}_{lm}^S(\cdot)\) is now considered to be a noisy estimate of the log energy spectral density for species \(S\), denoted \(\mathcal{E}^S(\cdot)\).
+The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \([0, \frac{f_S}{2}]\).
+\(\tilde{\mathcal{E}}_{lm}^S(\cdot)\) has been mapped to the decibel scale and also scaled according to the length, in time, of \(\tilde{x}_{lm}^S(\cdot)\).\cite{antoniou2006digital} We have now mapped each echolocation call on the same scale and \(\tilde{\mathcal{E}}_{lm}^S(\cdot)\) is now comparable to \(\tilde{\mathcal{E}}_{l'm'}^{S'}(f)\) at frequency \(f\).
+In order for these representations to be considered as function-valued traits however, \(\tilde{\mathcal{E}}_{lm}^S(\cdot)\) must be smoothed such the call representation in the frequency domain varies gradually and continuously.
 
-Bat echolocation calls can be considered as function-valued traits. A trait which is repeatedly
+The smoothed log energy spectral density is estimated by smoothing splines where \[
+\mathcal{E}_{lm}^S(f) = \arg \min_{\mathcal{E}_{lm}^S(\cdot)} \int_{0}^{\frac{f_S}{2}}  \{ \tilde{\mathcal{E}}_{lm}^S(\cdot) - \mathcal{E}_{lm}^S(f) \}^2 df + \lambda \int_{0}^{\frac{f_S}{2}} \{ {\mathcal{E}_{lm}^{S}}'' (f)\}^2 df.
+\]
+The smoothing parameter \(\lambda\) is chosen using a generalised cross-validation procedure.\cite{friedman2001elements} \cite{citeR}
+
+We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \mathcal{E}_{lm}^S(f)\}\) and \(\{f, \mathcal{E}_{l'm'}^(S')(f)\}\) are directly comparable.
+The function-valued traits can now be modelled for evolutionary inference.
 

### 2017-05-09 Paper Drafts/Draft1/Draft1.tex
First Draft: PGP

@@
 The log energy spectral density estimate for the signal \(\tilde{x}_{lm}^S(\cdot)\), \(\tilde{\mathcal{E}}_{lm}^S(\cdot)\) is now considered to be a noisy estimate of the log energy spectral density for species \(S\), denoted \(\mathcal{E}^S(\cdot)\).
-The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \([0, \frac{f_S}{2}]\).
+The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \(F = [0, \frac{f_S}{2}]\).
 \(\tilde{\mathcal{E}}_{lm}^S(\cdot)\) has been mapped to the decibel scale and also scaled according to the length, in time, of \(\tilde{x}_{lm}^S(\cdot)\).\cite{antoniou2006digital} We have now mapped each echolocation call on the same scale and \(\tilde{\mathcal{E}}_{lm}^S(\cdot)\) is now comparable to \(\tilde{\mathcal{E}}_{l'm'}^{S'}(f)\) at frequency \(f\).
@@
 
-We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \mathcal{E}_{lm}^S(f)\}\) and \(\{f, \mathcal{E}_{l'm'}^(S')(f)\}\) are directly comparable.
+We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \mathcal{E}_{lm}^S(f)\}\) and \(\{f, \mathcal{E}_{l'm'}^{S'}(f)\}\) are directly comparable.
 The function-valued traits can now be modelled for evolutionary inference.
+
+\section{Gaussian Process Regression on Phylogenies}
+
+The key innovation of Jones \& Moriarty\cite{jones2013evolutionary} in extending Gaussian Process Regression \cite{rasmussen2006gaussian} for use in evolutionary inference, was to replace the linear measure of distance between observations with a phylogenetic tree, denoted \(\mathbf{T}\).
+When this condition is imposed each of our observations \(\mathcal{E}^{S}(f)\) correspond to a point \((f, \mathbf{t})\) on the frequency-phylogeny \(F \times \mathbf{T}\).
+It is then by constructing a phylogenetic covariance function \(\Sigma_{\mathbf{T}}\left(\mathcal{E}^{S}(\cdot), \mathcal{E}^{S'}(\cdot)\right)\) that evolutionary inference can be performed.
+
+Deriving a tractable form of the phylogenetic covariance function requires some simplifying assumptions.
+Firstly, it is assumed that conditional on their common ancestors in the phylogenetic tree \(\mathbf{T}\), any two traits are statistically independent.
+
+The second assumption is that the statistical relationship between a trait and any of it's descendants in \(\mathbf{T}\) is independent of the topology of \(\mathbf{T}\).
+That is to say that the underlying process driving evolutionary changes is identical along all individual branches of the tree.
+We call this underlying process along each branch the marginal process.
+The marginal process depends on the date of \(\mathbf{t}\), the distance between a point \(\mathbf{t} \in \mathbf{T}\) and the root of \(\mathbf{T}\), denoted \(t\).
+
+Finally, we assume that the covariance function of the marginal process is separable over evolutionary time and the function-valued trait space.
+Thus, by defining the frequency only covariance function \(K(f,f')\) and the time only covariance function \(k(t,t')\) the covariance function of the marginal process is
+\[\Sigma\left((f,t), (f',t')\right) = K(f, f') k(t,t')\]
+
+Under these conditions, Jones \& Moriarty \cite{jones2013evolutionary} show that the phylogenetic covariance function is also separable, that is\[\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right) = K(f, f') k_{\mathbf{T}}(\mathbf{t},\mathbf{t}').\]
+
+It is also shown that for a phylogenetic Gaussian Process with this covariance function, \(Y\), there exists a set of \(n\) deterministic basis functions \(\phi_i: F \to \mathbf{R}\) and univariate Gaussian processes \(X_i\) for \(i = 1,\dots, n\) such that \[
+g(f, \mathbf{t}) = \sum_{i = 1}^{n} \phi_i(f) X_i(\mathbf{t})\] has the same distribution as \(Y\).
+Thus, by defining an appropriate set of basis functions and Gaussian Processes, evolutionary inference can be performed on our function-valued traits.
+
+\subsection{Basis Functions for Evolution}
 

### 2017-05-09 Paper Drafts/Draft1/Draft1.tex
First Draft: Basis functions

@@
 
-\section{Gaussian Process Regression on Phylogenies}
+\section{Phylogenetic Gaussian Process Regression for Bat Echolocation Calls}
+
+\subsection{Gaussian Process Regression on Phylogenies}
 
@@
 
-It is also shown that for a phylogenetic Gaussian Process with this covariance function, \(Y\), there exists a set of \(n\) deterministic basis functions \(\phi_i: F \to \mathbf{R}\) and univariate Gaussian processes \(X_i\) for \(i = 1,\dots, n\) such that \[
-g(f, \mathbf{t}) = \sum_{i = 1}^{n} \phi_i(f) X_i(\mathbf{t})\] has the same distribution as \(Y\).
-Thus, by defining an appropriate set of basis functions and Gaussian Processes, evolutionary inference can be performed on our function-valued traits.
+It is also shown that for a phylogenetic Gaussian Process with this covariance function, \(Y\), and a degenerate Mercer kernel, \(K(\cdot, \cdot)\), there exists a set of \(n\) deterministic basis functions \(\phi_i: F \to \mathbf{R}\) and univariate Gaussian processes \(X_i\) for \(i = 1,\dots, n\) such that
+\[
+g(f, \mathbf{t}) = \sum_{i = 1}^{n} \phi_i(f) X_i(\mathbf{t})
+\] 
+has the same distribution as \(Y\).
 
-\subsection{Basis Functions for Evolution}
+Thus, given an appropriate set of basis functions, \(\phi_{\mathcal{E}} = [\phi^{\mathcal{E}}_1(\cdot), \dots, \phi^{\mathcal{E}}_n(\cdot)]\), and Gaussian Processes, \(X_{\mathcal{E}} = [X_1^{\mathcal{E}}(\cdot), \dots\ X_n^{\mathcal{E}}(\cdot)]\), the set of observations of the echolocation function-valued trait can be expressed in matrix notation as
+\[
+\mathcal{E} = X_{\mathcal{E}} \phi_{\mathcal{E}}^{\mathsf{T}},
+\]
+where \( X_{\mathcal{E}}\) is the matrix of mixing coefficients of the fixed basis functions determining the function-valued trait.
+The values of \( X_{\mathcal{E}}\) are modelled as evolving by univariate phylogenetic Gaussian Processes.
+
+\subsection{Deterministic Basis Functions}
+
+Applying this model to observed traits requires that \(\phi\) be estimated somehow.
+Hajipantelis et al.
+\cite{hadjipantelis2013function} addressed this problem.
+The model outlined above implicitly assumes that the rows of \(X_{\mathcal{E}}\) are independent.
+This in turn implies that each of the basis, \(\phi_i(\cdot)\), evolved independently of one another.
+\(\hat{\phi}\), the estimate for \(\phi\), must reflect this.
+
+A Functional Principal Components Analysis \cite{ramsay2006functional} of the traits would return a set of orthogonal basis functions.
+This dimension reduction technique allows the selection of \(n\) basis functions which describe some proportion of the variation in the sample.
+However, this implicitly assumes that the basis functions are also Gaussian, a strong assumption which may not be realistic.
+
+A less stringent condition is to assumes only that the basis functions are independent.
+Such a set of components can be found by an Independent Components Analysis.
+Blasche \& Wiskott\cite{blaschke2004cubica} present a method for deriving Independent Components.
+This two step procedure first implements a Principal Components Analysis to estimate the effective dimensionality of the dataset, before passing the effective dimensions to the CuBICA algorithm.
+This algorithm then rotates these effective dimensions until the third and fourth cumulants have also been diagonalised, which produces approximately independent basis functions.
+
+\subsection{The Phylogenetic Ornstein-Uhelenbeck Process}
 

### 2017-05-10 Paper Drafts/Draft1/Draft1.tex
First Draft: Phylo OU

@@
 
-\subsection{The Phylogenetic Ornstein-Uhelenbeck Process}
+\subsection{The Phylogenetic Ornstein-Uhlenbeck Process for Evolutionary Inference}
+
+The relationships between the mixing coefficients, \(X_{\mathcal{E}}\), are modelled by a phylogenetic Gaussian Process, which must be defined.
+The Ornstein-Uhlenbeck process offers a popular method of modelling stabilising selection in comparative studies.\cite{hansen1997stabilizing} \cite{collen2012evolution} \cite{hadjipantelis2013function} Here, each independent phylogenetic Gaussian Process, \(X_i(\cdot)\), is modelled as an Ornstein-Uhlenbeck process.
+
+The phylogenetic Ornstein-Uhlenbeck process is defined by the kernel
+\[
+k_{\mathbf{T}}^i(\mathbf{t}, \mathbf{t}') = (\sigma_p^i)^2 \exp \left( \frac{-d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')}{\ell^i} \right) + (\sigma_n^i)^2 \delta_{\mathbf{t}, \mathbf{t}'}
+\] 
+where \(\delta\) is the Kronecker delta, \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) is the cophenetic distance between the points \(\mathbf{t}\) and \(\mathbf{t}'\) on the phylogeny \(\mathbf{T}\), and \(\mathbf{\theta}^i = [\sigma_p^i, \ell^i, \sigma_n^i]^{\mathsf{T}}\) is the vector of hyperparameters for the process \(X_i(\cdot)\).
+
+The full phylogenetic covariance function is then
+\[
+\Sigma_{\mathbf{T}}((f, \mathbf{t}), (f', \mathbf{t}')) = \sum_{i = 1}^{n}  k_{\mathbf{T}}^i(\mathbf{t}, \mathbf{t}') \phi_i^{\mathcal{E}}(f) \phi_i^{\mathcal{E}}(f')
+\]
+and the log likelihood associated with the model is
+\[
+\ell(\mathcal{E} | \theta) = -\frac{1}{2} \sum_{i = 1}^{n} \left( X_i(\cdot)^{\mathsf{T}} k_{\mathbf{T}}^i(\cdot, \cdot)^{-1}  X_i(\cdot) + \log( \det (k_{\mathbf{T}}^i(\cdot, \cdot))) + |X_i(\cdot)| \log 2\pi   \right),
+\]
+where \(\theta = [\theta^1, \dots , \theta^n]\), and \(|X_i(\cdot)|\) is the Euclidean length of the vector of realisations of \(X_i(\cdot)\).
+
+Model selection can be performed by a type II maximum likelihood estimation procedure which maximises the likelihood of the sample with respect to \(\theta\).
+
+The model hyperparameters have intuitive interpretations.
+The variance of observations in the sample is \(\sigma_p + \sigma_n\), where \(\sigma_p\) is the phylogenetic noise, and \(\sigma_n\) is the non-phylogenetic noise.
+\(\sigma_p\) is the proportion of the variance within the sample which due to phylogenetic relationships, while \(\sigma_n\) accounts for other sources of variation.
+The length-scale parameter, \(\ell\) then indicates the strength of the correlation between traits at various points on the phylogeny, where strong correlations are given by a large \(\ell\).
+
+Finally, ancestral reconstruction of the function-valued trait at some unobserved point in the phylogeny, \(\mathcal{E}^*\), is given by its posterior distribution 
+\[
+p(\mathcal{E}^* | \mathcal{E})  \sim \mathcal{N}(A, B)
+\]
+where 
+\[
+A = \sum_{i=1}^{n} k_{\mathbf{T}}^i(\mathbf{t}^*, \cdot) k_{\mathbf{T}}^i(\cdot, \cdot)
+^{-1} X_i^{\mathcal{E}} (\cdot) \phi_i^{\mathcal{E}}
+\]
+and
+\[
+B = \sum_{i=1}^{n}\left( k_{\mathbf{T}}^i(\mathbf{t}^*, \mathbf{t}^*) -  k_{\mathbf{T}}^i(\mathbf{t}^*, \cdot) k_{\mathbf{T}}^i(\cdot, \cdot)
+^{-1} k_{\mathbf{T}}^i(\mathbf{t}^*, \cdot)^{\mathsf{T}} \right) \phi_i^{\mathcal{E}}
+\]
+
+Thus, all the tools required to perform an ancestral reconstruction of the function-valued trait have been defined.
 

### 2017-05-10 Paper Drafts/Draft1/Draft1.tex
First Draft: Data Description and Simulation Study

@@
 
+\section{Results}
+
+\subsection{Data Description}
+
+The post processed echolocation call data accompanying Stathopoulos et al.
+\cite{stathopoulos2017bat} was used in this analysis.
+Echolocation calls were recorded across north and central Mexico with a Pettersson 1000x bat detector (Pettersson Elektronik AB, Uppsala, Sweden).
+Live trapped bats were measured and identified to species level using field keys.\cite{ceballos2005mamiferos} \cite{medellin2sanchez} Bats were recorded either while released from the hand or while tied to a zip line.
+The bat detector was set to record calls manually in real time, full spectrum at 500 kHz.
+In total the dataset consists of 22 species in five families, 449 individual bats and 1816 individual echolocation call recordings.
+
+Collen's \cite{collen2012evolution} Bat super-tree provided the basis for the phylogenetic tree of the recorded bat species, \(\mathbf{T}\).
+
+\begin{table}[ht]
+	\tbl{Echolocation Call Dataset Statistics}
+	{\begin{tabular}{@{}lccc@{}} \toprule
+		Species & Key & Individuals & Calls \\ \colrule
+		Family: Emballonuridae &&& \\
+		1 \textit{Balantiopteryx plicata} & Bapl & 16 & 100 \\
+		\\
+		Family: Molossidae &&& \\
+		2 \textit{Nyctinomops femorosaccus} & Nyfe & 16 & 100 \\
+		3 \textit{Tadarida brasiliensis} & Tabr & 49 & 100  \\
+		\\
+		Family: Vespertilionidae &&& \\
+		4 \textit{Antrozous pallidus} & Anpa & 58 & 100 \\
+		5 \textit{Eptesicus fuscus} & Epfu & 74 & 100 \\
+		6 \textit{Idionycteris phyllotis} & Idph & 6 & 100 \\
+		7 \textit{Lasiurus blossevillii} & Labl & 10 & 90 \\
+		8 \textit{Lasiurus cinereus} & Laci & 5 & 42 \\
+		9 \textit{Lasiurus xanthinus} & Laxa & 8 & 100 \\
+		10 \textit{Myotis volans} & Myvo & 8 & 100 \\
+		11 \textit{Myotis yumanensis} & Myyu & 5 & 89 \\
+		12 \textit{Pipistrellus hesperus} & Pihe & 85 & 100 \\
+		\\
+		Family: Mormoopidae &&& \\
+		13 \textit{Mormoops megalophylla} & Mome & 10 & 100 \\
+		14 \textit{Pteronotus davyi} & Ptda & 8 & 100 \\
+		15 \textit{Pteronotus parnellii} & Ptpa & 23 & 100 \\
+		16 \textit{Pteronotus personatus} & Ptpe & 7 & 51 \\
+		\\
+		Family:Phyllostomidae &&& \\
+		17 \textit{Artibeus jamaicensis} & Arja & 11 & 82 \\
+		18 \textit{Desmodus rotundus} & Dero & 6 & 38 \\
+		19 \textit{Leptonycteris yerbabuenae} & Leye & 26 & 100 \\
+		20 \textit{Macrotus californicus} & Maca & 6 & 53 \\
+		21 \textit{Sturnira ludovici} & Stlu & 8 & 51 \\
+		22 \textit{Sturnira lilium} & Stli & 4 & 20 \\
+		\botrule
+		\end{tabular}
+	}
+	\label{tab::dataset}
+\end{table}
+
+\subsection{Simulation Study}
+
+In order to demonstrate the efficacy of hyperparameter estimation for a phylogenetic Ornstein-Uhlenbeck process over the phylogeny \(\mathbf{T}\), a short simulation study was carried out.
+For this study, samples of a univariate phylogenetic Ornstein-Uhlenbeck processes with hyperparameters \(\theta= [1,50,1]^{\mathsf{T}}\) were generated.
+Type II maximum likelihood point estimates of the hyperparameters, were made for each sample.
+The samples varied in the number of observations taken at the leaf nodes of \(\mathbf{T}\), denoted \(n\), corresponding to the number of observations taken per species.
+This resulted in a full sample of size \(n \times 22\).
+The experiment was repeated until there were 1000 samples each for \(n \in \{1,2,4,8\}\).
+The distribution of the type II maximum likelihood hyperparameter estimates was then examined.
+
+\begin{table}[ht]
+	\tbl{\(\hat{\theta}\) Distribution}
+	{
+		\begin{tabular}{@{}lcccc@{}} \toprule
+			\(n\) & 1 & 2 & 4 & 8  \\ 
+			\colrule
+			\(\sigma_p \)  & mean (std dev) & mean (std dev) & mean (std dev)  \\
+			\(\ell\)  & mean (std dev) & mean (std dev) & mean (std dev)  \\
+			\(\sigma_n\)  & mean (std dev) & mean (std dev) & mean (std dev)  \\
+
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::simulation}
+\end{table}
+
+The results of this simulation study suggest that for any given sample \(\hat{\theta}\) does not provide a good estimate for \(\theta\), thus multiple samples will be required to produce estimates for model hyperparameters.
+It is also clear that having more observations at each node of the phylogeny results in better hyperparamenter estimation.
+therefore it is important to ensure that this is the case for inference on \(\mathbf{T}\) for bat echolocation calls.
+
 \bibliographystyle{ws-rv-van}

### 2017-05-11 Paper Drafts/Draft1/Draft1.tex
First Draft: Full Draft less graphics and results tables

@@
 
+\subsection{The Evolution of Echolocation in Bats}
+
+Given the functional representation of the bat echolocation calls described a set of deterministic basis functions need to be defined.
+These basis functions should be derived in a methodologically sound while still having an interpretation that reflects the specific dataset being considered.
+
+An important consideration in the analysis of any dataset is the notion of sampling bias.
+Even if we accept that that obtaining a representative sample of all the 1000 plus species of bat, is not possible, the small dataset analysed here is inherently imbalanced, as illustrated in Table \ref{tab:dataset}.
+At all stages of the analysis reasonable steps were taken to mitigate the effects of sampling bias.
+
+The first step in identifying a set of stable, interpretable, independent basis functions involved a Principal Components Analysis of 1000 samples of Echolocation Calls, balanced by both individual and species.
+These balanced samples randomly selected 4 individuals per species and then  randomly selected one of these individuals call recordings.
+Over all these samples it was found that the first 6 principal components generally accounted for 85\% of the variation in the sample, with the \(6^{th}\) component accounting for 3\% of the variation.
+The first and second Principal components were consistent across samples and so the mean of all these sampled components were identified as the first and second overall principal components.
+
+The remaining components were then grouped by the frequency with the absolute maximum weight.
+Grouping components in this way means that each component can be interpreted as the frequency with maximum energy.
+In this way 6 overall principal components were identified.
+These principal components were then passed through the CuBICA algorithm and independent basis functions for the bats echolocation calls identified.
+These basis functions were then passed on for analysis alongside the phylogenetic Ornstein-Uhlenbeck process.
+
+We are interested in the evolution of echolocation in bats to a species level, thus it is important to consider function valued traits that are representative of the species.
+Thus we consider estimates of the species level traits by taking balsanced estimates of the mean trait for that species.
+These mean estimates took one calls from 4 individuals.
+4 such estimates were taken to form a sample of traits for type II MLE.
+1000 samples were taken and hyperparameters estimated for each basis function.
+The distribution of these estimates were used to create hyperparameter estimates.
+
+\begin{table}[ht]
+	\tbl{\(\hat{\theta}\) Distribution}
+	{
+		\begin{tabular}{@{}lccccccc@{}} \toprule
+			Basis: & Bias & 1 & 2 & 3 & 4 & 5 & 6 \\ 
+			\colrule
+			\(\sigma_p \)  & mean (std dev) & mean (std dev) & mean (std dev) &&&&  \\
+			\(\ell\)  & mean (std dev) & mean (std dev) & mean (std dev) &&&& \\
+			\(\sigma_n\)  & mean (std dev) & mean (std dev) & mean (std dev) &&&&  \\
+
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::bat_results}
+\end{table}
+
+Ancestral reconstructions were performed by taking the sample mean for each species and obtaining posterior distributions based on these mean estimates.
+
+\section{Conclusions and Future Work}
+
+Demonstrated an application of phylogenetic Gaussian processes for a dataset. Higlighted issues with balanced samples, how to leverage the dataset to account for this.
+
+Identified features of calls, corresponding to frequencies. Demonstrated a possible evolutionary dynamic.
+
+Performed ancestral reconstructions which suggest...
+
+Early stages of this work, there are some extensions to be made. Spectrograms as function valued traits, account for temporal variation in call.
+
+consider other representations.
+
+Consider ways of extending these models to more realistic models of evolution, relax assumptions.
+
+When methodology has bee resolved extend to a larger dataset. EchoBank is there. Hopefully contribute to the understanding of the evolution of echolocation in bats.
+
 \bibliographystyle{ws-rv-van}

### 2017-05-12 Paper Drafts/Draft1/Draft1.tex
Results Tables
First Draft: Analysis Results

@@
 
+Removes estimates where \(\ell > 1000\) to stabilise distribution.
+
 \begin{table}[ht]
@@
 	{
-		\begin{tabular}{@{}lcccc@{}} \toprule
-			\(n\) & 1 & 2 & 4 & 8  \\ 
+		\begin{tabular}{@{}cccc@{}} \toprule
+			\(n\)  & \(\sigma_p \) & \(\ell\) & \(\sigma_n\) \\
 			\colrule
-			\(\sigma_p \)  & mean (std dev) & mean (std dev) & mean (std dev)  \\
-			\(\ell\)  & mean (std dev) & mean (std dev) & mean (std dev)  \\
-			\(\sigma_n\)  & mean (std dev) & mean (std dev) & mean (std dev)  \\
-
+			1 & 1.09 (0.47) & \(10^{14}\) (\(10^{15}\)) & 0.57 (0.54) \\
+			2 & 0.97 (0.29) & \(10^{13}\) (\(10^{14}\)) & 0.99 (0.15) \\
+			4 & 0.97 (0.25) & 63.66 (136.96) & 1.00 (0.09) \\
+			8 & 0.99 (0.24) & 56.21 (48.24) & 1.00 (0.06) \\			
 			\botrule
@@
 	}
-	\label{tab::simulation}
+	\label{tab::simulation_option2}
+\end{table}
+
+\begin{table}[ht]
+	\tbl{\(\hat{\theta}\) Distribution}
+	{
+		\begin{tabular}{@{}cccc@{}} \toprule
+			\(n\)  & \(\sigma_p \) & \(\ell\) & \(\sigma_n\) \\
+			\colrule
+			1 & 0.46 (0.21) & \(10^{14}\) (\(10^{15}\)) & 0.07 (0.08) \\
+			2 &  0.45 (0.20) & 259.18 (309.02) & 0.10 (0.02) \\
+			4 & 0.45 (0.20) & 279.50 (363.12) & 0.10 (0.01) \\
+			8 & 0.47 (0.21) & 291.61 (343.14) & 0.10 (0.01) \\			
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::simulation_option3}
+\end{table}
+
+\begin{table}[ht]
+	\tbl{\(\hat{\theta}\) Distribution}
+	{
+		\begin{tabular}{@{}cccc@{}} \toprule
+			& \(\sigma_p \) & \(\ell\) & \(\sigma_n\) \\
+			\colrule
+			Data Generating values & 0.50 & 250 & 0.10 \\
+			Full Sample Estimate &  0.41 & 156.75 & 0.10 \\
+			Resampled Estimates & 0.42 (0.01) & 161.06 (27.11) & 0.10 (0.01) \\		
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::simulation_option4}
 \end{table}
@@
 
+\begin{table}[ht]
+	\tbl{\(\hat{\theta}\) Distribution}
+	{
+		\begin{tabular}{@{}cccc@{}} \toprule
+			Basis & \(\sigma_p\) & \(\ell\) & \(\sigma_n\) \\
+			\colrule
+			1 & 0.35 (0.008) & 224.5 (39.15) & 0.08 (0.007) \\
+			2 & 0.51 (0.015) & 287.9 (44.76) & 0.09 (0.009) \\
+			3 & 0.42 (0.021)& 204.2 (30.39) & 0.08 (0.009) \\
+			4 & 0.59 (0.019) & 390.1 (74.44) & 0.14 (0.013) \\
+			5 & 0.51 (0.010) & 339.8 (42.81) & 0.08 (0.008) \\
+			6 & 0.48 (0.013) & 148.6 (23.29) & 0.09 (0.011) \\
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::bat_results_option2}
+\end{table}
+
 Ancestral reconstructions were performed by taking the sample mean for each species and obtaining posterior distributions based on these mean estimates.

### 2017-05-15 Paper Drafts/Draft2/Draft2.tex
Draft 2: Intro and Functional Representation

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Review Volume (last updated on 20-4-2015)                            %%
%% Trim Size: 9in x 6in                                                 %%
%% Text Area: 7.35in (include runningheads) x 4.5in                     %%
%% Main Text: 10 on 13pt                                                %%
%% For support: Yolande Koh, <ykoh@wspc.com.sg>                         %%
%%              D. Rajesh Babu, <rajesh@wspc.com.sg>                    %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
%\documentclass{ws-rv9x6}
\usepackage{subfigure}   % required only when side-by-side / subfigures are used
\usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
\usepackage{ws-rv-van}   % numbered citation & references (default)
%\usepackage{ws-index}   % to produce multiple indexes
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}       %author index
%\renewindex{default}{idx}{ind}{Subject Index}  %subject index

\begin{document}

\chapter[Ancestral Reconstruction of Bat Echolocation Calls]{Ancestral Reconstruction of Bat Echolocation Calls}\label{ra_ch1}

\author[J.P. Meagher et al.]{J.P. Meagher\footnote{Author footnote.}}
%\index[aindx]{Author, F.} % or \aindx{Author, F.}
%\index[aindx]{Author, S.} % or \aindx{Author, S.}

\address{Department of Statistics,\\
University of Warwick, \\
J.Meagher@Warwick.ac.uk\footnote{Affiliation footnote.}}

\begin{abstract}
 Something something bats something evolution something ancestral reconstruction something magic.
\end{abstract}
%\markright{Customized Running Head for Odd Page} % default is Chapter Title.
\body

%\tableofcontents
\section{Introduction}

The emerging field of Data Science is driven by research which lies at the nexus of Statistics and Computer Science. Bioacoustics is one such area generating vast quantities of data, often through citizen science initiatives. \cite{allen2006citizen} \cite{pettorelli2013indicator} Bioacoustic techniques for biodiversity monitoring \cite{stathopoulos2017bat} \cite{damoulas2010bayesian} have the potential to make real policy impacts, particularly with regard to sustainable economic development and nature conservation.

Bats (order \textit{Chiroptera}) have been identified as ideal bioindicators for monitoring climate change and habitat quality,\cite{jones2009carpe} and are of particular interest for monitoring biodiversity acoustically. Typically, a bat broadcasts information about itself in an ultrasonic echolocation call.\cite{griffin1944echolocation} The development of automatic acoustic monitoring algorithms \cite{stathopoulos2017bat} \cite{walters2012continental} means that large scale, non-invasive monitoring of bats is becoming possible.

Monitoring bat populations provides useful information, but an understanding the orders natural history is required to identify the cause and effect of any changes observed. The echolocation calls structure, which reflects a bats diet and habitat,\cite{aldridge1987morphology}  is a key aspect of this natural history. Reconstructing ancestral traits\cite{joy2016ancestral} relies on a statistical comparative analysis incorporating extant species and fossil records. \cite{felsenstein2004inferring} However, the fossil record is of limited use in inferring ancestral bats echolocation calls. Therefore, statistical data science techniques may shed some light on this topic.

Previous studies of bat echolocation calls for both classification \cite{walters2012continental} and ancestral reconstruction \cite{collen2012evolution} analysed features extracted from the call spectrogram. These call features relied domain knowledge to ensure they were sensibly selected and applied. More recently, general techniques for the classification of acoustic signals have been developed. \cite{stathopoulos2014bat} \cite{damoulas2010bayesian} General techniques for the ancestral reconstruction of function-valued traits have also been proposed. \cite{group2012phylogenetic} This piece of research applies some proposed techniques to bat echolocation calls.

A function-valued trait\cite{meyer2005up} is a characteristic of some organism measured along some continuous scale, usually time, and can be modelled as a continuous mathematical function by functional data analysis. \cite{ramsay2006functional} Jones \& Moriarty \cite{jones2013evolutionary} developed a method which extends Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits over a phylogeny. A full demonstration of the phylogenetic Gaussian Process method for ancestral reconstruction on a synthetic dataset is presented by Hajipantelis et al.\cite{hadjipantelis2013function}

This general approach to evolutionary inference for function-valued traits is implemented here for a set of bat echolocation calls. Our goal in doing so is twofold. These techniques had previously been considered in the context of modelling the evolution of human speech sounds in language. \cite{group2012phylogenetic} It is hoped that by applying these methods in a simpler context progress can be made towards resolving methodological problems. For example, how do we extend these methods to more realistic models of evolution?

We are also interested in what specifically these models tell us about bats and the evolutionary dynamics of echolocation. What impact might these results have on our understanding of ancestral bats and their behaviour?

This paper presents the early stages of our research and some preliminary results.

\section{Echolocation Calls as Function-Valued Traits}

A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time. \cite{ramsay2006functional} 
These measurements can be thought of as representing points on a curve that varies gradually and continuously. In the context of phylogenetics, these functional data objects are function-valued traits. \cite{meyer2005up}

Denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of species \(S\) by \(\{\tilde{x}_{lm}^S(n) : n = 0, \dots, N_{lm}^S - 1\}\). 
Thus, \(\{x_{lm}^S(\cdot)\}\) is a series of discrete measurements of the function \(x_{lm}^S(\cdot)\), the \(m^{th}\) call function for the \(l^{th}\) individual bat, observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording. Assume then that \(x_{lm}^S(\cdot) = x_{l}^S(\cdot) + z_{lm}^S(\cdot)\), where \(x_{l}^S(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^S(\cdot)\) is the noise process for the \(m^{th}\) call. Further to this, assume that \(x_{l}^S(\cdot) = x^S(\cdot) + z_{l}^S(\cdot)\) where \(x^S(\cdot)\) is the representative call function for the bat species \(S\) and \(z_{l}^S(\cdot)\) is the noise process for the \(l^{th}\) individual. It is the echolocation call functions at the species level that we re interested in.

The call recordings themselves are functional data objects, however modelling the phylogenetic relationships between \(\{x_{lm}^S(t)\}\) and \(\{x_{l'm'}^{S'}(t)\}\) directly implies that the processes are comparable at time \(t\). 
This is not the case for acoustic signals, a phenomenon which is often addressed by dynamic time warping.\cite{berndt1994using} Another approach to this issue is to consider an alternative functional representation of the signal.

The Fourier transform \(x_{lm}^{S}(\cdot)\) is given by
\[
X_{lm}^S(f) = \int_{-\infty}^{\infty} x_{lm}^{S}(t) e^{-i 2\pi f t} dt.
\label{eqn:dft}
\]
The energy spectral density of \(x_{lm}^{S}(\cdot)\) is the magnitude of the Fourier transform and the log energy spectral density is given by
\begin{equation}
\mathcal{E}_{lm}^S(\cdot) = 10 \log_{10} \left( |X_{lm}^S(\cdot)|^2 \right).
\label{eqn:esd}
\end{equation}

Similarly to the call functions, \(\mathcal{E}_{lm}^S(\cdot)\) is the log energy spectral density of the \(m^{th}\) call of the \(l^{th}\) individual from species \(S\) where \(\mathcal{E}_{lm}^S(\cdot) = \mathcal{E}_{l}^S(\cdot) + \mathcal{Z}_{lm}^S(\cdot)\) and \(\mathcal{E}_{l}^S(\cdot) = \mathcal{E}^S(\cdot) + \mathcal{Z}_{l}^S(\cdot)\) where \(\mathcal{Z}_{lm}^S(\cdot)\) and \(\mathcal{Z}_{l}^S(\cdot)\) are noise processes.  The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \(F = [0, \frac{f_S}{2}]\).\cite{antoniou2006digital} It is now assumed that \(\mathcal{E}_{lm}^S(f)\) and \(\mathcal{E}_{l'm'}^{S'}(f)\) are comparable at frequency \(f\).

The discrete Fourier Transform\cite{antoniou2006digital}
of \(\{\tilde{x}_{lm}^{S}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^S(k) : k = 0, \dots, \frac{ N_{lm}^S}{2} + 1\}\). Smoothing splines\cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{S}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{S}(\cdot)\).

We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \mathcal{E}_{lm}^S(f)\}\) and \(\{f, \mathcal{E}_{l'm'}^{S'}(f)\}\) are directly comparable. These function-valued traits can now be modelled for evolutionary inference.

\bibliographystyle{ws-rv-van}
\bibliography{../BatBiblio}

%\blankpage
%\printindex[aindx]                 % to print author index
%\printindex                         % to print subject index

\end{document}

### 2017-05-15 R Code/Components/Components.R

@@ -3,6 +3,7 @@ set.seed(8)
 
 # Create the samples of calls balanced by species and individual.
 N <- 1000
+
 resampled_calls <- balanced_samples(smooth_and_regularise_call_densities(), n_samples = N)
 
 # Proportion of variance accounted for by each component.
@@ -139,4 +140,4 @@ spectral_mean <- mean_centring_curve
 
 
 devtools::use_data(spectral_fpca)
-devtools::use_data(spectral_mean)
+devtools::use_data(spectral_mean, overwrite = T)

### 2017-05-16 Paper Drafts/Draft2/Draft2.tex
Draft 2: PGP and Simulation Study

@@
 \section{Introduction}
-
 The emerging field of Data Science is driven by research which lies at the nexus of Statistics and Computer Science.
@@
 Monitoring bat populations provides useful information, but an understanding the orders natural history is required to identify the cause and effect of any changes observed.
-The echolocation calls structure, which reflects a bats diet and habitat,\cite{aldridge1987morphology}  is a key aspect of this natural history.
+The echolocation call structure, which reflects a bats diet and habitat,\cite{aldridge1987morphology}  is a key aspect of this natural history.
 Reconstructing ancestral traits\cite{joy2016ancestral} relies on a statistical comparative analysis incorporating extant species and fossil records.
@@
 \section{Echolocation Calls as Function-Valued Traits}
-
 A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time. \cite{ramsay2006functional} 
@@
 
-Denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of species \(S\) by \(\{\tilde{x}_{lm}^S(n) : n = 0, \dots, N_{lm}^S - 1\}\). 
-Thus, \(\{x_{lm}^S(\cdot)\}\) is a series of discrete measurements of the function \(x_{lm}^S(\cdot)\), the \(m^{th}\) call function for the \(l^{th}\) individual bat, observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording.
-Assume then that \(x_{lm}^S(\cdot) = x_{l}^S(\cdot) + z_{lm}^S(\cdot)\), where \(x_{l}^S(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^S(\cdot)\) is the noise process for the \(m^{th}\) call.
-Further to this, assume that \(x_{l}^S(\cdot) = x^S(\cdot) + z_{l}^S(\cdot)\) where \(x^S(\cdot)\) is the representative call function for the bat species \(S\) and \(z_{l}^S(\cdot)\) is the noise process for the \(l^{th}\) individual.
+Given a phylogenetic tree \(\mathbf{T}\), representing the evolutionary relationships between the recorded bat species, we denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of the species observed at point \(\mathbf{t} \in \mathbf{T}\) by \(\{\tilde{x}_{lm}^{\mathbf{t}}(n) : n = 0, \dots, N_{lm}^{\mathbf{t}} - 1\}\).
+Thus, \(\{x_{lm}^{\mathbf{t}}(\cdot)\}\) is a series of discrete measurements of the function \(x_{lm}^{\mathbf{t}}(\cdot)\), the \(m^{th}\) call function for the \(l^{th}\) individual bat, observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording.
+Assume then that \(x_{lm}^{\mathbf{t}}(\cdot) = x_{l}^{\mathbf{t}}(\cdot) + z_{lm}^{\mathbf{t}}(\cdot)\), where \(x_{l}^{\mathbf{t}}(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^{\mathbf{t}}(\cdot)\) is the noise process for the \(m^{th}\) call.
+Further to this, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual.
 It is the echolocation call functions at the species level that we re interested in.
 
-The call recordings themselves are functional data objects, however modelling the phylogenetic relationships between \(\{x_{lm}^S(t)\}\) and \(\{x_{l'm'}^{S'}(t)\}\) directly implies that the processes are comparable at time \(t\).
+The call recordings themselves are functional data objects, however modelling the phylogenetic relationships between \(\{x_{lm}^{\mathbf{t}}(t)\}\) and \(\{x_{l'm'}^{{\mathbf{t}}'}(t)\}\) directly implies that the processes are comparable at time \(t\).
 This is not the case for acoustic signals, a phenomenon which is often addressed by dynamic time warping.\cite{berndt1994using} Another approach to this issue is to consider an alternative functional representation of the signal.
 
-The Fourier transform \(x_{lm}^{S}(\cdot)\) is given by
+The Fourier transform of \(x_{lm}^{\mathbf{t}}(\cdot)\) is given by
 \[
-X_{lm}^S(f) = \int_{-\infty}^{\infty} x_{lm}^{S}(t) e^{-i 2\pi f t} dt.
+X_{lm}^{\mathbf{t}}(f) = \int_{-\infty}^{\infty} x_{lm}^{\mathbf{t}}(t) e^{-i 2\pi f t} dt.
 \label{eqn:dft}
 \]
-The energy spectral density of \(x_{lm}^{S}(\cdot)\) is the magnitude of the Fourier transform and the log energy spectral density is given by
+The energy spectral density of \(x_{lm}^{\mathbf{t}}(\cdot)\) is the magnitude of the Fourier transform and the log energy spectral density is given by
 \begin{equation}
-\mathcal{E}_{lm}^S(\cdot) = 10 \log_{10} \left( |X_{lm}^S(\cdot)|^2 \right).
+\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = 10 \log_{10} \left( |X_{lm}^{\mathbf{t}}(\cdot)|^2 \right).
 \label{eqn:esd}
@@
 
-Similarly to the call functions, \(\mathcal{E}_{lm}^S(\cdot)\) is the log energy spectral density of the \(m^{th}\) call of the \(l^{th}\) individual from species \(S\) where \(\mathcal{E}_{lm}^S(\cdot) = \mathcal{E}_{l}^S(\cdot) + \mathcal{Z}_{lm}^S(\cdot)\) and \(\mathcal{E}_{l}^S(\cdot) = \mathcal{E}^S(\cdot) + \mathcal{Z}_{l}^S(\cdot)\) where \(\mathcal{Z}_{lm}^S(\cdot)\) and \(\mathcal{Z}_{l}^S(\cdot)\) are noise processes.
-The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \(F = [0, \frac{f_S}{2}]\).\cite{antoniou2006digital} It is now assumed that \(\mathcal{E}_{lm}^S(f)\) and \(\mathcal{E}_{l'm'}^{S'}(f)\) are comparable at frequency \(f\).
+Similarly to the call functions, \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\) is the log energy spectral density of the \(m^{th}\) call of the \(l^{th}\) individual from the species at \({\mathbf{t}}\) where \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = \mathcal{E}_{l}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{E}_{l}^{\mathbf{t}}(\cdot) = \mathcal{E}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) where \(\mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) are noise processes.
+The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \(F = [0, \frac{f_S}{2}]\).\cite{antoniou2006digital}
 
 The discrete Fourier Transform\cite{antoniou2006digital}
-of \(\{\tilde{x}_{lm}^{S}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^S(k) : k = 0, \dots, \frac{ N_{lm}^S}{2} + 1\}\).
-Smoothing splines\cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{S}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{S}(\cdot)\).
+of \(\{\tilde{x}_{lm}^{\mathbf{t}}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^{\mathbf{t}}(k) : k = 0, \dots, \frac{ N_{lm}^{\mathbf{t}}}{2} + 1\}\).
+Smoothing splines\cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\).
 
-We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \mathcal{E}_{lm}^S(f)\}\) and \(\{f, \mathcal{E}_{l'm'}^{S'}(f)\}\) are directly comparable.
+We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \mathcal{E}_{lm}^{\mathbf{t}}(f)\}\) and \(\{f, \mathcal{E}_{l'm'}^{{\mathbf{t}}'}(f)\}\) are directly comparable.
 These function-valued traits can now be modelled for evolutionary inference.
+
+\section{Phylogenetic Gaussian Processes}
+A Gaussian Process is a collection of random variables, any finite number of which have a joint Gaussian distribution.
+A Gaussian process prior can be defined as a distribution over functions, \(f(x) \sim \mathcal{GP}(m(x), k (x, x'),\) where \(x \in \mathbf{R}^P\) is some input variable, the mean function \(m(x) = \mathbf{E}[f(x)]\), and the covariance kernel \(k(x, x') = \mathnormal{cov}(f(x), f(x') )\).
+Any collection of function values has a joint Gaussian distribution
+\(
+[f(x_1), f(x_2), \dots f(x_N)]^{\mathsf{T}} \sim \mathcal{N}(\)\boldmath\(\mu\)\unboldmath, \(\mathbf{K}
+\))
+where the \(N \times N\) covariance matrix \(\mathbf{K}\)  has the entries \(K_{ij} = k(x_i, x_j)\), and the mean \boldmath\(\mu\)\unboldmath\(\) has entries \(\mu_i = m(x_i)\).
+Thus, the properties of the functions are determined by the kernel function.
+Assuming that the function values \(\mathbf{y}\) observed at locations \(\{x_n\}_{n=1}^N\) are subject to Gaussian noise, and given the kernel hyperparameters \(\theta\), a posterior predictive distribution over Gaussian process functions can be inferred analytically.
+A marginal likelihood \(p(\mathbf{y} | \theta, \{x_n\}_{n=1}^N)\), which can in turn be used to estimate the kernel hyperparameters, can also be derived analytically.
+An in depth treatment of Gaussian processes is given by Rasmussen \& Williams.\cite{rasmussen2006gaussian}
+
+Jones \& Moriarty\cite{jones2013evolutionary} extend Gaussian processes for inference on function-valued traits over a phylogeny.
+Consider \(\mathcal{E}^{\mathbf{t}}(\cdot)\), a functional representation of the echolocation call of the species observed at the point \(\mathbf{t}\) on the phylogenetic tree \(\mathbf{T}\) with respect to frequency.
+Modelling this as Gaussian process function where \(\mathcal{E}^{\mathbf{t}}(f)\) corresponds to a point \((f, \mathbf{t})\) on the frequency-phylogeny \(F \times \mathbf{T}\) requires that a suitable phylogenetic covariance function, \(\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right)\), is defined.
+
+Deriving a tractable form of the phylogenetic covariance function requires some simplifying assumptions.
+Firstly, it is assumed that conditional on their common ancestors in the phylogenetic tree \(\mathbf{T}\), any two traits are statistically independent.
+
+The second assumption is that the statistical relationship between a trait and any of it's descendants in \(\mathbf{T}\) is independent of the topology of \(\mathbf{T}\).
+That is to say that the underlying process driving evolutionary changes is identical along all individual branches of the tree.
+We call this underlying process along each branch the marginal process.
+The marginal process depends on the date of \(\mathbf{t}\), the distance between \(\mathbf{t}\) and the root of \(\mathbf{T}\), denoted \(t\).
+
+Finally, it is assumed that the covariance function of the marginal process is separable over evolutionary time and the function-valued trait space.
+Thus, by defining the frequency only covariance function \(K(f,f')\) and the time only covariance function \(k(t,t')\) the covariance function of the marginal process is \(\Sigma\left((f,t), (f',t')\right) = K(f, f') k(t,t')\).
+
+Under these conditions the phylogenetic covariance function is also separable and so
+\begin{equation}
+\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right) = K(f, f') k_{\mathbf{T}}(\mathbf{t},\mathbf{t}').
+\label{eqn:phy}
+\end{equation}
+
+For a phylogenetic Gaussian Process \(Y\) with covariance function given by \ref{eqn:phy}, when \(K\) is a degenerate Mercer kernel, there exists a set of \(n\) deterministic basis functions \(\phi_i: F \to \mathbf{R}\) and univariate Gaussian processes \(X_i\) for \(i = 1,\dots, n\) such that
+\[
+g(f, \mathbf{t}) = \sum_{i = 1}^{n} \phi_i(f) X_i(\mathbf{t})
+\] 
+has the same distribution as \(Y\). The full phylogenetic covariance function of this phylogenetic Gaussian process is
+\[
+\Sigma_{\mathbf{T}}((f, \mathbf{t}), (f', \mathbf{t}')) = \sum_{i = 1}^{n}  k_{\mathbf{T}}^i(\mathbf{t}, \mathbf{t}') \phi_i(f) \phi_i(f'),
+\]
+where \(\int \phi_i(f) \phi_j(f) df = \delta_{ij}\), \(\delta\) being the Kronecker delta, and so the phylogenetic covariance function depends only on \(\mathbf{T}\).
+
+Thus, given function-valued traits observed at \(\mathbf{f} \times \mathcal{t}\) on the frequency-phylogeny, where \(\mathbf{f} = [f_1, \dots, f_q]^{\mathsf{T}}\) and \(\mathcal{t} = [\mathbf{t}_1, \dots, \mathbf{t}_Q]^{\mathbf{T}}\), an appropriate set of basis functions \(\phi_{F} = [\phi^{F}_1(\mathbf{f}), \dots, \phi^{F}_n(\mathbf{f})]\) for the traits \(\mathcal{E} = [\mathcal{E}^{\mathbf{t}}(\mathbf{f}), \dots, \mathcal{E}^{\mathbf{t}'}(\mathbf{f})]\), and Gaussian Processes, \(X_{\mathbf{T}} = [X_1^{\mathbf{T}}(\mathcal{t}), \dots\ X_n^{\mathbf{T}}(\mathcal{t})]\), the set of observations of the echolocation function-valued trait are then
+\begin{equation}
+\mathcal{E} = X_{\mathbf{T}} \phi_F^{\mathsf{T}}.
+\label{eqn:inv}
+\end{equation}
+
+The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis et al.
+\cite{hadjipantelis2013function} \(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott\cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for samples of traits balanced by species.
+Given \(\hat{\phi}_F\), the estimated matrix of mixing coefficients is \(\hat{X}_{\mathbf{T}} = \mathcal{E} (\hat{\phi}_F^{\mathsf{T}})^{-1}\).
+
+Each column of \(X_{\mathbf{T}}\) is an independent, univariate, phylogenetic Gaussian process, \(X_i^{\mathbf{T}}(\mathcal{t})\), modelled here with phylogenetic Ornsetin-Uhlenbeck process kernel.
+
+The phylogenetic Ornstein-Uhlenbeck process is defined by the kernel
+\begin{equation}
+k_{\mathbf{T}}^i(\mathbf{t}, \mathbf{t}') = (\sigma_p^i)^2 \exp \left( \frac{-d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')}{\ell^i} \right) + (\sigma_n^i)^2 \delta_{\mathbf{t}, \mathbf{t}'}
+\label{eqn:oukernel}
+\end{equation}
+
+where \(\delta\) is the Kronecker delta, \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) is the cophenetic distance between \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), and \(\mathbf{\theta}^i = [\sigma_p^i, \ell^i, \sigma_n^i]^{\mathsf{T}}\) is the vector of hyperparameters for \(X_i^{\mathbf{T}}(\cdot)\).
+The phylogenetic covariance matrix for \(X_i^{\mathbf{T}}(\mathcal{t})\) is denoted \(\Sigma_{\mathbf{T}}^i(\mathcal{t}, \mathcal{t})\) and the marginal likelihood of the observed data given \(\theta\) is
+\begin{equation}
+\log(p(\mathcal{E} | \theta)) \propto -\frac{1}{2} \sum_{i = 1}^{n} \left( X_i(\mathcal{t})^{\mathsf{T}} \Sigma_{\mathbf{T}}^i(\mathcal{t}, \mathcal{t})^{-1}  X_i(\mathcal{t}) + \log |\Sigma_{\mathbf{T}}^i(\mathcal{t}, \mathcal{t})|    \right)
+\label{eqn:t2mle}
+\end{equation}
+and so \(\theta\) can be estimated by type II maximum likelihood estimation.
+
+Ancestral Reconstruction of the function valued trait for the species at \(\mathbf{t}^*\) then amounts to inferring the posterior predictive distribution \(p(\mathcal{E}^{\mathbf{t}^*} (\cdot) | \mathcal{E}) \sim \mathcal{N}(A, B)\) where
+\begin{equation}
+A = \sum_{i=1}^{n} \left( \Sigma_{\mathbf{T}}^i(\mathbf{t}^*, \mathcal{t}) \left( \Sigma_{\mathbf{T}}^i(\mathcal{t}, \mathcal{t})
+\right)^{-1} X_i^{\mathcal{E}} (\mathcal{t}) \phi_i(\cdot) \right)
+\end{equation}
+\begin{equation}
+B = \sum_{i=1}^{n}\left( \Sigma_{\mathbf{T}}^i(\mathbf{t}^*, \mathbf{t}^*) -  \Sigma_{\mathbf{T}}^i(\mathbf{t}^*, \mathcal{t}) \left( \Sigma_{\mathbf{T}}^i(\mathcal{t}, \mathcal{t})
+\right)^{-1} \Sigma_{\mathbf{T}}^i(\mathbf{t}^*, \mathcal{t})^{\mathsf{T}} \right) \phi_i (\cdot)
+\end{equation}
+
+We note that the elements of \(\theta\) each have intuitive interpretations.
+The variance of observations in the sample is \(\sigma_p + \sigma_n\), where \(\sigma_p\) is the phylogenetic noise, and \(\sigma_n\) is the non-phylogenetic noise.
+\(\sigma_p\) is the proportion of variation which can be accounted for by the cophenetic distance between any \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), while \(\sigma_n\) accounts for other sources of variation which may be observed at \(\mathbf{t}\).
+The length-scale parameter, \(\ell\) then indicates the strength of the correlation along \(\mathbf{T}\), where large values of \(\ell\) indicate a slowly decaying correlation.
+
+\section{Results}
+
+\subsection{Data Description}
+
+The post processed echolocation call data accompanying Stathopoulos et al.
+\cite{stathopoulos2017bat} was used in this analysis.
+Echolocation calls were recorded across north and central Mexico with a Pettersson 1000x bat detector (Pettersson Elektronik AB, Uppsala, Sweden).
+Live trapped bats were measured and identified to species level using field keys.\cite{ceballos2005mamiferos} \cite{medellin2sanchez} Bats were recorded either while released from the hand or while tied to a zip line.
+The bat detector was set to record calls manually in real time, full spectrum at 500 kHz.
+In total the dataset consists of 22 species in five families, 449 individual bats and 1816 individual echolocation call recordings.
+The distribution of these call recordings across species is summarised in Table \ref{tab::dataset}.
+
+Collen's \cite{collen2012evolution} Bat super-tree provided the basis for the phylogenetic tree of the recorded bat species, \(\mathbf{T}\).
+
+\begin{table}[ht]
+	\tbl{Echolocation Call Dataset}
+	{\begin{tabular}{@{}lccc@{}} \toprule
+			Species & Key & Individuals & Calls \\ \colrule
+			Family: Emballonuridae &&& \\
+			1 \textit{Balantiopteryx plicata} & Bapl & 16 & 100 \\
+			\\
+			Family: Molossidae &&& \\
+			2 \textit{Nyctinomops femorosaccus} & Nyfe & 16 & 100 \\
+			3 \textit{Tadarida brasiliensis} & Tabr & 49 & 100  \\
+			\\
+			Family: Vespertilionidae &&& \\
+			4 \textit{Antrozous pallidus} & Anpa & 58 & 100 \\
+			5 \textit{Eptesicus fuscus} & Epfu & 74 & 100 \\
+			6 \textit{Idionycteris phyllotis} & Idph & 6 & 100 \\
+			7 \textit{Lasiurus blossevillii} & Labl & 10 & 90 \\
+			8 \textit{Lasiurus cinereus} & Laci & 5 & 42 \\
+			9 \textit{Lasiurus xanthinus} & Laxa & 8 & 100 \\
+			10 \textit{Myotis volans} & Myvo & 8 & 100 \\
+			11 \textit{Myotis yumanensis} & Myyu & 5 & 89 \\
+			12 \textit{Pipistrellus hesperus} & Pihe & 85 & 100 \\
+			\\
+			Family: Mormoopidae &&& \\
+			13 \textit{Mormoops megalophylla} & Mome & 10 & 100 \\
+			14 \textit{Pteronotus davyi} & Ptda & 8 & 100 \\
+			15 \textit{Pteronotus parnellii} & Ptpa & 23 & 100 \\
+			16 \textit{Pteronotus personatus} & Ptpe & 7 & 51 \\
+			\\
+			Family: Phyllostomidae &&& \\
+			17 \textit{Artibeus jamaicensis} & Arja & 11 & 82 \\
+			18 \textit{Desmodus rotundus} & Dero & 6 & 38 \\
+			19 \textit{Leptonycteris yerbabuenae} & Leye & 26 & 100 \\
+			20 \textit{Macrotus californicus} & Maca & 6 & 53 \\
+			21 \textit{Sturnira ludovici} & Stlu & 8 & 51 \\
+			22 \textit{Sturnira lilium} & Stli & 4 & 20 \\
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::dataset}
+\end{table}
+
+\subsection{Hyperparameter Estimation for Phylogenetic Gaussian Processes}
+
+We are interested in modelling the evolution of \(\mathcal{E}^{\mathbf{t}}(\cdot)\), the function valued trait representing the echolocation call of the species of bat observed at \(\mathbf{t} \in \mathbf{T}\).
+However, only 22 species of bat are represented in \(\mathbf{T}\).
+The relatively small size of this dataset presents challenges for the estimation of the kernel hyperparameters in (\ref{eqn:oukernel}).
+A short simulation study was performed to investigate the dynamics at play.
+
+Although there are only 22 leaf nodes of the phylogeny \(\mathbf{T}\), we are not limited to a single observation at any given \(\mathbf{t}\).
+By sampling at each leaf node of \(\mathbf{T}\) multiple times, larger samples can be obtained, improving the quality of the estimators \(\hat{\theta}\).
+With this in mind, 1000 independent, univariate phylogenetic Gaussian processes were simulated for each of \(n = \{1,2,4,8\}\) according to the kernel (\ref{eqn:oukernel}) with \(\theta = [1,50,1]^{\mathsf{T}}\), where \(n\) is the number of samples generated at each leaf node.
+The likelihood of each of these samples (\ref{eqn:t2mle}) is then maximised to give a type II maximum likelihood estimator \(\hat{\theta}\) and the results summarised in Table \ref{tab::ind_simulation}.
+This simulation study indicates that at least \(n = 4\) observations are needed at each leaf node to provide stable estimators \(\hat{\theta}\).
+Thus, when implementing the phylogenetic Ornstein-Uhlenbeck process for bat echolocation call traits, resampling methods must be used to provide multiple estimates for \(\mathcal{E}^{\mathbf{t}}(\cdot)\)
+
+\begin{table}[ht]
+	\tbl{Summary of \(\hat{\theta}\) for 1000 samples reporting the sample mean (standard error) }
+	{
+		\begin{tabular}{@{}cccc@{}} \toprule
+			\(n\)  & \(\hat{\sigma}_p \) & \(\hat{\ell}\) & \(\hat{\sigma}_n\) \\
+			\colrule
+			1 & 1.09 (0.47) & \(10^{14}\) (\(10^{15}\)) & 0.57 (0.54) \\
+			2 & 0.97 (0.29) & \(10^{13}\) (\(10^{14}\)) & 0.99 (0.15) \\
+			4 & 0.97 (0.25) & 63.66 (136.96) & 1.00 (0.09) \\
+			8 & 0.99 (0.24) & 56.21 (48.24) & 1.00 (0.06) \\			
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::ind_simulation}
+\end{table}
 

### 2017-05-19 Paper Drafts/Draft2/Draft2.tex
Submission Version

@@
 
-\chapter[Ancestral Reconstruction of Bat Echolocation Calls]{Ancestral Reconstruction of Bat Echolocation Calls}\label{ra_ch1}
+\chapter[Ancestral Reconstruction of Bat Echolocation Calls]{Phylogenetic Gaussian Processes for the Ancestral Reconstruction of Bat Echolocation Calls}\label{ra_ch1}
 
-\author[J.P. Meagher et al.]{J.P. Meagher\footnote{Author footnote.}}
+\author[J.P. Meagher et al.]{J.P. Meagher, T. Damoulas, K.E. Jones, and M. Girolami\footnote{Author footnote.}}
 %\index[aindx]{Author, F.} % or \aindx{Author, F.}
@@
 \begin{abstract}
- Something something bats something evolution something ancestral reconstruction something magic.
+The reconstruction of ancestral function-valued traits in phylogenetics requires the use computational Statistics for comparative analysis.
+The reconstruction of ancestral bat echolocation calls is an important part of understanding bats natural history.
+General techniques for the ancestral reconstruction of function-valued traits have recently been proposed.
+A full implementation of phylogenetic Gaussian processes for the ancestral reconstruction of function-valued traits representing bat echolocation calls is presented here.
+A phylogenetic signal was found in the data and ancestral reconstruction performed.
+Further interpretation of these results will be required to deduce the full implication of this model on our understanding of the evolution of echolocation in bats.
 \end{abstract}
@@
 \section{Introduction}
+\label{sec:intro}
 The emerging field of Data Science is driven by research which lies at the nexus of Statistics and Computer Science.
@@
 \section{Echolocation Calls as Function-Valued Traits}
+\label{sec:funrep}
 A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time. \cite{ramsay2006functional} 
@@
 
+\begin{figure}
+	\subfigure[]{\includegraphics[width = 0.3\textwidth]{../Figures/callrec.eps} \label{fig:call}}
+	\subfigure[]{\includegraphics[width = 0.3\textwidth]{../Figures/esd.eps} \label{fig:esd}}
+	\subfigure[]{\includegraphics[width = 0.3\textwidth]{../Figures/funrep.eps} \label{fig:funrep}}
+	\caption{A recording of a bat echolocation call (a) along with the log energy spectral density of the call (b) and the smooth functional representation of that spectral density (c).}
+\end{figure}
+
 Given a phylogenetic tree \(\mathbf{T}\), representing the evolutionary relationships between the recorded bat species, we denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of the species observed at point \(\mathbf{t} \in \mathbf{T}\) by \(\{\tilde{x}_{lm}^{\mathbf{t}}(n) : n = 0, \dots, N_{lm}^{\mathbf{t}} - 1\}\).
 Thus, \(\{x_{lm}^{\mathbf{t}}(\cdot)\}\) is a series of discrete measurements of the function \(x_{lm}^{\mathbf{t}}(\cdot)\), the \(m^{th}\) call function for the \(l^{th}\) individual bat, observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording.
-Assume then that \(x_{lm}^{\mathbf{t}}(\cdot) = x_{l}^{\mathbf{t}}(\cdot) + z_{lm}^{\mathbf{t}}(\cdot)\), where \(x_{l}^{\mathbf{t}}(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^{\mathbf{t}}(\cdot)\) is the noise process for the \(m^{th}\) call.
-Further to this, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual.
+Assume then that \(x_{lm}^{\mathbf{t}}(\cdot) = x_{l}^{\mathbf{t}}(\cdot) + z_{lm}^{\mathbf{t}}(\cdot)\), where \(x_{l}^{\mathbf{t}}(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^{\mathbf{t}}(\cdot)\) is the noise process for the \(m^{th}\) call, where \(\mathbf{E}[z_{lm}^{\mathbf{t}}(\cdot)] = 0\).
+Further to this, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual, with \(\mathbf{E}[z_{l}^{\mathbf{t}}(\cdot)] = 0\).
 It is the echolocation call functions at the species level that we re interested in.
@@
 
-Similarly to the call functions, \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\) is the log energy spectral density of the \(m^{th}\) call of the \(l^{th}\) individual from the species at \({\mathbf{t}}\) where \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = \mathcal{E}_{l}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{E}_{l}^{\mathbf{t}}(\cdot) = \mathcal{E}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) where \(\mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) are noise processes.
+Similarly to the call functions, \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\) is the log energy spectral density of the \(m^{th}\) call of the \(l^{th}\) individual from the species at \({\mathbf{t}}\) where \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = \mathcal{E}_{l}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{E}_{l}^{\mathbf{t}}(\cdot) = \mathcal{E}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) where \(\mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) are noise processes, each with ae expected value of zero.
 The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \(F = [0, \frac{f_S}{2}]\).\cite{antoniou2006digital}
@@
 
-We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \mathcal{E}_{lm}^{\mathbf{t}}(f)\}\) and \(\{f, \mathcal{E}_{l'm'}^{{\mathbf{t}}'}(f)\}\) are directly comparable.
+We now have a functional representation of each bats echolocation call where the pairs of observations \(\{f, \hat{\mathcal{E}}_{lm}^{\mathbf{t}}(f)\}\) and \(\{f, \hat{\mathcal{E}}_{l'm'}^{{\mathbf{t}}'}(f)\}\) are directly comparable.
 These function-valued traits can now be modelled for evolutionary inference.
@@
 \section{Phylogenetic Gaussian Processes}
+\label{sec:pgp}
 A Gaussian Process is a collection of random variables, any finite number of which have a joint Gaussian distribution.
@@
 Modelling this as Gaussian process function where \(\mathcal{E}^{\mathbf{t}}(f)\) corresponds to a point \((f, \mathbf{t})\) on the frequency-phylogeny \(F \times \mathbf{T}\) requires that a suitable phylogenetic covariance function, \(\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right)\), is defined.
+
+\begin{figure}[h]
+	\subfigure{\includegraphics[width = 0.45\textwidth]{../Figures/tree3}}
+	\subfigure{\includegraphics[width = 0.45\textwidth]{../Figures/unipgp}}
+	\caption{An example of a Phylogenetic Tree and a simulated univariate phylogenetic Gaussian process of that tree.}
+\end{figure}
 
@@
 
+\begin{figure}[h]
+	\centering
+	\includegraphics[width = 0.9\textwidth]{../Figures/basis.eps}
+	\caption{Set of basis functions identified}
+\end{figure}
+
 Each column of \(X_{\mathbf{T}}\) is an independent, univariate, phylogenetic Gaussian process, \(X_i^{\mathbf{T}}(\mathcal{t})\), modelled here with phylogenetic Ornsetin-Uhlenbeck process kernel.
@@
 \right)^{-1} X_i^{\mathcal{E}} (\mathcal{t}) \phi_i(\cdot) \right)
+\label{eqn:postmean}
 \end{equation}
@@
 \right)^{-1} \Sigma_{\mathbf{T}}^i(\mathbf{t}^*, \mathcal{t})^{\mathsf{T}} \right) \phi_i (\cdot)
+\label{eqn:postvar}
 \end{equation}
@@
 
-\subsection{Hyperparameter Estimation for Phylogenetic Gaussian Processes}
+\subsection{Hyperparameter Estimation and Ancestral Trait Reconstruction with Phylogenetic Gaussian Processes}
 
@@
 \begin{table}[ht]
-	\tbl{Summary of \(\hat{\theta}\) for 1000 samples reporting the sample mean (standard error) }
+	\tbl{Summary of \(\hat{\theta}\) for 1000 simulations of independent Ornstein-Uhlenbeck processes with \(\theta = [1,50,1]^{\mathsf{T}}\) reporting: sample mean (standard error) }
 	{
@@
 			4 & 0.97 (0.25) & 63.66 (136.96) & 1.00 (0.09) \\
-			8 & 0.99 (0.24) & 56.21 (48.24) & 1.00 (0.06) \\			
+			8 & 0.99 (0.24) & 56.21 ( 48.24) & 1.00 (0.06) \\			
 			\botrule
@@
 \end{table}
+
+Given the modelling assumptions made in Section \ref{sec:funrep} the best estimator for \(\mathcal{E}^{\mathbf{t}}(\cdot)\) is the sample mean given by 
+\begin{equation}
+\hat{\mathcal{E}}^{\mathbf{t}}~(\cdot) = \frac{1}{l_{\mathbf{t}}} \sum_{l = 1}^{l_{\mathbf{t}}} \frac{1}{m_l} \sum_{m = 1}^{m_l} \hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)
+\label{eqn:est}
+\end{equation}
+where \(m_l\) is the total number of recordings for the \(l^{th}\) individual and \(l_{\mathbf{t}}\) is the number of individuals recorded from the species at \(\mathbf{t} \in \mathbf{T}\).
+However, simply calculating \(\hat{\mathcal{E}}^{\mathbf{t}}~(\cdot)\) for each species performing the analysis laid out in Section \ref{sec:pgp} means that we have too few datapoints to obtain a stable model.
+For this reason we implement a resampling procedure to leverage more datapoints from the dataset in order to produce stable estimates for the model parameters.
+
+A resampled estimator \(\hat{\mathcal{E}}_r^{\mathbf{t}}(\cdot)\) is obtained by sampling at random one call from \(n_r\) individuals of the species at \(\mathbf{t}\) and calculating the arithmetic mean of the sample, similarly to (\ref{eqn:est}).
+This can be repeated to create an arbitrary number of estimates for \(\mathcal{E}^{\mathbf{t}}\).
+Resampling across all the species in the dataset we create a resampled dataset \(\hat{\mathcal{E}}_r = [\hat{\mathcal{E}}_{r,1}^{\mathbf{t}_1}(\mathbf
+f), \hat{\mathcal{E}}_{r,2}^{\mathbf{t}_1}(\mathbf
+f), \dots, \hat{\mathcal{E}}_{r,1}^{\mathbf{t}_2}(\mathbf
+f), \dots]\), where \(\mathbf{f}\) is the vector of frequencies over which \(\hat{\mathcal{E}}_{r}^{\mathbf{t}_2}(
+\cdot)\) is sampled. The methods outlined in Section \ref{sec:pgp} can then be applied to each resampled \(\hat{\mathcal{E}}_r\)
+
+Our analysis took \(n_r = 4\) and included 4 samples of \(\hat{\mathcal{E}}_r^{\mathbf{t}}(\mathbf{f})\) in each \(\hat{\mathcal{E}}_r\).
+This reflected the structure of the dataset, for which the minimum number of individuals per species was 4, and the results of the simulations study which showed that 4 observations per species provided reasonably stable estimates for \(\theta\).
+Note also that \(\mathbf{f} = [9,10, \dots, 212]^{\mathsf{T}}\), which reflects the spectrum of frequencies over which bats emit echolocation calls.
+\(\hat{\phi}_F\) was obtained by averaging the basis identified over all \(\hat{\mathcal{E}}_r\) and applying a single set of basis functions to each dataset.
+Thus \(\hat{X}_r\), the matrix of mixing coefficients described by (\ref{eqn:inv}), the columns of which are modelled a phylogenetic Ornstein-Uhlenbeck processes, is obtained for each \(\hat{\mathcal{E}}_r\).
+\(\hat{\theta}_r\) is then the type II maximum likelihood estimator of (\ref{eqn:t2mle}) given \(\hat{\mathcal{E}}_r\).
+Table \ref{tab::bat_results} presents the results of the hyperparameter estimation procedure.
+
+\begin{table}[ht]
+	\tbl{Summary of \(\hat{\theta}_r\) over 1000 \(\hat{\mathcal{E}_r}\) samples reporting: sample mean (standard error)}
+	{
+		\begin{tabular}{@{}cccc@{}} \toprule
+			Basis & \(\hat{\sigma}_p\) & \(\hat{\ell}\) & \(\hat{\sigma}_n\) \\
+			\colrule
+			1 & 2.30 (0.11) & 12.27 ( 4.18) & 1.18 (0.11) \\
+			2 & 3.17 (0.11) & 27.63 ( 3.70) & 1.26 (0.13) \\
+			3 & 4.05 (0.32) & 70.50 (20.31) & 1.19 (0.12) \\
+			4 & 3.32 (0.17) & 22.86 ( 8.95) & 1.96 (0.19) \\
+			5 & 3.00 (0.13) & 26.93 ( 2.85) & 1.21 (0.11) \\
+			6 & 3.70 (0.14) & 12.82 ( 4.52) & 1.28 (0.15) \\
+			\botrule
+		\end{tabular}
+	}
+	\label{tab::bat_results}
+\end{table}
+
+Ancestral reconstruction by a phylogenetic Gaussian process involves obtaining the posterior predictive distribution of the trait at the ancestral node \(\mathbf{t}^* \in \mathbf{T}\) given by (\ref{eqn:postmean}) and (\ref{eqn:postvar}).
+
+To perform ancestral trait reconstruction for \(\mathcal{E}^{\mathbf{t}^*}(\cdot)\) the species level traits are estimated by (\ref{eqn:est}) and and the model hyperparameters by the mean values of \(\theta_r\) reported in Table \ref{tab::bat_results}.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width = 0.9\textwidth]{../Figures/ancrec.eps}
+	\caption{Ancestral Reconstruction}
+\end{figure}
+
+\section{Conclusions and Further Work}
+
+This analysis has developed a limited model for the evolution of echolocation in bats and identified a phylogenetic signal which allows the construction of a posterior predictive distribution for ancestral traits.
+The log energy spectral density has been identified as a trait representative of the echolocation call in bats.
+This trait, representing the energy intensity of the call across the frequency spectrum, is modelled as a series of independent components, combinations of energy intensities across the spectrum, each of which evolves according to a phylogenetic Ornstein-Uhlenbeck process.
+Estimating the hyperparameters governing these Ornstein-Uhlenbeck processes from observed traits provides an insight into the evolutionary path of these traits.
+Each of the hyperparameters has an intuitive interpretation where \(\frac{\sigma_p}{\sigma_p + \sigma_n}\) indicates the proportion of variation in the sample attributable solely to the evolutionary path while \(\ell\) provides a measure of how quickly correlation between observations along the evolutionary path decays.
+We are working towards understanding what the results of this analysis mean for with respect to the evolution of echolocation in bats.
+
+One particular limitation of the model is the representation of the echolocation call by a log energy spectral density.
+Echolocation calls have complex spectral and temporal structures, much of which is lost in the log energy spectral density representation.
+An alternative representation, which preserves more of this structure, is the spectrogram.
+Implementing this model using the spectrogram as a functional representation of the echolocation call is a priority, as the results of such an analysis will hopefully be more meaningful.
 

### 2017-05-19 Paper Drafts/submissionDraft/submissionDraft.tex
Submission Version

--- 2017-05-19 Paper Drafts/Draft2/Draft2.tex
@@
 %%
-\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
-%\documentclass{ws-rv9x6}
+%\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
+\documentclass{ws-rv9x6}
 \usepackage{subfigure}   % required only when side-by-side / subfigures are used
@@
 \begin{abstract}
-The reconstruction of ancestral function-valued traits in phylogenetics requires the use computational Statistics for comparative analysis.
-The reconstruction of ancestral bat echolocation calls is an important part of understanding bats natural history.
+The reconstruction of ancestral echolocation calls is an important part of understanding the evolutionary history of bats.
 General techniques for the ancestral reconstruction of function-valued traits have recently been proposed.
@@
 A phylogenetic signal was found in the data and ancestral reconstruction performed.
-Further interpretation of these results will be required to deduce the full implication of this model on our understanding of the evolution of echolocation in bats.
+This promising preliminary analysis paves the way for more realistic models of the evolution of echolocation in bats.
 \end{abstract}
@@
 
-Monitoring bat populations provides useful information, but an understanding the orders natural history is required to identify the cause and effect of any changes observed.
+Monitoring bat populations provides useful information, but an understanding the orders evolutionary history is required to identify the cause and effect of any changes observed.
 The echolocation call structure, which reflects a bats diet and habitat,\cite{aldridge1987morphology}  is a key aspect of this natural history.
 Reconstructing ancestral traits\cite{joy2016ancestral} relies on a statistical comparative analysis incorporating extant species and fossil records.
-\cite{felsenstein2004inferring} However, the fossil record is of limited use in inferring ancestral bats echolocation calls.
+\cite{felsenstein2004inferring} However, the fossil record is of limited use in inferring ancestral echolocation calls in bats.
 Therefore, statistical data science techniques may shed some light on this topic.
@@
 Previous studies of bat echolocation calls for both classification \cite{walters2012continental} and ancestral reconstruction \cite{collen2012evolution} analysed features extracted from the call spectrogram.
-These call features relied domain knowledge to ensure they were sensibly selected and applied.
+These call features relied upon domain knowledge to ensure they were sensibly selected and applied.
 More recently, general techniques for the classification of acoustic signals have been developed.
 \cite{stathopoulos2014bat} \cite{damoulas2010bayesian} General techniques for the ancestral reconstruction of function-valued traits have also been proposed.
-\cite{group2012phylogenetic} This piece of research applies some proposed techniques to bat echolocation calls.
+\cite{group2012phylogenetic} Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits\cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis et al.\cite{hadjipantelis2013function}.
+This current research investigates these techniques in the context of the evolution of bat echolocation calls.
 
-A function-valued trait\cite{meyer2005up} is a characteristic of some organism measured along some continuous scale, usually time, and can be modelled as a continuous mathematical function by functional data analysis.
-\cite{ramsay2006functional} Jones \& Moriarty \cite{jones2013evolutionary} developed a method which extends Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits over a phylogeny.
-A full demonstration of the phylogenetic Gaussian Process method for ancestral reconstruction on a synthetic dataset is presented by Hajipantelis et al.\cite{hadjipantelis2013function}
-
-This general approach to evolutionary inference for function-valued traits is implemented here for a set of bat echolocation calls.
-Our goal in doing so is twofold.
-These techniques had previously been considered in the context of modelling the evolution of human speech sounds in language.
-\cite{group2012phylogenetic} It is hoped that by applying these methods in a simpler context progress can be made towards resolving methodological problems.
-For example, how do we extend these methods to more realistic models of evolution?
-
-We are also interested in what specifically these models tell us about bats and the evolutionary dynamics of echolocation.
-What impact might these results have on our understanding of ancestral bats and their behaviour?
-
-This paper presents the early stages of our research and some preliminary results.
-
+Section \ref{sec:funrep} presents details on representing echolocation call recordings as function-valued traits, and Section \ref{sec:pgp} then illustrates the model for evolution developed by a phylogenetic Gaussian Process.
+The results of our analysis of bat echolocation calls are then presented and discussed.
 \section{Echolocation Calls as Function-Valued Traits}
@@
 \begin{figure}
+	\centering
 	\subfigure[]{\includegraphics[width = 0.3\textwidth]{../Figures/callrec.eps} \label{fig:call}}
@@
 	\subfigure[]{\includegraphics[width = 0.3\textwidth]{../Figures/funrep.eps} \label{fig:funrep}}
-	\caption{A recording of a bat echolocation call (a) along with the log energy spectral density of the call (b) and the smooth functional representation of that spectral density (c).}
+	\caption{A recording of a bat echolocation call (a) along with the log energy spectral density of the call (b) and the smooth functional representation of that spectral density restricted to the range \([9-212]\)kHz (c).}
 \end{figure}
 
-Given a phylogenetic tree \(\mathbf{T}\), representing the evolutionary relationships between the recorded bat species, we denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of the species observed at point \(\mathbf{t} \in \mathbf{T}\) by \(\{\tilde{x}_{lm}^{\mathbf{t}}(n) : n = 0, \dots, N_{lm}^{\mathbf{t}} - 1\}\).
-Thus, \(\{x_{lm}^{\mathbf{t}}(\cdot)\}\) is a series of discrete measurements of the function \(x_{lm}^{\mathbf{t}}(\cdot)\), the \(m^{th}\) call function for the \(l^{th}\) individual bat, observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording.
-Assume then that \(x_{lm}^{\mathbf{t}}(\cdot) = x_{l}^{\mathbf{t}}(\cdot) + z_{lm}^{\mathbf{t}}(\cdot)\), where \(x_{l}^{\mathbf{t}}(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^{\mathbf{t}}(\cdot)\) is the noise process for the \(m^{th}\) call, where \(\mathbf{E}[z_{lm}^{\mathbf{t}}(\cdot)] = 0\).
-Further to this, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual, with \(\mathbf{E}[z_{l}^{\mathbf{t}}(\cdot)] = 0\).
+Given a phylogenetic tree \(\mathbf{T}\), representing the evolutionary relationships between the recorded bat species, we denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of the species observed at point \(\mathbf{t} \in \mathbf{T}\) by \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}_{n=0}^{N_{lm}^{\mathbf{t}} - 1}\).
+Thus, \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}\) is a series of discrete measurements of the function \(x_{lm}^{\mathbf{t}}(\cdot)\), observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording.
+Assume then that \(x_{lm}^{\mathbf{t}}(\cdot) = x_{l}^{\mathbf{t}}(\cdot) + z_{lm}^{\mathbf{t}}(\cdot)\), where \(x_{l}^{\mathbf{t}}(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^{\mathbf{t}}(\cdot)\) is the noise process for the \(m^{th}\) call.
+Further, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual.
 It is the echolocation call functions at the species level that we re interested in.
@@
 The energy spectral density of \(x_{lm}^{\mathbf{t}}(\cdot)\) is the magnitude of the Fourier transform and the log energy spectral density is given by
-\begin{equation}
+\[
 \mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = 10 \log_{10} \left( |X_{lm}^{\mathbf{t}}(\cdot)|^2 \right).
-\label{eqn:esd}
-\end{equation}
+\]
 
-Similarly to the call functions, \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\) is the log energy spectral density of the \(m^{th}\) call of the \(l^{th}\) individual from the species at \({\mathbf{t}}\) where \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = \mathcal{E}_{l}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{E}_{l}^{\mathbf{t}}(\cdot) = \mathcal{E}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) where \(\mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) are noise processes, each with ae expected value of zero.
+Similarly to the call functions, \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\) is the log energy spectral density of the \(m^{th}\) call of the \(l^{th}\) individual from the species at \({\mathbf{t}}\) where \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = \mathcal{E}_{l}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{E}_{l}^{\mathbf{t}}(\cdot) = \mathcal{E}^{\mathbf{t}}(\cdot) + \mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) where \(\mathcal{Z}_{lm}^{\mathbf{t}}(\cdot)\) and \(\mathcal{Z}_{l}^{\mathbf{t}}(\cdot)\) are noise processes, each with an expected value of zero.
 The log energy spectral density is a periodic function of frequency which describes the energy of a signal at each frequency on the interval \(F = [0, \frac{f_S}{2}]\).\cite{antoniou2006digital}
@@
 The discrete Fourier Transform\cite{antoniou2006digital}
-of \(\{\tilde{x}_{lm}^{\mathbf{t}}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^{\mathbf{t}}(k) : k = 0, \dots, \frac{ N_{lm}^{\mathbf{t}}}{2} + 1\}\).
+of \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^{\mathbf{t}}(k) : k = 0, \dots, \frac{ N_{lm}^{\mathbf{t}}}{2} + 1\}\).
 Smoothing splines\cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\).
@@
 \label{sec:pgp}
-A Gaussian Process is a collection of random variables, any finite number of which have a joint Gaussian distribution.
-A Gaussian process prior can be defined as a distribution over functions, \(f(x) \sim \mathcal{GP}(m(x), k (x, x'),\) where \(x \in \mathbf{R}^P\) is some input variable, the mean function \(m(x) = \mathbf{E}[f(x)]\), and the covariance kernel \(k(x, x') = \mathnormal{cov}(f(x), f(x') )\).
-Any collection of function values has a joint Gaussian distribution
-\(
-[f(x_1), f(x_2), \dots f(x_N)]^{\mathsf{T}} \sim \mathcal{N}(\)\boldmath\(\mu\)\unboldmath, \(\mathbf{K}
-\))
-where the \(N \times N\) covariance matrix \(\mathbf{K}\)  has the entries \(K_{ij} = k(x_i, x_j)\), and the mean \boldmath\(\mu\)\unboldmath\(\) has entries \(\mu_i = m(x_i)\).
-Thus, the properties of the functions are determined by the kernel function.
-Assuming that the function values \(\mathbf{y}\) observed at locations \(\{x_n\}_{n=1}^N\) are subject to Gaussian noise, and given the kernel hyperparameters \(\theta\), a posterior predictive distribution over Gaussian process functions can be inferred analytically.
-A marginal likelihood \(p(\mathbf{y} | \theta, \{x_n\}_{n=1}^N)\), which can in turn be used to estimate the kernel hyperparameters, can also be derived analytically.
-An in depth treatment of Gaussian processes is given by Rasmussen \& Williams.\cite{rasmussen2006gaussian}
+Gaussian Processes (GP) provide a concise way of fitting a prior belief over the distribution of functions.
+The structure of this prior belief is described by the GP model kernel, defining it's covariance function.
+GPs are useful for classification and regression problems, and allow posterior predictive distributions to be obtained.
+An in depth treatment of GPs is given by Rasmussen \& Williams.\cite{rasmussen2006gaussian}
 
-Jones \& Moriarty\cite{jones2013evolutionary} extend Gaussian processes for inference on function-valued traits over a phylogeny.
+Jones \& Moriarty\cite{jones2013evolutionary} extend GPs for the inference of function-valued traits over a phylogeny.
 Consider \(\mathcal{E}^{\mathbf{t}}(\cdot)\), a functional representation of the echolocation call of the species observed at the point \(\mathbf{t}\) on the phylogenetic tree \(\mathbf{T}\) with respect to frequency.
-Modelling this as Gaussian process function where \(\mathcal{E}^{\mathbf{t}}(f)\) corresponds to a point \((f, \mathbf{t})\) on the frequency-phylogeny \(F \times \mathbf{T}\) requires that a suitable phylogenetic covariance function, \(\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right)\), is defined.
+Modelling this as GP function, where \(\mathcal{E}^{\mathbf{t}}(f)\) corresponds to a point \((f, \mathbf{t})\) on the frequency-phylogeny \(F \times \mathbf{T}\), requires that a suitable phylogenetic covariance function, \(\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right)\), is defined.
 
 \begin{figure}[h]
-	\subfigure{\includegraphics[width = 0.45\textwidth]{../Figures/tree3}}
-	\subfigure{\includegraphics[width = 0.45\textwidth]{../Figures/unipgp}}
-	\caption{An example of a Phylogenetic Tree and a simulated univariate phylogenetic Gaussian process of that tree.}
+	\subfigure[]{\includegraphics[width = 0.45\textwidth]{../Figures/tree3} }
+	\subfigure[]{\includegraphics[width = 0.45\textwidth]{../Figures/unipgp} }
+	\caption{A sub tree from the full phylogeny \(\mathbf{T}\) (a) and a simulated univariate phylogenetic OU GP over that sub tree (b).}
+	\label{fig:subtree}
 \end{figure}
@@
 
-For a phylogenetic Gaussian Process \(Y\) with covariance function given by \ref{eqn:phy}, when \(K\) is a degenerate Mercer kernel, there exists a set of \(n\) deterministic basis functions \(\phi_i: F \to \mathbf{R}\) and univariate Gaussian processes \(X_i\) for \(i = 1,\dots, n\) such that
+For a phylogenetic Gaussian Process \(Y\) with covariance function given by (\ref{eqn:phy}), when \(K\) is a degenerate Mercer kernel, there exists a set of \(n\) deterministic basis functions \(\phi_i: F \to \mathbf{R}\) and univariate GPs \(X_i\) for \(i = 1,\dots, n\) such that
 \[
@@
 \] 
-has the same distribution as \(Y\). The full phylogenetic covariance function of this phylogenetic Gaussian process is
+has the same distribution as \(Y\). The full phylogenetic covariance function of this phylogenetic GP is
 \[
@@
 \]
-where \(\int \phi_i(f) \phi_j(f) df = \delta_{ij}\), \(\delta\) being the Kronecker delta, and so the phylogenetic covariance function depends only on \(\mathbf{T}\).
+where \(\int \phi_i(f) \phi_j(f) df = \delta_{ij}\), \(\delta\) being the Kronecker delta, and so the phylogenetic covariance function depends only on \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\).
 
@@
 The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis et al.
-\cite{hadjipantelis2013function} \(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott\cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for samples of traits balanced by species.
+\cite{hadjipantelis2013function} \(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott\cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for the observed traits.
 Given \(\hat{\phi}_F\), the estimated matrix of mixing coefficients is \(\hat{X}_{\mathbf{T}} = \mathcal{E} (\hat{\phi}_F^{\mathsf{T}})^{-1}\).
@@
 	\includegraphics[width = 0.9\textwidth]{../Figures/basis.eps}
-	\caption{Set of basis functions identified}
+	\caption{Set of independently evolving basis functions identified for bat echolocation calls. }
 \end{figure}
 
-Each column of \(X_{\mathbf{T}}\) is an independent, univariate, phylogenetic Gaussian process, \(X_i^{\mathbf{T}}(\mathcal{t})\), modelled here with phylogenetic Ornsetin-Uhlenbeck process kernel.
+Each column of \(X_{\mathbf{T}}\) is an independent, univariate, phylogenetic GP, \(X_i^{\mathbf{T}}(\mathcal{t})\), modelled here with phylogenetic Ornstein-Uhlenbeck (OU) process kernel.
 
-The phylogenetic Ornstein-Uhlenbeck process is defined by the kernel
+The phylogenetic OU process is defined by the kernel
 \begin{equation}
@@
 
-where \(\delta\) is the Kronecker delta, \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) is the cophenetic distance between \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), and \(\mathbf{\theta}^i = [\sigma_p^i, \ell^i, \sigma_n^i]^{\mathsf{T}}\) is the vector of hyperparameters for \(X_i^{\mathbf{T}}(\cdot)\).
+where \(\delta\) is the Kronecker delta, \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) is the distance along \(\mathbf{T}\) between \(\mathbf{t}\) and \(\mathbf{t}' \in \mathbf{T}\), and \(\mathbf{\theta}^i = [\sigma_p^i, \ell^i, \sigma_n^i]^{\mathsf{T}}\) is the vector of hyperparameters for \(X_i^{\mathbf{T}}(\cdot)\).
 The phylogenetic covariance matrix for \(X_i^{\mathbf{T}}(\mathcal{t})\) is denoted \(\Sigma_{\mathbf{T}}^i(\mathcal{t}, \mathcal{t})\) and the marginal likelihood of the observed data given \(\theta\) is
@@
 The variance of observations in the sample is \(\sigma_p + \sigma_n\), where \(\sigma_p\) is the phylogenetic noise, and \(\sigma_n\) is the non-phylogenetic noise.
-\(\sigma_p\) is the proportion of variation which can be accounted for by the cophenetic distance between any \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), while \(\sigma_n\) accounts for other sources of variation which may be observed at \(\mathbf{t}\).
-The length-scale parameter, \(\ell\) then indicates the strength of the correlation along \(\mathbf{T}\), where large values of \(\ell\) indicate a slowly decaying correlation.
+\(\sigma_p\) is the proportion of variation which can be accounted for by the distance between any \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), while \(\sigma_n\) accounts for other sources of variation which may be observed at \(\mathbf{t}\).
+The length-scale parameter, \(\ell\) then indicates the strength of the correlation along \(\mathbf{T}\), where large values of \(\ell\) indicate a correlation that decays slowly as \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) increases.
 
@@
 			1 \textit{Balantiopteryx plicata} & Bapl & 16 & 100 \\
-			\\
+			\colrule
 			Family: Molossidae &&& \\
@@
 			3 \textit{Tadarida brasiliensis} & Tabr & 49 & 100  \\
-			\\
+			\colrule
 			Family: Vespertilionidae &&& \\
@@
 			12 \textit{Pipistrellus hesperus} & Pihe & 85 & 100 \\
-			\\
+			\colrule
 			Family: Mormoopidae &&& \\
@@
 			16 \textit{Pteronotus personatus} & Ptpe & 7 & 51 \\
-			\\
+			\colrule
 			Family: Phyllostomidae &&& \\
@@
 
-We are interested in modelling the evolution of \(\mathcal{E}^{\mathbf{t}}(\cdot)\), the function valued trait representing the echolocation call of the species of bat observed at \(\mathbf{t} \in \mathbf{T}\).
+We are interested in modelling the evolution of \(\mathcal{E}^{\mathbf{t}}(\cdot)\), the function valued trait representing \(x_{lm}^{\mathbf{t}}(\cdot)\).
 However, only 22 species of bat are represented in \(\mathbf{T}\).
@@
 
-Although there are only 22 leaf nodes of the phylogeny \(\mathbf{T}\), we are not limited to a single observation at any given \(\mathbf{t}\).
-By sampling at each leaf node of \(\mathbf{T}\) multiple times, larger samples can be obtained, improving the quality of the estimators \(\hat{\theta}\).
-With this in mind, 1000 independent, univariate phylogenetic Gaussian processes were simulated for each of \(n = \{1,2,4,8\}\) according to the kernel (\ref{eqn:oukernel}) with \(\theta = [1,50,1]^{\mathsf{T}}\), where \(n\) is the number of samples generated at each leaf node.
+We are not limited to a single observation at any given \(\mathbf{t}\).
+By repeatedly sampling at each observed \(\mathbf{t}\), larger samples can be obtained, improving the quality of the estimators \(\hat{\theta}\).
+With this in mind, 1000 independent, univariate phylogenetic GPs were simulated for each of \(n = \{1,2,4,8\}\) according to the kernel (\ref{eqn:oukernel}) with \(\theta = [1,50,1]^{\mathsf{T}}\), where \(n\) is the number of samples generated at each leaf node.
 The likelihood of each of these samples (\ref{eqn:t2mle}) is then maximised to give a type II maximum likelihood estimator \(\hat{\theta}\) and the results summarised in Table \ref{tab::ind_simulation}.
 This simulation study indicates that at least \(n = 4\) observations are needed at each leaf node to provide stable estimators \(\hat{\theta}\).
-Thus, when implementing the phylogenetic Ornstein-Uhlenbeck process for bat echolocation call traits, resampling methods must be used to provide multiple estimates for \(\mathcal{E}^{\mathbf{t}}(\cdot)\)
+Thus, when implementing the phylogenetic OU process for bat echolocation call traits, resampling methods must be used to provide multiple estimates for \(\mathcal{E}^{\mathbf{t}}(\cdot)\)
 
 \begin{table}[ht]
-	\tbl{Summary of \(\hat{\theta}\) for 1000 simulations of independent Ornstein-Uhlenbeck processes with \(\theta = [1,50,1]^{\mathsf{T}}\) reporting: sample mean (standard error) }
+	\tbl{Summary of \(\hat{\theta}\) for 1000 simulations of independent OU processes with \(\theta = [1,50,1]^{\mathsf{T}}\) reporting: sample mean (standard error) }
 	{
@@
 where \(m_l\) is the total number of recordings for the \(l^{th}\) individual and \(l_{\mathbf{t}}\) is the number of individuals recorded from the species at \(\mathbf{t} \in \mathbf{T}\).
-However, simply calculating \(\hat{\mathcal{E}}^{\mathbf{t}}~(\cdot)\) for each species performing the analysis laid out in Section \ref{sec:pgp} means that we have too few datapoints to obtain a stable model.
-For this reason we implement a resampling procedure to leverage more datapoints from the dataset in order to produce stable estimates for the model parameters.
+However, Table \ref{tab::ind_simulation} indicates that 22 samples is not enough to obtain a stable \(\hat{\theta}\) by type II maximum likelihood estimation.
+We implement a resampling procedure to leverage more estimates from the dataset in order to produce more stable \(\hat{\theta}\).
 
@@
 \(\hat{\phi}_F\) was obtained by averaging the basis identified over all \(\hat{\mathcal{E}}_r\) and applying a single set of basis functions to each dataset.
-Thus \(\hat{X}_r\), the matrix of mixing coefficients described by (\ref{eqn:inv}), the columns of which are modelled a phylogenetic Ornstein-Uhlenbeck processes, is obtained for each \(\hat{\mathcal{E}}_r\).
+Thus \(\hat{X}_r\), the matrix of mixing coefficients described by (\ref{eqn:inv}), the columns of which are modelled a phylogenetic OU processes, is obtained for each \(\hat{\mathcal{E}}_r\).
 \(\hat{\theta}_r\) is then the type II maximum likelihood estimator of (\ref{eqn:t2mle}) given \(\hat{\mathcal{E}}_r\).
@@
 
-Ancestral reconstruction by a phylogenetic Gaussian process involves obtaining the posterior predictive distribution of the trait at the ancestral node \(\mathbf{t}^* \in \mathbf{T}\) given by (\ref{eqn:postmean}) and (\ref{eqn:postvar}).
+Ancestral reconstruction by a phylogenetic GP involves obtaining the posterior predictive distribution of the trait at the ancestral node \(\mathbf{t}^* \in \mathbf{T}\) given by (\ref{eqn:postmean}) and (\ref{eqn:postvar}).
 
@@
 	\includegraphics[width = 0.9\textwidth]{../Figures/ancrec.eps}
-	\caption{Ancestral Reconstruction}
+	\caption{Ancestral Reconstruction of the function valued trait representing the echolocation calls of the bat species included in the sub tree shown in Figure  \ref{fig:subtree} (a).
+Grey shaded region represents one standard deviation of variation around \(\hat{\mathcal{E}}^{\mathbf{t}}(\cdot)\).}
 \end{figure}
@@
 
-This analysis has developed a limited model for the evolution of echolocation in bats and identified a phylogenetic signal which allows the construction of a posterior predictive distribution for ancestral traits.
+This preliminary analysis has developed a limited model for the evolution of echolocation in bats and identified a phylogenetic signal which allows the construction of a posterior predictive distribution for ancestral traits.
 The log energy spectral density has been identified as a trait representative of the echolocation call in bats.
@@
 Each of the hyperparameters has an intuitive interpretation where \(\frac{\sigma_p}{\sigma_p + \sigma_n}\) indicates the proportion of variation in the sample attributable solely to the evolutionary path while \(\ell\) provides a measure of how quickly correlation between observations along the evolutionary path decays.
-We are working towards understanding what the results of this analysis mean for with respect to the evolution of echolocation in bats.
+We are working towards understanding what the results of this analysis could mean with respect to the evolution of echolocation in bats.
 
@@
 An alternative representation, which preserves more of this structure, is the spectrogram.
-Implementing this model using the spectrogram as a functional representation of the echolocation call is a priority, as the results of such an analysis will hopefully be more meaningful.
+Implementing this model using the spectrogram as a functional representation of the echolocation call is a priority, as the results of such an analysis are expected to be more meaningful.
 

### 2017-05-19 Paper Drafts/submissionDraft/ws-rv9x6.tex
Submission Version

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Review Volume (last updated on 20-4-2015)                            %%
%% Trim Size: 9in x 6in                                                 %%
%% Text Area: 7.35in (include runningheads) x 4.5in                     %%
%% Main Text: 10 on 13pt                                                %%
%% For support: Yolande Koh, <ykoh@wspc.com.sg>                         %%
%%              D. Rajesh Babu, <rajesh@wspc.com.sg>                    %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
\documentclass{ws-rv9x6}
\usepackage{subfigure}   % required only when side-by-side / subfigures are used
\usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
\usepackage{ws-rv-van}   % numbered citation & references (default)
%\usepackage{ws-index}   % to produce multiple indexes
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}       % author index
%\renewindex{default}{idx}{ind}{Subject Index}  % subject index

\begin{document}

\chapter[Using World Scientific's Review Volume Document Style]{Using World Scientific's Review Volume\\ Document Style}\label{ra_ch1}

\author[F. Author and S. Author]{First Author and Second Author\footnote{Author footnote.}}
%\index[aindx]{Author, F.} % or \aindx{Author, F.}
%\index[aindx]{Author, S.} % or \aindx{Author, S.}

\address{World Scientific Publishing Co, Production Department,\\
5 Toh Tuck Link, Singapore 596224, \\
f\_author@wspc.com.sg\footnote{Affiliation footnote.}}

\begin{abstract}
The abstract should summarize the context, content and conclusions of
the paper in less than 200 words. It should not contain any references
or displayed equations. Typeset the abstract in 9 pt Times roman with
baselineskip of 11 pt, making an indentation of 1.5 pica on the left
and right margins.
\end{abstract}
%\markright{Customized Running Head for Odd Page} % default is Chapter Title.
\body

%\tableofcontents

\section{Using Other Packages}\label{ra_sec1}
The WSPC class file has already loaded the packages \verb|amsfonts,|
\verb|amsmath, amssymb, graphicx, rotating| and \verb|url| at the
startup. \verb|Check.tex| is an utility to test for all the files
required by World Scientific review volume project are available in
your present \LaTeX\ installation.

{\bf Usage:} \verb|latex check.tex|

Please try to limit your use of additional packages. They frequently
introduce incompatibilities. This problem is not specific to the
WSPC styles, it is a general \LaTeX{} problem. Check this manual
whether the required functionality is already provided by the WSPC
class file. If you do need third-party packages, send them along
with the paper. In general, you should use standard \LaTeX{}
commands as much as possible.

\section{Layout}
In order to facilitate our processing of your article, please give
easily identifiable structure to the various parts of the text by
making use of the usual \LaTeX{} commands or by your own commands
defined in the preamble, rather than by using explicit layout
commands, such as \verb|\hspace, \vspace, \large|,
etc. Also, do not redefine the page-layout parameters. For more
information on layout and font please refer
\url{http://www.worldscientific.com/sda/1020/rv-instruction9x6_2e.pdf}.

\subsection{Input used to produce a chapter}
\begin{verbatim}
\documentclass{ws-rv9x6}
%\usepackage{subfigure}% to produce side-by-side / subfigures
\usepackage{ws-rv-thm}% comment when other thm package is used
\usepackage{ws-rv-van}% default - numbered citation/references
%\usepackage{ws-index}% to produce multiple indexes
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}      % author index
%\renewindex{default}{idx}{ind}{Subject Index} % subject index
\begin{document}
\chapter[Short Title]{Full Title}
\author[F. Author]{First Author}
%\aindx{Author, F.}                       % author index entry
\address{World Scientific Publishing...}
\begin{abstract}
The abstract should ...
\end{abstract}
\body
\section{Using Other Packages}
The class file has...
\begin{appendix}[Optional Appendix Title]
\section{Sample Appendix}
Text...
\end{appendix}
%\begin{thebibliography}{9}        % for non BIBTeX users
%\bibitem{ams04} \AmS, \emph{\AmS-\LaTeX{} ...
%\end{thebibliography}
\bibliographystyle{ws-rv-van}
\bibliography{ws-rv-sample}
%\printindex[aindx]                % to print author index
\printindex
\end{document}
\end{verbatim}

\section{Using ws-rv9x6}
You can obtain these files from our web pages at:

\begin{itemlist}
\item \url{http://www.worldscientific.com/page/authors/review-style} and
\item \url{http://www.icpress.co.uk/authors/stylefiles.shtml#review}.
\end{itemlist}

\subsection{Class options and extra packages}
The class file, \verb|ws-rv9x6.cls| provides the following options:
\begin{alphlist}[onethmnum]
\item [\tt wsdraft] To draw border line around text area.\\
Default: no border line around text area.
\item[\tt addchapnum] Appends chapter number, e.g. 1.1. Section, Theorem 1.1., Table 1.1., etc\\
Default: 1. Section, Theorem 1., Table 1., etc.,
\item[\tt onethmnum] To number all theorem-like objects in a
single sequence, e.g. Theorem~1., Definition 2., Lemma 3., etc.\\
Default: individual numbering on different theorem-like objects,
e.g. Theorem 1., Definition 1., Lemma 1., etc.\\
\verb|ws-rv-thm| package is required.
\end{alphlist}

Apart from the packages mentioned in Sec~\ref{ra_sec1}, the WSPC
class also requires the following in-house packages for customizing
the theorems, citations and references.\\[6pt]

\noindent
{\tablefont
\begin{tabular}{@{}ll@{}}
{\bf Vancouver (numbered)}\\
\quad\verb|\usepackage{ws-rv-van}| & -- Superscript$^1$ (Default style)\\
\quad\verb|\usepackage[square]{ws-rv-van}| & -- Bracketed [1]\\[6pt]
{\bf Harvard (author-date)}\\
\quad\verb|\usepackage{ws-rv-har}| & -- (Author, 1994)
\end{tabular}\\[6pt]}

\noindent{\bf The contributors are advised to consult the managing editor for the chosen option (e.g., \verb|ws-rv-har|, \verb|addchapnum|, \verb|onethmnum|).}

\section{User Defined Macros}
User defined macros should be placed in the preamble of the article, and not at any other place in the document. Such private definitions, i.e. definitions made using the commands \verb|\newcommand, \renewcommand|, \verb|\newenvironment| or \verb|\renewenvironment|, should be used with great care. Sensible, restricted usage of private definitions is encouraged. Large macro packages and Definitions that are not used in the article should be avoided. Do not change existing environments, commands and other standard parts of \LaTeX.

\section{Running Head}\index{running head}

The WSPC prefers ``Author Name(s)'' as even page header and ``Chapter Title'' as odd page header.

\begin{table}[ht]
\tbl{This table shows how the author names should appear in the
running head and TOC depending upon the number of authors
contributing that paper.} {\begin{tabular}{@{}ll@{}} \toprule
No. of Authors & Author Names\\
\colrule
1 & L. Hatcher\\
2 & I. A. Pedrosa \& I. Guedes\\
3 & B. Feng, X. Gong \& X. Wang\\
4 and more & S. R. Choudhury {\it et al.}\\\botrule
\end{tabular}}
\begin{tabnote}
For TOC and Running Heads, the author names should appear in initial
and surname format, e.g. Lee Hatcher should be abbreviated as
L.~Hatcher.
\end{tabnote}\label{ra_tbl1}
\end{table}

Running heads are obtained by the arguments supplied in the square
brackets of \verb|\chapter[#1]{#2}| and \verb|\author[#1]{#2}|
commands, e.g.,
\begin{verbatim}
\chapter[Short Title for Running Head]{Full title}
\author[F. Author and S. Author]
       {First Author and Second Author}
\end{verbatim}

When more than one \verb|\author| command is used, the names are
combined and included as the last \verb|\author|'s argument, e.g.,

\begin{verbatim}
\chapter[Short Title for Running Head]{Full title}
\author{First Author}
\address{First author's address}
\author[F. Author and S. Author]{Second Author}
\address{Second author's address}
\end{verbatim}

\section{Chapters}
Each chapter should normally be in a separate file. The chapter
title is typeset by using the \verb|\chapter[#1]{#2}| command, where
\verb|[#1]| is an optional short title to be used as a running head
if the title is too long and \verb|#2| is the full title of the
chapter. The short, edited version of the title appears in the table
of contents and running head. The chapter title should be typed in a
way such that only the initial character is in upper case and the
rest is in lower case.

\section{Sectional Units}
Sectional units are obtained in the usual way, i.e. with the \LaTeX{}
instructions \verb|\section|, \verb|\subsection|,
\verb|\subsubsection|, and \verb|\paragraph|.

\section{Section}
Text...

\subsection{Subsection}
Text...

\subsubsection{Subsubsection}
Text...

\paragraph{Paragraph}
Text...

\section*{Unnumbered Section}
Unnumbered sections can be obtained by \verb|\section*|.

\section{Lists of Items}\index{lists}
Lists are broadly classified into four major categories that can
randomly be used as desired by the author:
\begin{alphlist}[(d)]
\item Numbered list.
\item Lettered list.
\item Unnumbered list.
\item Bulleted list.
\end{alphlist}

\subsection{Numbered and lettered list}\index{lists!numbered and lettered}

\begin{enumerate}
\item The \verb|\begin{arabiclist}[]| command is used for the arabic
number list (arabic numbers appearing within or without parenthesis), e.g., (1),
(2); 1., 2. etc.\index{lists!numbered and lettered!arabic (1, 2, 3...)}

\smallskip

\item The \verb|\begin{romanlist}[]| command is used for the roman
number list (roman numbers appearing within parenthesis), e.g., (i),
(ii), etc.\index{lists!numbered and lettered!roman (i, ii, iii...)}

\smallskip

\item The \verb|\begin{Romanlist}[]| command is used for the capital roman
\hbox{number list} (capital roman numbers appearing within parenthesis),
e.g., (I), (II), etc.\index{lists!numbered and lettered!Roman (I, II, III...)}

\smallskip

\item The \verb|\begin{alphlist}[]| command is used for the alphabetical
list (alphabetical characters appearing within parenthesis),
e.g., (a), (b), etc.\index{lists!numbered and lettered!alphabetical (a, b, c...)}

\smallskip

\item The \verb|\begin{Alphlist}[]| command is used for the capital
alphabetical list (capital alphabetical characters appearing within
parenthesis), e.g., (A), (B), etc.\index{lists!numbered and lettered!Alphabetical (A, B, C...)}
\end{enumerate}
Note: For all the above mentioned lists, it is obligatory to enter the last entry's number
in the list within the square bracket, to enable unit alignment.

Items numbered with lowercase Roman numerals:
\begin{romanlist}[(iv)]
\item item one
\item item two
    \begin{alphlist}[(d)]
    \item subitem one
    \item lists within lists can be numbered with lowercase alphabets
    \end{alphlist}
\item item three.
\end{romanlist}

\subsection{Bulleted and unnumbered list}\index{lists!bulleted and unnumbered}

\begin{itemlist}
\item The \verb|\begin{itemlist}| command is used for the bulleted list.

\smallskip

\item The \verb|\begin{unnumlist}| command is used for creating the
  unnumbered list with the turnovers hangindent by 1\,pica.
\end{itemlist}

Lists may be laid out with each item marked by a dot:
\begin{itemlist}
\item item one
\item item two
\begin{itemlist}
\item subitem one
\item subitem two
\item subitem three
\item subitem four.
\end{itemlist}
\item item three
\item item four
\item item five.
\end{itemlist}

\section{Theorems and Definitions}\index{theorems}
The WSPC document styles contain a set of pre-defined
environments for theorems, definitions, proofs, remarks etc.

All theorem-like objects use individual numbering scheme by default.
To number them in a single sequence, load the option
\verb|onethmnum| in the preamble., e.g.,
\verb|\usepackage[onethmnum]{ws-rv-thm}|.

\begin{verbatim}
\begin{theorem}
We have $\# H^2 (M \supset N) < \infty$ for an inclusion ...
\label{ra_thm1}
\end{theorem}
\end{verbatim}

\noindent produces

\begin{theorem}
We have $\# H^2 (M \supset N) < \infty$ for an inclusion $M \supset
N$ of factors of finite index.\label{ra_thm1}
\end{theorem}

\begin{verbatim}
\begin{theorem}[Longo, 1998]
For a given $Q$-system...
\[
N = \{x \in N; T x = \gamma (x) T, T x^* = \gamma (x^*) T\}\,,
\]
and $E_\Xi (\cdot) = T^* \gamma (\cdot) T$ gives ...
\label{ra_thm2}
\end{theorem}
\end{verbatim}

\noindent generates

\begin{theorem}[Longo, 1998]
For a given $Q$-system...
\[
N = \{x \in N; T x = \gamma (x) T, T x^* = \gamma (x^*) T\}\,,
\]
and $E_\Xi (\cdot) = T^* \gamma (\cdot) T$ gives a conditional
expectation onto $N$.
\label{ra_thm2}
\end{theorem}

The following environments are available by default with WSPC
document styles:

\begin{center}
{\tablefont
\begin{tabular}{ll}
\toprule Environment & Heading\\\colrule
\verb|algorithm| & Algorithm\\
\verb|answer| & Answer\\
\verb|assertion| & Assertion\\
\verb|assumption| & Assumption\\
\verb|case| & Case\\
\verb|claim| & Claim\\
\verb|comment| & Comment\\
\verb|condition| & Condition\\
\verb|conjecture| & Conjecture\\
\verb|convention| & Convention\\
\verb|corollary| & Corollary\\
\verb|criterion| & Criterion\\
\verb|definition| & Definition\\
\verb|example| & Example\\
\verb|lemma| & Lemma\\
\verb|notation| & Notation\\
\verb|note| & Note\\
\verb|observation| & Observation\\
\verb|problem| & Problem\\
\verb|proposition| & Proposition\\
\verb|question| & Question\\
\verb|remark| & Remark\\
\verb|solution| & Solution\\
\verb|step| & Step\\
\verb|summary| & Summary\\
\verb|theorem| & Theorem\\\botrule
\end{tabular}}\label{ra_theo}
\end{center}

\LaTeX{} provides \verb|\newtheorem| to create new theorem
environments. To add a new theorem-type environments to a chapter, use

\begin{verbatim}
\newtheorem{example}{Example}[section]
\let\Examplefont\upshape
\def\Exampleheadfont{\bfseries}
\end{verbatim}

\subsection{Proofs}
The WSPC document styles also provide a predefined proof environment for proofs.
The proof environment produces the heading
`Proof' with \hbox{appropriate} spacing and punctuation. A `Q.E.D.' symbol, $\square$,
can be appended at the end of a proof with the command \verb|\qed|, e.g.,

\begin{verbatim}
\begin{proof}
This is to test.
\end{proof}
\end{verbatim}

\noindent produces

\begin{proof}
This is to test.
\end{proof}

The proof environment takes an argument in curly
braces, which allows you to substitute a different name for the standard
`Proof'. If you want to display, `Proof of Lemma', then write

\begin{verbatim}
\begin{proof}[Proof of Lemma]
This is to test.
\end{proof}
\end{verbatim}

\noindent produces

\begin{proof}[Proof of Lemma]
This is to test.
\end{proof}

\section{Mathematical Formulas}
\paragraph{Inline:}
For in-line formulas use \verb|\( ... \) or $ ... $|. Avoid
built-up constructions, for example fractions and matrices, in
in-line formulas. Fractions in inline can be typed with a solidus e.g. \verb|x+y/z=0|.
\index{equations!inline}

\paragraph{Display:}
For numbered display formulas use the displaymath environment
\index{equations!display}
\verb|\begin{equation} ...| \verb|\end{equation}|.

And for unnumbered display formula use
\verb|\[ ... \]|. For numbered displayed
one line formulas always use the equation environment. Do not use
\verb|$$ ... $$|. For example, the input for:
\begin{equation}
\mu(n, t) = \frac{\sum\limits^\infty_{i=1}1
(d_i < t, N(d_i) = n)}
{\int\limits^t_{\sigma=0}1(N(\sigma)=n)d\sigma}\,.
\label{ra_eq1}
\end{equation}

\noindent is:

\begin{verbatim}
\begin{equation}
\mu(n, t)=\frac{\sum\limits^\infty_{i=1}1(d_i < t, N(d_i)=n)}
{\int\limits^t_{\sigma=0}1(N(\sigma)=n)d\sigma}\,.\label{ra_eq1}
\end{equation}
\end{verbatim}

For displayed multi-line formulas use the eqnarray environment.

\begin{verbatim}
\begin{eqnarray}
\zeta\mapsto\hat{\zeta}&=&a\zeta+b\eta\label{ra_eq2}\\
\eta\mapsto\hat{\eta}&=&c\zeta+d\eta\label{ra_eq3}
\end{eqnarray}
\end{verbatim}

\noindent\begin{eqnarray}
\zeta\mapsto\hat{\zeta}&=&a\zeta+b\eta\label{ra_eq2}\\
\eta\mapsto\hat{\eta}&=&c\zeta+d\eta\label{ra_eq3}
\end{eqnarray}

Superscripts and subscripts that are words or abbreviations, as in
\( \pi_{\mathrm{low}} \), should be typed as roman letters; this is
done as \verb|\( \pi_{\mathrm{low}} \)| instead of \( \pi_{low} \)
done by \verb|\( \pi_{low} \)|.

For geometric functions, e.g. exp, sin, cos, tan, etc. please use the macros
\verb|\sin, \cos, \tan|. These macros gives proper spacing in mathematical formulas.

It is also possible to use the \AmS-\LaTeX{}
package,\cite{ams04} which can be obtained from the \AmS, from various \TeX{}
archives.

\section{Floats}\index{floats}
\subsection{Tables}\index{floats!tables}
Put the tables and figures in the text with the table and figure environments,
and position them near the first reference of the table or figure in
the text. Please avoid long caption texts in figures and tables.
Do not put them at the end of the article.

\begin{verbatim}
\begin{table}[ht]
\tbl{Sample table caption.}
{\begin{tabular}{@{}cccc@{}} \toprule
Piston mass$^{\text a}$ & Analytical ...\\
& (Rad/s) & (Rad/s) \\ \colrule
1.0... \\
0.001...\\ \botrule
\end{tabular}}
\begin{tabnote}
$^{\text a}$Sample table footnote.
\end{tabnote}
\label{ra_tbl2}
\end{table}
\end{verbatim}

\begin{table}[ht]
\tbl{Sample table caption.}
{\begin{tabular}{@{}cccc@{}} \toprule
Piston mass$^{\text a}$ & Analytical frequency & TRIA6-$S_1$ model & \% Error \\
& (Rad/s) & (Rad/s) \\ \colrule
1.000 & \hphantom{0}281.0 & \hphantom{0}280.81 & 0.07 \\
0.010 & 2441.0 & 2441.00 & 0.00 \\
0.001 & 4130.0 & 4129.30 & 0.16\\ \botrule
\end{tabular}
}
\begin{tabnote}
$^{\text a}$Sample table footnote.
\end{tabnote}
\label{ra_tbl2}
\end{table}

For most tables, the horizontal rules are obtained by:
\begin{description}\index{floats!tables!rules}
\item[toprule] one rule at the top
\item[colrule] one rule separating column heads from data cells
\item[botrule] one bottom rule
\item[Hline] one thick rule at the top and bottom of the tables with multiple column heads
\end{description}

To avoid the rules sticking out at either end
of the table add \verb|@{}| before the first and after the last descriptors, e.g.
{@{}llll@{}}. Please avoid vertical rules in tables. But if you think the vertical rule is must,
you can use the standard \LaTeX{} \verb|tabular| environment.

By using \verb|\tbl| command in table environment, long captions will
be justified to the table width while the short or single line captions
are centered. \verb|\tbl{table caption}{tabular environment}|.
If we need the fixed width for
the tables, the command is \verb|\begin{tabular*}{#1}{@{}ll@{}}|
and \verb|\end{tabular*}|. In the argument \verb|#1| the width of
the table has to be given. For example if we need the table to be
of 25pc width, then the command is
\verb|\begin{tabular*}{25pc}{@{\extracolsep{fill}}ll@{}}|.

Headings which span for more than one column should be set using
\verb|\multicolumn{#1}{#2}{#3}| where \verb|#1| is the number of
columns to be spanned, \verb|#2| is the argument for the alignment
of the column head which may be either {c} --- for center
alignment; {l} --- for left alignment; or {r} --- for right
alignment, as desired by the users. Use {c} for column heads as
this is the WS style and \verb|#3| is the heading. A simplified
alternative version is \verb|\centre{#1}{#2}| where \verb|#1| is
the number of columns to be spanned and \verb|#2| the heading.
There should be a rule spanning the same columns below the
heading. Termed as spanner or bridge rule, it is generated using
the command \verb|\cline{n-m}| where \verb|n| is the number of the
first spanned column and \verb|m| that of the last spanned column.
\verb|\cline| should not be part of a row but follow immediately
after a \verb|\\|.

\def\p{\phantom{$-$}}
\def\pc{\phantom{,}}
\def\p0{\phantom{0}}
\begin{sidewaystable}
\tbl{Positive values of $X_0$ by eliminating $Q_0$ from Eqs.~(15)
and (16) for different values of the parameters $f_0$, $\lambda_0$
and $\alpha_0$ in various dimension.}
{\begin{tabular}{@{}ccccccccccc@{}}
\toprule\\[-6pt]
$f_0$ &$\lambda_0$ &$\alpha_0$
&\multicolumn{8}{c}{Positive roots ($X_0$)}\\[3pt]
\hline\\[-6pt]
&& &4D &5D &6D &7D &8D &10D &12D &16D\\[3.5pt]
\hline\\[-6pt]
\phantom{1}$-0.033$ &0.034 &\phantom{0}0.1\phantom{.01}
&6.75507,\p0 &4.32936,\p0 &3.15991,\p0 &2.44524,\p0
&1.92883,\p0 &0.669541, &--- &---\\[3.5pt]
&&&1.14476\pc\p0 &1.16321\pc\p0 &1.1879\pc\phantom{00}
&1.22434\pc\p0 &1.29065\pc\p0
&0.415056\pc\\[3.5pt]
\phantom{1}$-0.1$\phantom{33} &0.333 &\phantom{0}0.2\phantom{.01}
&3.15662,\p0 &1.72737,\p0 &--- &--- &--- &--- &--- &---\\[3.5pt]
&&&1.24003\pc\p0 &1.48602\pc\p0\\[3.5pt]
\phantom{1}$-0.301$ &0.302 &0.001
&2.07773,\p0 &--- &--- &--- &--- &--- &--- &---\\[3.5pt]
&&&1.65625\pc\p0\\[3.5pt]
\phantom{1}$-0.5$\phantom{01} &0.51\phantom{2} &\phantom{0}0.001
&--- &--- &--- &--- &--- &--- &--- &---\\[3.5pt]
$\phantom{1-}$0.1\phantom{01} &0.1\phantom{02}
&\phantom{0}2\phantom{.001} &1.667,\phantom{000} &1.1946\phantom{00,}
&--- &--- &--- &--- &--- &---\\[3.5pt]
&&&0.806578\pc &0.858211\pc\\[3.5pt]
$\phantom{1-}$0.1\phantom{01} &0.1\phantom{33} &10\phantom{.001}
&0.463679\pc &0.465426\pc &0.466489\pc &0.466499\pc
&0.464947\pc &0.45438\pc\p0 &0.429651\pc &0.35278\pc\\[3.5pt]
$\phantom{1-}$0.1\phantom{01} &1\phantom{.333}
&\phantom{0}0.2\phantom{01}
&--- &--- &--- &--- &--- &--- &--- &---\\[3.5pt]
$\phantom{-0}$1\phantom{.033} &0.001 &\phantom{0}2\phantom{.001}
&0.996033, &0.968869, &0.91379,\p0 &0.848544,&0.783787, &0.669541,
&0.577489, &---\\[3.5pt]
&&&0.414324\pc &0.41436\pc\p0 &0.414412\pc &0.414489\pc &0.414605\pc
&0.415056\pc &0.416214\pc\\[3.5pt]
\phantom{10}\phantom{.033} &0.001 &\phantom{0}0.2\phantom{01}
&0.316014, &0.309739, &--- &--- &--- &--- &--- &---\\[3.5pt]
&&&0.275327\pc &0.275856\pc\\[3.5pt]
\phantom{10}\phantom{.033} &0.1\phantom{33}
&\phantom{0}5\phantom{.001}
&0.089435\pc &0.089441\pc &0.089435\pc &0.089409\pc &0.08935\pc\p0
&0.089061\pc &0.088347\pc &0.084352\pc\\[3.5pt]
\phantom{10}\phantom{.033} &1\phantom{.333} &\phantom{0}3\phantom{.001}
&0.128192\pc &0.128966\pc &0.19718,\p0 &0.169063, &0.142103,
&--- &--- &---\\[3.5pt]
&&&& &0.41436\pc\p0 &0.414412\pc &0.414489\pc\\[3pt]
\Hline
\end{tabular}}\label{ra_tbl3}
\end{sidewaystable}

If a table contains note(s), as a
universal thumb-rule they should appear beneath the table set to its
width and seldom at the foot of the page. For the footnotes in the
table environment the command is
\verb|{\begin{tabnote}<text>\end{tabnote}}|. Appropriate symbols
should be included in the body of the table matching their
corresponding symbols in the footnotes where the footnotes are to be
placed immediately after the \verb|{\begin{tabnote}| command and
terminated before \verb|\end{tabnote}}\end{table}| command.

Landscape tables and figures can be typeset with following environments:
\begin{itemize}\index{floats!tables!landscape}
\item \verb|sidewaystable| and
\item \verb|sidewaysfigure|.
\end{itemize}

\noindent {\bf Example:}

\begin{verbatim}
\begin{sidewaystable}
\tbl{Positive values of ...}
{\begin{tabular}{@{}ccccccccccc@{}}
...
\end{tabular}} \label{ra_tbl3}
\end{sidewaystable}
\end{verbatim}

\subsection{Figures}\index{floats!figures}
The preferred graphics are tiff and Encapsulated
PostScript, eps in short, for any type of graphic. Our
\TeX\ installation requires eps, but we can easily convert tiff to eps.
Many other formats, e.g. pict (Macintosh), wmf (Windows) and various proprietary
formats, are not suitable. Even if we can read such files, there is no guarantee
that they will look the same on our systems as on yours.

Next adjust the scaling of the figure until it's correctly positioned,
and remove the declarations of the lines and any anomalous spacing.
If instead you wish to use some other method, then it's most
important to leave the right amount of vertical space in the figure
declaration to accommodate your figure. A figure is obtained with the following commands

\begin{verbatim}
\begin{figure}
\centerline{\includegraphics[width=3.8cm]{rv-fig1}}
\caption{Sample figure caption.} \label{ra_fig1}
\end{figure}
\end{verbatim}

\begin{figure}
\centerline{\includegraphics[width=3.8cm]{rv-fig1}}
\caption{Sample figure caption.}
\label{ra_fig1}
\end{figure}

Sub-figures are obtained with the following commands
\begin{verbatim}
\begin{figure}[ht]
\centerline{
  \subfigure[]
     {\includegraphics[width=2in]{rv-fig2a}\label{ra_fig2a}}
  \hspace*{4pt}
  \subfigure[Optional subcaption]
     {\includegraphics[width=2in]{rv-fig2b}\label{ra_fig2b}}
}
\caption{Common caption here.} \label{ra_fig2} % common label
\end{figure}
\end{verbatim}

\begin{figure}[ht]
\centerline{
  \subfigure[]
     {\includegraphics[width=2in]{rv-fig2a}\label{ra_fig2a}}
  \hspace*{4pt}
  \subfigure[Optional subcaption.]
     {\includegraphics[width=2in]{rv-fig2b}\label{ra_fig2b}}
}
\caption{Common caption here.}\label{ra_fig2}
\end{figure}

Sub-figures \fref{ra_fig2a} and Fig.~\ref{ra_fig2b} are
referred with \verb|\fref{ra_fig2a}| and \verb|Fig.~\ref{ra_fig2b}| commands.

\begin{sidewaysfigure}
\centerline{\includegraphics[width=6.6in]{rv-fig3}}
\caption{The bifurcating response curves of system
$\alpha=0.5$, $\beta=1.8$; $\delta=0.2$, $\gamma=0$:
(a) $\mu=-1.3$; and (b) $\mu=0.3$.}
\label{ra_fig3}
\end{sidewaysfigure}

Large figures and tables should be placed on a page by
themselves, e.g.,

\index{floats!figures!landscape}
\begin{verbatim}
\begin{sidewaysfigure}
\centerline{\includegraphics[width=6.6in]{rv-fig3}}
\caption{Sample figure caption.} \label{ra_fig3}
\end{sidewaysfigure}
\end{verbatim}

\begin{figure}[ht]
\centerline{
  \minifigure[Sample caption.]
     {\includegraphics[width=2in]{rv-fig2a}\label{ra_fig4}}
  \hspace*{4pt}
  \minifigure[Sample caption.]
     {\includegraphics[width=2in]{rv-fig2b}\label{ra_fig5}}
}
\end{figure}

Side-by-side figures Fig.~\ref{ra_fig4} and \fref{ra_fig5} are
obtained with \verb|\minifigure| commands.

\begin{verbatim}
\begin{figure}[ht]
\centerline{
  \minifigure[Sample caption.]
     {\includegraphics[width=2in]{rv-fig2a}\label{ra_fig4}}
  \hspace*{4pt}
  \minifigure[Sample caption.]
     {\includegraphics[width=2in]{rv-fig2b}\label{ra_fig5}}
}
\end{figure}
\end{verbatim}

\section{Cross-references}
Use \verb|\label| and \verb|\ref| for cross-references to
equations, figures, tables, sections, etc., instead
of plain numbers. Every numbered part to which one wants to refer,
should be labelled with the instruction \verb|\label|, for e.g.,

\begin{verbatim}
\begin{equation}
\mu(n, t)=\frac{\sum\limits^\infty_{i=1}1(d_i < t, N(d_i)=n)}
{\int\limits^t_{\sigma=0}1(N(\sigma)=n)d\sigma}. \label{ra_eq1}
\end{equation}
\end{verbatim}

With the instruction \verb|\ref| one can refer to a numbered part
that has been labelled, e.g., \verb|..., see also Eq. (\ref{ra_eq1})|.

\begin{center}{\tablefont
Some useful shortcut commands.
\begin{tabular}{lll}
\toprule
Shortcut & Equivalent & Output \\
command & \TeX\ command\\\colrule
\multicolumn{3}{@{}l}{In the middle of a sentence:}\\
\verb|\eref{ra_eq1}|  & Eq.~(\verb|\ref{ra_eq1}|) & \eref{ra_eq1}\\
\verb|\sref{ra_sec1}| & Sec.~\verb|\ref{ra_sec1}| & \sref{ra_sec1}\\
\verb|\cref{ra_ch1}|  & Chap.~\verb|\ref{ra_ch1}| & \cref{ra_ch1}\\
\verb|\fref{ra_fig1}| & Fig.~\verb|\ref{ra_fig1}| & \fref{ra_fig1}\\
\verb|\tref{ra_tbl1}| & Table~\verb|\ref{ra_tbl1}| & \tref{ra_tbl1}\\[3pt]
\multicolumn{2}{@{}l}{At the starting of a sentence:}\\
\verb|\Eref{ra_eq1}|  & Equation (\verb|\ref{ra_eq1}|) & \Eref{ra_eq1}\\
\verb|\Sref{ra_sec1}| & Section~\verb|\ref{ra_sec1}| & \Sref{ra_sec1}\\
\verb|\Cref{ra_ch1}|  & Chapter~\verb|\ref{ra_ch1}| & \Cref{ra_ch1}\\
\verb|\Fref{ra_fig1}| & Figure~\verb|\ref{ra_fig1}| & \Fref{ra_fig1}\\
\verb|\Tref{ra_tbl1}| & Table~\verb|\ref{ra_tbl1}| & \Tref{ra_tbl1}\\\botrule
\end{tabular}}
\end{center}

\begin{itemize}
\item The \verb|\label| instruction should be typed
immediately after (or one line below), e.g.,
\verb|\caption{ ... caption ... }\label{ra_fig2}|. Labels should not be typed inside the argument of
a number-generating instruction such as \verb|\section| or \verb|\caption|.
\item labels should not be repeated.
\end{itemize}

\section{Citations}\label{ra_cit}\index{citation}

World Scientific's preferred style for Review Volume is the Vancouver (numbered) system,
unless if the text is not very heavily referenced in which case the
Harvard (author-date) system may be used.

\begin{center}
\tablefont
\begin{tabular}{@{}ll@{}}\toprule
System & Package\\\colrule
 Vancouver (numbered)\\
 \quad$\bullet$ Bracketed [1] & \verb|\usepackage[square]{ws-rv-van}|\\
 \quad$\bullet$ Superscript$^1$ & \verb|\usepackage{ws-rv-van}| (Default style)\\[3pt]
 Harvard (author-date) & \verb|\usepackage{ws-rv-har}|\\\botrule
\end{tabular}
\end{center}

Citations in the text use the labels defined in the bibitem declaration,
for example, the first paper by Jarlskog\cite{jarl88} is cited using the
command \verb|\cite{jarl88}|. The bibitem labels should not be repeated.

For multiple citations do not use \verb|\cite{1}\cite{2}|, but use
\verb|\cite{1,2}| instead.

When superscripted citations are used, there should not be a space
before \verb|\cite{key}|, e.g.,

citation:
\verb|see\cite{zipf}|\hskip-60pt\lower8pt\hbox{$\uparrow$}\hskip-4pt\lower16pt\hbox{no
character space here}

\subsection{Vancouver Style}\index{citation!numbered}

Reference citations in the text are to be numbered consecutively in
Arabic numerals, in the order of first appearance. The numbered citations
can appear in two ways:

\begin{romanlist}[(ii)]
\item bracketed
\item superscript (default style)
\end{romanlist}

\subsubsection{Bracketed}\index{citation!numbered!bracketed}
References cited in the text are within square brackets, e.g.,

\begin{arabiclist}[(2)]
\item \verb|``One can deduce from Ref.~\cite{benh93} that...''|\\
            ``One can deduce from Ref.~[3] that...''
\smallskip
\item \verb|``See Refs.~\cite{ams04,bake72,benh93,brow88} and|\\
      \verb|\cite{davi93} for more details.''|\\
      ``See Refs.~[1--3, 5] and [7] for more details.''
\end{arabiclist}

\subsubsection{Superscript}\index{citation!numbered!superscript}

References cited in the text appear as superscripts, e.g.,

\begin{arabiclist}[(2)]
\item \verb|``...in the statement.\cite{ams04}''|\\
            ``...in the statement.$^1$''
\smallskip
\item \verb|``...have proven\cite{bake72} that this equation...''|\\
            ``...have proven$^2$ that this equation...''
\end{arabiclist}

When the reference forms part of the sentence, it should appear with
``Reference'' or ``Ref.'', e.g.,

\begin{arabiclist}[(2)]
\item \verb|``One can deduce from Ref.~\refcite{benh93} that...''|\\
      ``One can deduce from Ref.~3 that...''
\smallskip
\item \verb|``See Refs.~\citen{ams04,bake72,benh93,brow88}|
      \verb|\citen{davi93} for more details.''|\\
     ``See Refs.~1--3, 5 and 7 for more details.''
\end{arabiclist}

\subsection{Harvard Style}\index{citation!author-date}
Citations in the text use the labels defined in the
\verb|bibitem| declaration, for example, [Jarlskog (1988)]
is cited using the command \verb|\cite{jarl88}|.
While \verb|\citet {jarl88}| produces Jarlskog (1988).
See Sec.~\ref{ra_secbib} for more details on coding references in Vancouver and Harvard styles.

\section{Footnote}
Footnotes are denoted by a Roman letter superscript in the text. Footnotes can be used as

\verb|... total.\footnote{Sample footnote text.}|

\noindent {\bf Output:}

\noindent ... in total.\footnote{Sample footnote text.}

\section{Miscellaneous}
\subsection{Quote}\index{quote}
Here is an example for the \verb|quote| environment.

\begin{quote}
This is an example for the \verb|quote| environment. Quote text is
indented by 1pc on the left and right sides. The point size for the
quote text is 9/11pt.
\end{quote}

\subsection{Boxed text}\index{boxed text}
Here is an example for the \verb|boxedtxt| environment.

\begin{boxedtxt}
This is an example for the \verb|boxedtxt| environment. The text will be
placed inside a box with 6pt space on all sides. The box rule
thickness is 0.5pt.
\end{boxedtxt}

\section{Acknowledgments}
Acknowledgments to funding bodies etc. may be placed in a separate
section at the end of the text, before the Appendices. This should not
be numbered so use \verb|\section*{Acknowledgements}|.

\section{Appendix}\index{appendix}
Appendices should be used only when absolutely necessary. They
should come before the References.

\begin{verbatim}
\begin{appendix}[Optional Appendix Title]
\section{Sample Appendix}
Text...
\begin{equation}
\mu(n, t) = ...\label{ra_appen1}
\end{equation}
\subsection{Sample Subsection}
Text...
\begin{equation}
\zeta\mapsto...\label{ra_appen2}
\end{equation}
\end{appendix}
\end{verbatim}

\section{Bibliography}\label{ra_secbib}\index{bibliography}
Use \verb|\bibitem| to produce the bibliography.
The bibitem labels should not be repeated.

\subsection{\btex\ users}

\btex\index{BIBTeX} users should use our bibliography style file
\verb|ws-rv-van.bst| or \verb|ws-rv-har.bst|.

If you use the \btex\ program to maintain your bibliography, you
don't use the \verb|thebibliography| environment. Instead, you
include the lines

\begin{verbatim}
   \usepackage{ws-rv-van}
   ...
   \bibliographystyle{ws-rv-van}
   \bibliography{bibfile}
\end{verbatim}

\noindent where \verb|ws-rv-van| refers to a file \verb|ws-rv-van.bst|, which
defines how your references will look.
The argument to \verb|\bibliography| refers to the file
\verb|bibfile.bib|, which should contain your database in \btex\
format. Only the entries referred to via \verb|\cite| will be listed
in the bibliography.

To complete the job, compile your file as follows:

\begin{enumerate}[(7)]
\item latex ws-rv9x6
\item latex ws-rv9x6
\item bibtex ws-rv9x6
\item latex ws-rv9x6
\item latex ws-rv9x6
\end{enumerate}

\begin{center}
\tablefont
\begin{tabular}{@{}ll@{}}\toprule
\multicolumn{1}{c}{\btex}\\
\multicolumn{1}{c}{Database}  & \multicolumn{1}{c}{Sample citation}\\
\multicolumn{1}{c}{entry type}\\\colrule

article & ... text.\citep{best03,pier02}\\

proceedings & ... text.\cite{weis94}\\

inproceedings & ... text.\cite{gupt97}\\

book & ... text.\cite{rich60,jarl88}\\

edition & ... text.\cite{chur90}\\

editor & ... text.\cite{benh93}\\

series & ... text.\cite{bake72}\\

tech report & ... text.\cite{hobb92,bria84}\\

unpublished & ... text.\citet{hear94}\\

phd thesis & ... text.\cite{brow88}\\

masters thesis & ... text.\cite{lodh74}\\

incollection & ... text.\cite{dani73}\\

misc & ... Ref. \citen{davi93} ...\\
\botrule
\end{tabular}
\end{center}

To use Harvard (author-date) system \verb|ws-rv-har.bst| is used
with \verb|\usepackage{ws-rv-har}|.

\subsection{Non-\btex\ users}
For Vancouver (numbered) style users, references are to be listed in the order cited in the text.

Use the style shown in the following examples.

\begin{verbatim}
\begin{thebibliography}{9}
\bibitem{ams04}
   \AmS, \emph{\AmS-\LaTeX{} Version 2 User's Guide},
   American Mathematical Society, Providence (2004).
   \url{http://www.ams.org/tex/amslatex.html}.
\bibitem{jarl88}
   C.~Jarlskog, \emph{CP Violation}. World Scientific,
   Singapore (1988).

\bibitem{best03}
   B.~W. Bestbury, {$R$}-matrices and the magic square,
   \emph{J. Phys. A}. {\bf 36}(7), 1947--1959 (2003).

\bibitem{pier02}
   P.~X. Deligne and B.~H. Gross, On the exceptional series,
   and its descendants, \emph{C. R. Math. Acad. Sci. Paris}.
   {\bf 335}(11), 877--881 (2002).
\end{thebibliography}
\end{verbatim}

For Harvard (author-date) style users, the references are to be listed in
alphabetical order according to the surname of the first author.

Use the style shown in the following examples.

\begin{verbatim}
\begin{thebibliography}{9}
\bibitem[{Baker and Carter(1972)}]{bake72}
   Baker, D.~W. and Carter, N.~L. (1972). \emph{Seismic
   Velocity Anisotropy Calculated for Ultramafic Minerals
   and Aggregates}, \emph{Geophys. Mono.}, Vol.~16,
   Am. Geophys. Union, pp. 157--166.

\bibitem[{Benhamou and Colmerauer(1993)}]{benh93}
   Benhamou, F. and Colmerauer, A. eds. (1993).
   \emph{Constraint Logic Programming, Selected Research},
   MIT Press.

\bibitem[{Bestbury(2003)}]{best03}
   Bestbury, B.~W. (2003). {$R$}-matrices and the magic
   square, \emph{J. Phys. A} \textbf{36}, 7, pp. 1947--1959.

\bibitem[{Brown(1988)}]{brow88}
   Brown, M.~E. (1988). \emph{An Interactive Environment for
   Literate Programming}, Ph.D. thesis, Texas A\&M University,
   TX, USA.

\bibitem[{Churchill and Brown(1990)}]{chur90}
   Churchill, R.~V. and Brown, J.~W. (1990). \emph{Complex
   Variables and Applications}, 5th edn., McGraw-Hill.
\end{thebibliography}
\end{verbatim}

\section{Single Indexing}\index{index}
The first step in producing the index is to put the necessary
\verb|\index| commands in your document. The following example shows
some simple \verb|\index| commands and the index entries that they
produce.

\begin{table}[ht]
\tablefont
\begin{tabular}{ll|l}
Page ii:   & \verb|\index{Alpha}              | &  Alpha, ii           \\
Page viii: & \verb|\index{alpha}              | &  alpha, viii, ix, 22 \\
Page ix:   & \verb|\index{alpha}              | &  bites               \\
Page 7:    & \verb|\index{gnat!size of}       | &  \sitem animal       \\
Page 8:    & \verb|\index{bites!animal!gnats} | &  \ssitem gnats, 8, 10\\
Page 10:   & \verb|\index{bites!animal!gnats} | &  \ssitem gnus, 10    \\
Page 10:   & \verb|\index{bites!animal!gnus}  | &  \sitem vegetable, 12\\
Page 12:   & \verb|\index{bites!vegetable}    | &  gnat, 32            \\
Page 22:   & \verb|\index{alpha}              | &  \sitem anatomy, 35  \\
Page 32:   & \verb|\index{gnat}               | &  \sitem size of, 7   \\
Page 35:   & \verb|\index{gnat!anatomy}       | &  gnus                \\
           & \verb|\index{gnus!good}          | &  \sitem bad, 38      \\
Page 38:   & \verb|\index{gnus!bad}           | &  \sitem good, 35
\end{tabular}
\end{table}

You then run \LaTeX\ on your entire document, causing it to generate
the file {\tt ws-rv9x6.idx}. Next, run the \verb|MakeIndex|
program by typing the command, \verb|makeindex ws-rv9x6|.
This produces the file \verb|ws-rv9x6.ind|. If \verb|MakeIndex|
generated no error messages, you can now rerun \LaTeX\ on your
document and the index will appear.
Compile your file as follows:

\begin{enumerate}[(4)]
\item latex ws-rv9x6
\item latex ws-rv9x6
\item bibtex ws-rv9x6\verb|            % when bibtex is used|
\item makeindex ws-rv9x6
\item latex ws-rv9x6
\item latex ws-rv9x6
\end{enumerate}

Reading the index, you may discover mistakes, which
should be corrected by changing the appropriate \verb|\index|
commands in the document and regenerating the {\tt ind} file. If
there are problems that cannot be corrected in this way, you can
edit the {\tt ind} file directly. However, such editing is
to be avoided because it must be repeated every time you generate a
new version of the index.

If you are making changes in the .toc or .ind files directly,
then include \verb"\nofiles" before
\verb"\begin{document}" to avoid overwriting. However, the command
\verb"\nofiles" should be used as the last option.

\section{Multiple Indexes}\label{ra_sec3.1}\index{index!multiple indexes}

To create a ``subject'' and an ``author'' index, the following
packages and declarations should be included in the \TeX{}
file:

\begin{verbatim}
...
\usepackage{ws-index}
\makeindex
\newindex{aindx}{adx}{and}{Author Index}       % author index
\renewindex{default}{idx}{ind}{Subject Index}  % subject index
...
\printindex[aindx]                  % to print author index
\printindex                         % to print subject index
\end{verbatim}

In text, the subject or default index entries are tagged with
\verb|\index{entry}|, and the author index entries are marked
with \verb|\index[aindx]{entry}| or \verb|\aindx{entry}|.

\begin{verbatim}
\index[aindx]{Author, F.}  % or \aindx{Author, F.}
\index{FAQ}
\end{verbatim}

To complete the job, compile your file as follows:

\begin{enumerate}[(7)]
\item latex ws-rv9x6
\item latex ws-rv9x6
\item bibtex ws-rv9x6
\item makeindex ws-rv9x6
\item makeindex -o ws-rv9x6.and ws-rv9x6.adx
\item latex ws-rv9x6
\item latex ws-rv9x6
\end{enumerate}

\begin{appendix}[Appendix Title]
Appendices should be used only when absolutely necessary. They
should come before the References.

\begin{table}[ht]
\tbl{Commonly used macros.}{
\begin{tabular}{@{}ll@{}}
\toprule
Macro name&Purpose\\
\colrule
{\tt$\backslash$chapter[\#1]}\{{\tt\#2}\} & Chapter title\\
{\tt$\backslash$author[\#1]}\{{\tt\#2}\} & Author Name(S)\\
{\tt$\backslash$address}\{{\tt\#1}\} & Address\\
{\tt$\backslash$begin}\{{\tt{abstract}}\} & Start Abstract\\
{\tt$\backslash$end}\{{\tt{abstract}}\} & End Abstract\\
{\tt$\backslash$bigtoc} & For longer TOCs (e.g. 1.99. Section Title)\\
{\tt$\backslash$smalltoc} & For smaller TOCs (e.g. 1.9. Section Title)\\
{\tt$\backslash$tableofcontents} & Table of Contents\\
{\tt$\backslash$body} & Start Body Text\\
{\tt$\backslash$section}\{{\tt\#1}\} & Section heading\\
{\tt$\backslash$subsection}\{{\tt\#1}\} & Subsection heading\\
{\tt$\backslash$subsubsection}\{{\tt\#1}\} & Subsubsection heading\\
{\tt$\backslash$section*}\{{\tt\#1}\} & Unnumbered Section head\\
{\tt$\backslash$begin}\{{\tt{itemlist}}\} & Start bulleted lists\\
{\tt$\backslash$end}\{{\tt{itemlist}}\} & End bulleted lists\\
{\tt$\backslash$begin}\{{\tt{arabiclist}}\} & Start arabic lists (1, 2, 3...)\\
{\tt$\backslash$end}\{{\tt{arabiclist}}\} & End arabic lists\\
{\tt$\backslash$begin}\{{\tt{romanlist}}\} & Start roman lists (i, ii, iii...)\\
{\tt$\backslash$end}\{{\tt{romanlist}}\} & End roman lists\\
{\tt$\backslash$begin}\{{\tt{Romanlist}}\} & Start roman lists (I, II, III...)\\
{\tt$\backslash$end}\{{\tt{Romanlist}}\} & End roman lists\\
{\tt$\backslash$begin}\{{\tt{alphlist}}\} & Start alpha lists (a, b, c...)\\
{\tt$\backslash$end}\{{\tt{alphlist}}\} & End alpha lists\\
{\tt$\backslash$begin}\{{\tt{Alphlist}}\} & Start alpha lists (A, B, C...)\\
{\tt$\backslash$end}\{{\tt{Alphlist}}\} & End alpha lists\\
{\tt$\backslash$begin}\{{\tt{proof}}\} & Start of Proof\\
{\tt$\backslash$end}\{{\tt{proof}}\} & End of Proof\\
{\tt$\backslash$begin}\{{\tt{theorem}}\} & Start of Theorem\\
{\tt$\backslash$end}\{{\tt{theorem}}\} & End of Theorem (see Page \pageref{ra_theo} for detailed list)\\
{\tt$\backslash$begin}\{{\tt{appendix}}\}[{\tt\#1}] & Start Appendix\\
{\tt$\backslash$end}\{{\tt{appendix}}\} & End Appendix\\
{\tt$\backslash$begin}\{{\tt{thebibliography}}\}\{{\tt\#1}\} & Start of reference list\\
{\tt$\backslash$end}\{{\tt{thebibliography}}\} & End of reference list\\
{\tt$\backslash$bibitem}[\{{\tt\#1}\}]\{{\tt\#2}\}& reference item in author-date style\\
{\tt$\backslash$bibitem}\{{\tt\#1}\} & reference item in numbered style\\
{\tt$\backslash$bibliographystyle}\{{\tt\#1}\} & To include \btex{} style file\\
{\tt$\backslash$bibliography}\{{\tt\#1}\} & To include \btex{} database\\ \botrule
\end{tabular}}
\end{table}

\section{Appendices}\index{appendix}
If there is more than one appendix, number them alphabetically.
Sectional units are obtained with the
\LaTeX{} instructions \verb|\section|, \verb|\subsection|.
Number displayed equations occurring in the Appendix in this way, e.g.~(\ref{ra_appeq1}), (\ref{ra_appeq2}), etc.

\noindent\begin{eqnarray}
\zeta\mapsto\hat{\zeta}&=&a\zeta+b\eta\label{ra_appeq1}\\
\eta\mapsto\hat{\eta}&=&c\zeta+d\eta\label{ra_appeq2}
\end{eqnarray}

\begin{table}[ht]
\tbl{Macros available for Tables/Figures.}{
\begin{tabular}{@{}ll@{}}
\toprule
Environment name&Purpose\\
\colrule
{\tt figure} & figures\\
{\tt sidewaysfigure} & landscape figures\\
{\tt table} & tables\\
{\tt sidewaystable} & landscape tables\\
{\tt$\backslash$tbl}\{{\tt\#1}\}\{{\tt\#2}\} & \#1 - table caption;\\
& \#2 - tabular environment\\[3pt]
\multicolumn{2}{@{}l}{Horizantal Rules for tables}\\
{\tt$\backslash$toprule} & one rule at the top\\
{\tt$\backslash$colrule} & one rule separating column heads from\\ & data cells\\
{\tt$\backslash$botrule} & one bottom rule\\
{\tt$\backslash$Hline} & one thick rule at the top and bottom of\\ & the tables with multiple column heads\\
\botrule
\end{tabular}}
\end{table}
\end{appendix}

\bibliographystyle{ws-rv-van}
\bibliography{ws-rv-sample}

%\blankpage
%\printindex[aindx]                 % to print author index
%\printindex                         % to print subject index

\end{document}

### 2017-05-19 Paper Drafts/submissionDraft/submissionDraft.tex
Submitted version

@@
 
-\author[J.P. Meagher et al.]{J.P. Meagher, T. Damoulas, K.E. Jones, and M. Girolami\footnote{Author footnote.}}
+\author[J.P.
+Meagher et al.]{J.P.
+Meagher\footnote{Department of Statistics, The University of Warwick}, T.
+Damoulas\footnote{The University of Warwick and The Alan Turing Institute}, K.E.
+Jones\footnote{Centre for Biodiversity and Environment Research, University College London}, and M.
+Girolami\footnote{Imperial College London and The Alan Turing Institute}\\}
 %\index[aindx]{Author, F.} % or \aindx{Author, F.}
@@
 
-\address{Department of Statistics,\\
-University of Warwick, \\
-J.Meagher@Warwick.ac.uk\footnote{Affiliation footnote.}}
+\address{
+J.Meagher@Warwick.ac.uk}
 
@@
 A phylogenetic signal was found in the data and ancestral reconstruction performed.
-This promising preliminary analysis paves the way for more realistic models of the evolution of echolocation in bats.
+This promising preliminary analysis paves the way for more realistic models for the evolution of echolocation in bats.
 \end{abstract}
@@
 \cite{group2012phylogenetic} Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits\cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis et al.\cite{hadjipantelis2013function}.
-This current research investigates these techniques in the context of the evolution of bat echolocation calls.
+This current research investigates these techniques in the context of the bat echolocation calls.
 
@@
 \label{sec:pgp}
-Gaussian Processes (GP) provide a concise way of fitting a prior belief over the distribution of functions.
-The structure of this prior belief is described by the GP model kernel, defining it's covariance function.
-GPs are useful for classification and regression problems, and allow posterior predictive distributions to be obtained.
-An in depth treatment of GPs is given by Rasmussen \& Williams.\cite{rasmussen2006gaussian}
+A Gaussian process places a prior distribution over functions, \(f(x) \sim \mathcal{GP}(m(x), k (x, x'),\) where \(x \in \mathbf{R}^P\) is some input variable, the mean function \(m(x) = \mathbf{E}[f(x)]\), and the covariance function \(k(x, x') = \mathnormal{cov}(f(x), f(x') )\).
+Given observations \(\mathbf{y}\) at locations \(\{x_n\}_{n=1}^N\), Gaussian noise, and kernel hyperparameters \(\theta\), a posterior predictive distribution over functions can be inferred analytically.
+See Rasmussen \& Williams\cite{rasmussen2006gaussian} for an in depth treatment.
 
@@
 
-The post processed echolocation call data accompanying Stathopoulos et al.
+The post processed echolocation call data for bats found across north and central Mexico which accompanied Stathopoulos et al.
 \cite{stathopoulos2017bat} was used in this analysis.
-Echolocation calls were recorded across north and central Mexico with a Pettersson 1000x bat detector (Pettersson Elektronik AB, Uppsala, Sweden).
-Live trapped bats were measured and identified to species level using field keys.\cite{ceballos2005mamiferos} \cite{medellin2sanchez} Bats were recorded either while released from the hand or while tied to a zip line.
-The bat detector was set to record calls manually in real time, full spectrum at 500 kHz.
+Live bats were caught, identified, and recorded either while released from the hand or while tied to a zip line.
+Calls were recorded at a sampling frequency of 500 kHz.
 In total the dataset consists of 22 species in five families, 449 individual bats and 1816 individual echolocation call recordings.
@@
 Echolocation calls have complex spectral and temporal structures, much of which is lost in the log energy spectral density representation.
-An alternative representation, which preserves more of this structure, is the spectrogram.
-Implementing this model using the spectrogram as a functional representation of the echolocation call is a priority, as the results of such an analysis are expected to be more meaningful.
+An alternative time-frequency representation, which preserves more of this structure, is the spectrogram.
+Modelling the evolution of bat echolocation calls with spectrograms is an active area of future research.
 

### 2017-06-16 Paper Drafts/Final/finalDraft.tex
Final Draft

--- 2017-05-19 Paper Drafts/submissionDraft/submissionDraft.tex
@@
 \usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
-\usepackage{ws-rv-van}   % numbered citation & references (default)
+\usepackage[square]{ws-rv-van}   % numbered citation & references (default)
 %\usepackage{ws-index}   % to produce multiple indexes
@@
 Meagher et al.]{J.P.
-Meagher\footnote{Department of Statistics, The University of Warwick}, T.
+Meagher\footnote{J.P Meagher would like to thank the EPSRC for funding this work, and also acknowledge the support of the Alan Turing Institute and Gaelic Players Association.}, T.
 Damoulas\footnote{The University of Warwick and The Alan Turing Institute}, K.E.
@@
 
-\address{
+\address{Department of Statistics,\\ The University of Warwick.\\
 	J.Meagher@Warwick.ac.uk}
@@
 The emerging field of Data Science is driven by research which lies at the nexus of Statistics and Computer Science.
-Bioacoustics is one such area generating vast quantities of data, often through citizen science initiatives.
-\cite{allen2006citizen} \cite{pettorelli2013indicator} Bioacoustic techniques for biodiversity monitoring \cite{stathopoulos2017bat} \cite{damoulas2010bayesian} have the potential to make real policy impacts, particularly with regard to sustainable economic development and nature conservation.
+Bioacoustics is one such area generating vast quantities of data, often through citizen science initiatives \cite{pettorelli2013indicator}.
+Bioacoustic techniques for biodiversity monitoring \cite{stathopoulos2017bat} \cite{damoulas2010bayesian} have the potential to make real policy impacts, particularly with regard to sustainable economic development and nature conservation.
 
-Bats (order \textit{Chiroptera}) have been identified as ideal bioindicators for monitoring climate change and habitat quality,\cite{jones2009carpe} and are of particular interest for monitoring biodiversity acoustically.
-Typically, a bat broadcasts information about itself in an ultrasonic echolocation call.\cite{griffin1944echolocation} The development of automatic acoustic monitoring algorithms \cite{stathopoulos2017bat} \cite{walters2012continental} means that large scale, non-invasive monitoring of bats is becoming possible.
+Bats (order \textit{Chiroptera}) have been identified as ideal bioindicators for monitoring climate change and habitat quality \cite{jones2009carpe}, and are of particular interest for monitoring biodiversity acoustically.
+Typically, a bat broadcasts information about itself in an ultrasonic echolocation call \cite{griffin1944echolocation}.
+The development of automatic acoustic monitoring algorithms \cite{stathopoulos2017bat} \cite{walters2012continental} means that large scale, non-invasive monitoring of bats is becoming possible.
 
 Monitoring bat populations provides useful information, but an understanding the orders evolutionary history is required to identify the cause and effect of any changes observed.
-The echolocation call structure, which reflects a bats diet and habitat,\cite{aldridge1987morphology}  is a key aspect of this natural history.
-Reconstructing ancestral traits\cite{joy2016ancestral} relies on a statistical comparative analysis incorporating extant species and fossil records.
-\cite{felsenstein2004inferring} However, the fossil record is of limited use in inferring ancestral echolocation calls in bats.
+The echolocation call structure, which reflects a bats diet and habitat \cite{aldridge1987morphology}, is a key aspect of this evolutionary history.
+Reconstructing ancestral traits \cite{joy2016ancestral} relies on a statistical comparative analysis incorporating extant species and fossil records \cite{felsenstein2004inferring}.
+However, the fossil record is of limited use in inferring ancestral echolocation calls in bats.
 Therefore, statistical data science techniques may shed some light on this topic.
@@
 These call features relied upon domain knowledge to ensure they were sensibly selected and applied.
-More recently, general techniques for the classification of acoustic signals have been developed.
-\cite{stathopoulos2014bat} \cite{damoulas2010bayesian} General techniques for the ancestral reconstruction of function-valued traits have also been proposed.
-\cite{group2012phylogenetic} Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits\cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis et al.\cite{hadjipantelis2013function}.
+More recently, general techniques for the classification of acoustic signals have been developed \cite{stathopoulos2014bat} \cite{damoulas2010bayesian}.
+General techniques for the ancestral reconstruction of function-valued traits have also been proposed \cite{group2012phylogenetic}.
+Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits \cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis et al \cite{hadjipantelis2013function}.
 This current research investigates these techniques in the context of the bat echolocation calls.
 
-Section \ref{sec:funrep} presents details on representing echolocation call recordings as function-valued traits, and Section \ref{sec:pgp} then illustrates the model for evolution developed by a phylogenetic Gaussian Process.
-The results of our analysis of bat echolocation calls are then presented and discussed.
+The structure of this paper is as follows, section \ref{sec:funrep} presents details on representing echolocation call recordings as function-valued traits.
+Section \ref{sec:pgp} develops the model for evolution given by a phylogenetic Gaussian Process.
+The results of the analysis of bat echolocation calls are then presented and discussed in sections \ref{sec:results} and \ref{sec:conc}.
+
 \section{Echolocation Calls as Function-Valued Traits}
 \label{sec:funrep}
-A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time. \cite{ramsay2006functional} 
+A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time \cite{ramsay2006functional}. 
 These measurements can be thought of as representing points on a curve that varies gradually and continuously.
-In the context of phylogenetics, these functional data objects are function-valued traits.
-\cite{meyer2005up}
+In the context of phylogenetics, these functional data objects are function-valued traits \cite{meyer2005up}.
 
@@
 Further, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual.
-It is the echolocation call functions at the species level that we re interested in.
+It is the phylogenetic relationship between the species level echolocation call functions that we are interested in modelling.
 
 The call recordings themselves are functional data objects, however modelling the phylogenetic relationships between \(\{x_{lm}^{\mathbf{t}}(t)\}\) and \(\{x_{l'm'}^{{\mathbf{t}}'}(t)\}\) directly implies that the processes are comparable at time \(t\).
-This is not the case for acoustic signals, a phenomenon which is often addressed by dynamic time warping.\cite{berndt1994using} Another approach to this issue is to consider an alternative functional representation of the signal.
+This is not the case for acoustic signals, a phenomenon which is often addressed by dynamic time warping \cite{berndt1994using}.
+Another approach to this issue is to consider an alternative functional representation of the signal.
 
@@
 \]
-The energy spectral density of \(x_{lm}^{\mathbf{t}}(\cdot)\) is the magnitude of the Fourier transform and the log energy spectral density is given by
+The energy spectral density of \(x_{lm}^{\mathbf{t}}(\cdot)\) is the squared magnitude of the Fourier transform and the log energy spectral density is given by
 \[
@@
 
-The discrete Fourier Transform\cite{antoniou2006digital}
+The discrete Fourier Transform \cite{antoniou2006digital}
 of \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^{\mathbf{t}}(k) : k = 0, \dots, \frac{ N_{lm}^{\mathbf{t}}}{2} + 1\}\).
-Smoothing splines\cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\).
+Smoothing splines \cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\).
 
@@
 \label{sec:pgp}
-A Gaussian process places a prior distribution over functions, \(f(x) \sim \mathcal{GP}(m(x), k (x, x'),\) where \(x \in \mathbf{R}^P\) is some input variable, the mean function \(m(x) = \mathbf{E}[f(x)]\), and the covariance function \(k(x, x') = \mathnormal{cov}(f(x), f(x') )\).
+A Gaussian process places a prior distribution over functions, \(f(x) \sim \mathcal{GP}(m(x), k (x, x'))\), where \(x \in \mathbf{R}^P\) is some input variable, the mean function \(m(x) = \mathbf{E}[f(x)]\), and the covariance function \(k(x, x') = \mathrm{cov}(f(x), f(x') )\).
 Given observations \(\mathbf{y}\) at locations \(\{x_n\}_{n=1}^N\), Gaussian noise, and kernel hyperparameters \(\theta\), a posterior predictive distribution over functions can be inferred analytically.
-See Rasmussen \& Williams\cite{rasmussen2006gaussian} for an in depth treatment.
+See Rasmussen \& Williams \cite{rasmussen2006gaussian} for an in depth treatment.
 
-Jones \& Moriarty\cite{jones2013evolutionary} extend GPs for the inference of function-valued traits over a phylogeny.
+Jones \& Moriarty \cite{jones2013evolutionary} extend GPs for the inference of function-valued traits over a phylogeny.
 Consider \(\mathcal{E}^{\mathbf{t}}(\cdot)\), a functional representation of the echolocation call of the species observed at the point \(\mathbf{t}\) on the phylogenetic tree \(\mathbf{T}\) with respect to frequency.
@@
 
-The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis et al.
-\cite{hadjipantelis2013function} \(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott\cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for the observed traits.
+The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis et al \cite{hadjipantelis2013function}.
+\(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott \cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for the observed traits.
 Given \(\hat{\phi}_F\), the estimated matrix of mixing coefficients is \(\hat{X}_{\mathbf{T}} = \mathcal{E} (\hat{\phi}_F^{\mathsf{T}})^{-1}\).
@@
 We note that the elements of \(\theta\) each have intuitive interpretations.
-The variance of observations in the sample is \(\sigma_p + \sigma_n\), where \(\sigma_p\) is the phylogenetic noise, and \(\sigma_n\) is the non-phylogenetic noise.
-\(\sigma_p\) is the proportion of variation which can be accounted for by the distance between any \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), while \(\sigma_n\) accounts for other sources of variation which may be observed at \(\mathbf{t}\).
-The length-scale parameter, \(\ell\) then indicates the strength of the correlation along \(\mathbf{T}\), where large values of \(\ell\) indicate a correlation that decays slowly as \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) increases.
+The total variation of observed points is \(\sigma_p + \sigma_n\), where \(\sigma_p\) is the phylogenetic noise, and \(\sigma_n\) is the non-phylogenetic noise.
+\(\sigma_p\) is the variation depending on the evolutionary time between any \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), while \(\sigma_n\) accounts for variation that does not depend on the phylogeny.
+The length-scale parameter, \(\ell\), indicates the strength of the correlation between points on \(\mathbf{T}\), where large values of \(\ell\) indicate a correlation that decays slowly as \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) increases.
 
 \section{Results}
-
+\label{sec:results}
 \subsection{Data Description}
 
-The post processed echolocation call data for bats found across north and central Mexico which accompanied Stathopoulos et al.
+Post processed echolocation call data accompanying Stathopoulos et al.
 \cite{stathopoulos2017bat} was used in this analysis.
-Live bats were caught, identified, and recorded either while released from the hand or while tied to a zip line.
-Calls were recorded at a sampling frequency of 500 kHz.
-In total the dataset consists of 22 species in five families, 449 individual bats and 1816 individual echolocation call recordings.
+Live bats were caught, identified, and recorded in at a sampling frequency of 500 kHz.
+In total the dataset consists of 22 species from five families, 449 individual bats and 1816 individual echolocation call recordings.
 The distribution of these call recordings across species is summarised in Table \ref{tab::dataset}.
 
-Collen's \cite{collen2012evolution} Bat super-tree provided the basis for the phylogenetic tree of the recorded bat species, \(\mathbf{T}\).
+Collen's \cite{collen2012evolution} Bat super-tree provided the phylogenetic tree of the recorded bat species, \(\mathbf{T}\).
 
@@
 
-We are interested in modelling the evolution of \(\mathcal{E}^{\mathbf{t}}(\cdot)\), the function valued trait representing \(x_{lm}^{\mathbf{t}}(\cdot)\).
+We are interested in modelling the evolution of \(\mathcal{E}^{\mathbf{t}}(\cdot)\), the function valued trait representing \(x_{lm}^{\mathbf{t}}(\cdot)\), with a phylogenetic GP.
 However, only 22 species of bat are represented in \(\mathbf{T}\).
 The relatively small size of this dataset presents challenges for the estimation of the kernel hyperparameters in (\ref{eqn:oukernel}).
-A short simulation study was performed to investigate the dynamics at play.
+A short simulation study was performed to investigate the the accuracy of estimated hyperparameters for a phylogenetic GP over \(\mathbf{T}\).
 
@@
 This simulation study indicates that at least \(n = 4\) observations are needed at each leaf node to provide stable estimators \(\hat{\theta}\).
-Thus, when implementing the phylogenetic OU process for bat echolocation call traits, resampling methods must be used to provide multiple estimates for \(\mathcal{E}^{\mathbf{t}}(\cdot)\)
 
@@
 
-Given the modelling assumptions made in Section \ref{sec:funrep} the best estimator for \(\mathcal{E}^{\mathbf{t}}(\cdot)\) is the sample mean given by 
+Given the modelling assumptions made in Section \ref{sec:funrep} an unbiased estimator for \(\mathcal{E}^{\mathbf{t}}(\cdot)\) is the sample mean given by 
 \begin{equation}
@@
 However, Table \ref{tab::ind_simulation} indicates that 22 samples is not enough to obtain a stable \(\hat{\theta}\) by type II maximum likelihood estimation.
-We implement a resampling procedure to leverage more estimates from the dataset in order to produce more stable \(\hat{\theta}\).
+We implement a resampling procedure to leverage multiple estimates for each \(\mathcal{E}^{\mathbf{t}}(\cdot)\) from the dataset.
+This will produce a stable estimator, \(\hat{\theta}\).
 
@@
 
-Our analysis took \(n_r = 4\) and included 4 samples of \(\hat{\mathcal{E}}_r^{\mathbf{t}}(\mathbf{f})\) in each \(\hat{\mathcal{E}}_r\).
+Our analysis set \(n_r = 4\) and included 4 samples of \(\hat{\mathcal{E}}_r^{\mathbf{t}}(\mathbf{f})\) in each \(\hat{\mathcal{E}}_r\) for \(r = 1,\dots,1000\).
 This reflected the structure of the dataset, for which the minimum number of individuals per species was 4, and the results of the simulations study which showed that 4 observations per species provided reasonably stable estimates for \(\theta\).
 Note also that \(\mathbf{f} = [9,10, \dots, 212]^{\mathsf{T}}\), which reflects the spectrum of frequencies over which bats emit echolocation calls.
-\(\hat{\phi}_F\) was obtained by averaging the basis identified over all \(\hat{\mathcal{E}}_r\) and applying a single set of basis functions to each dataset.
+\(\hat{\phi}_F\) was obtained by identifying the first six principal components, which accounted for approximately 85\% of the variation, in each  \(\hat{\mathcal{E}}_r\).
+By averaging over each sample, a single set of six stable, approximately orthogonal, basis functions were identified.
+These basis functions were then passed through Blaschke \& Wiscott's \cite{blaschke2004cubica} algorithm to produce a set of six independent basis functions for \(\mathcal{E}^\mathbf{t}(\cdot)\).
 Thus \(\hat{X}_r\), the matrix of mixing coefficients described by (\ref{eqn:inv}), the columns of which are modelled a phylogenetic OU processes, is obtained for each \(\hat{\mathcal{E}}_r\).
@@
 
-To perform ancestral trait reconstruction for \(\mathcal{E}^{\mathbf{t}^*}(\cdot)\) the species level traits are estimated by (\ref{eqn:est}) and and the model hyperparameters by the mean values of \(\theta_r\) reported in Table \ref{tab::bat_results}.
+To perform ancestral trait reconstruction for \(\mathcal{E}^{\mathbf{t}^*}(\cdot)\) the species level traits are estimated by (\ref{eqn:est}) and and the model hyperparameters are set to be the mean values of \(\theta_r\) reported in Table \ref{tab::bat_results}.
 
@@
 \section{Conclusions and Further Work}
-
-This preliminary analysis has developed a limited model for the evolution of echolocation in bats and identified a phylogenetic signal which allows the construction of a posterior predictive distribution for ancestral traits.
+\label{sec:conc}
+This preliminary analysis has developed a model for the evolution of echolocation in bats and identified a phylogenetic signal which allows the construction of a posterior predictive distribution for ancestral traits.
 The log energy spectral density has been identified as a trait representative of the echolocation call in bats.
 This trait, representing the energy intensity of the call across the frequency spectrum, is modelled as a series of independent components, combinations of energy intensities across the spectrum, each of which evolves according to a phylogenetic Ornstein-Uhlenbeck process.
-Estimating the hyperparameters governing these Ornstein-Uhlenbeck processes from observed traits provides an insight into the evolutionary path of these traits.
-Each of the hyperparameters has an intuitive interpretation where \(\frac{\sigma_p}{\sigma_p + \sigma_n}\) indicates the proportion of variation in the sample attributable solely to the evolutionary path while \(\ell\) provides a measure of how quickly correlation between observations along the evolutionary path decays.
+Estimating the hyperparameters governing these Ornstein-Uhlenbeck processes from observed traits provides an insight into the evolution of these traits.
+Each of the hyperparameters has an intuitive interpretation where \(\frac{\sigma_p}{\sigma_p + \sigma_n}\) indicates the proportion of variation in the sample accounted for by the phylogenetic distance between species, while \(\ell\) provides a measure of how quickly correlation along the phylogeny decays.
 We are working towards understanding what the results of this analysis could mean with respect to the evolution of echolocation in bats.
@@
 An alternative time-frequency representation, which preserves more of this structure, is the spectrogram.
-Modelling the evolution of bat echolocation calls with spectrograms is an active area of future research.
+Modelling the evolution of bat echolocation calls with spectrograms, and implementing this model for a larger dataset of bat echolocation calls, is to be the subject of future research.
 
+The interested reader can access the datasets and code used to produce these results through the R package 'sdsBAT' which is still under development and can be found at https://github.com/jpmeagher/sdsBAT.
 \bibliographystyle{ws-rv-van}

### 2017-06-28 Paper Drafts/Final/finalDraft.tex
Fixing final typos

@@
 \author[J.P.
-Meagher et al.]{J.P.
+Meagher \textit{et al.}]{J.P.
 Meagher\footnote{J.P Meagher would like to thank the EPSRC for funding this work, and also acknowledge the support of the Alan Turing Institute and Gaelic Players Association.}, T.
@@
 
-Monitoring bat populations provides useful information, but an understanding the orders evolutionary history is required to identify the cause and effect of any changes observed.
+Monitoring bat populations provides useful information, but an understanding of the evolutionary history is required to identify the cause and effect of any changes observed.
 The echolocation call structure, which reflects a bats diet and habitat \cite{aldridge1987morphology}, is a key aspect of this evolutionary history.
@@
 General techniques for the ancestral reconstruction of function-valued traits have also been proposed \cite{group2012phylogenetic}.
-Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits \cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis et al \cite{hadjipantelis2013function}.
+Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits \cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis \textit{et al.} \cite{hadjipantelis2013function}.
 This current research investigates these techniques in the context of the bat echolocation calls.
@@
 
-The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis et al \cite{hadjipantelis2013function}.
+The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis \textit{et al.} \cite{hadjipantelis2013function}.
 \(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott \cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for the observed traits.
@@
 
-Post processed echolocation call data accompanying Stathopoulos et al.
-\cite{stathopoulos2017bat} was used in this analysis.
-Live bats were caught, identified, and recorded in at a sampling frequency of 500 kHz.
+Post processed echolocation call data accompanying Stathopoulos \textit{et al.} \cite{stathopoulos2017bat} was used in this analysis.
+Live bats were caught, identified, and recorded at a sampling frequency of 500 kHz.
 In total the dataset consists of 22 species from five families, 449 individual bats and 1816 individual echolocation call recordings.

### 2017-07-02 cconferenceTalk.Rmd
Conference Talk and Slides

---
title: "Conference Talk"
author: "J.P. Meagher"
date: "1 July 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Title Slide (1 min 15 sec) (1 min 15 sec)

Good morning everyone, and thank you very much for that introduction _______ .

My name is Joe Meagher, and I am going to speak to you today about some of the research I have done so far for my PhD at the University of Warwick's Department of Statistics, titled "Phylogenetic Gaussian Processes for the Ancestral Reconstruction of Bat Echolocation Calls".

Over the next 25 minutes or so I will try to:

* Provide some background and motivation for modelling Bat echolocation calls.

* Give a gentle introduction to Ancestral Reconstruction and define function-valued traits in this context.

* Present results reducing models for the evolution of function-valued traits to Gaussian Process Regression.

* Explain how we applied these models to bat echolocation calls and discuss the results.

* Finally, I'll talk a little about where I see this work going in the future.

Before I go any further however, there are some organisations and people I want to thank.

# Acknowlegdements (1 min 15 sec) (2 min 30 sec)

I have been supported by a few organisations throughout my PhD.

* This work is funded by the EPSRC through the Engage project, which aims to enable non-programming scientists develop systems for detection in audio and visual data.

* The Alan Turing Institute has facilitated interdisciplinary collaboration on this project, particularly last summer.

* The Gaelic Players Association has also supported me, personally, during this research project.

I also want to thank my PhD supervisors, co-authors on this conference paper.

* Prof. Mark Girolami, Chair of Statistics at the Imperial and Director for the Alan Turing Institute-Lloyd’s Register Foundation Programme in Data-Centric Engineering.

* Dr. Theo Damoulas, an Assistant Professor in Data Science at Warwick and a Faculty Fellow at the ATI.

* Prof. Kate Jones, Chair of Ecology and Biodiversity at UCL, based at the Centre for Biodiversity and Environment Research.

# Background and Motivation (4 min) (6 min 30 sec)

So, why might one be interested in studying the evolution of echolocation in bats?

## Bat Background

Well, bats, as a whole, comprise the second most speciose order of mammals, after rodents.

There are over 1000 known species of bat, with about 17 species breeding in the UK & Ireland.

Bats are found on every continent, except Antartica, and they have successfully adapted to a massive variety of habitats and foraging strategies. The are also considered to be keystone species in desert and tropical environments. If any of you are unfamiliar with the notion of keystone species I recommend taking a look at a video on the impact of reintroducing wolves in Yellowstone. It presents a pretty incredible narrative.

What sets bats apart from other mammals is that they fly, in the dark most of the time, and even though they are not actually blind, most bats use echolocation to percieve the world around them.

Given all this, I think it's fair to say that bats are pretty remarkable creatures, and worthy of our attention, although its not difficult to convince people that this is the case.

## Public Engagement

Bats capture peoples imagination.

The appear consistently in folklore and mythology, from being symbols of good fortune in ancient China, to Bram Stoker's more sinister Count Dracula.

Research into bats generates public interest, Mark did a BBC television interview discussing some of the groups research and Kate supervised work, gathering the data used in this analysis actually, that featured in this BBC article.

This level of public engagement means that citizen science initiatives for gathering and labelling bat call data have been successful.

This brings me on to the specific motivation for wanting to reconstruct ancestral bat calls.

## Acoustic Biodiversity Monitoring

Bats have been identified as ideal Bioindicatiors for biodiversity monitoring.

Because they leak information about themselves into their environment through their echolocation calls, they can be monitored non-invasively using acoustic monitoring equipment. This picture shows the sort of monitoring device that can be used.

Ideally, these recordings will provide us with information about bats activity which can then be used as a proxy for the state of the environment in general. For this process to be really effective, a good understanding of bats natural history is required.

Historically, how have bats interacted with the flora and fauna in a given environment. Understanding the evolution of echolocation is a part of this.

## General Application of Techniques

The final aspect of motivation for this project that I am going to discuss, is the general applicability of these techniques.

The techniques used is this analysis have a lot in common with to those used for bat call and bird song classification.

There is also work being done to reconstruct latin phonetically from modern romance languages.

Phyologenetics is a big, active area of research.

I don't thing that it is too much of a stretch to think that insights uncovered in this research project could eventually feed into these other areas.

## Summary of Research Project

So, to summarise what we are hoping to achieve with this research project, let me refer you to this diagram.

We have audio recordings of the echolocation calls of various species of bat.

Assuming that these bats are related according to some phylogenetic tree, can we then compare and model the calls of existing bats in such a way that we can make some sensible statments about the calls of ancestral bats.

# Ancestral Reconstruction and Function-Valued Traits

Ok, so by now you probably have a fair idea what I mean when I say Ancestral Recostruction, but to be precise:

## Definition

Ancestral Reconstruction is the extrapolation back in time from measured characteristics of individuals (or populations) to their common ancestors.

It is an application of phylogenetics. Phylogenetics being the reconstruction and study of evolutionary relationships.

Some examples. Ancestral Reconstruction can be applied to genetic sequences, the amino acid sequence of a protien, or observed traits of an organism, an echolocation call, for example.

The results of an Ancestral Reconstruction study are heavily reliant on the modelling assumptions made. They are the phylogenetic tree of evolutionary relationships and also the model for the evolutionary dynamics, how the object of interest changes through evolutionary time.

## Evolutionary Models for Continuous Character Traits

There are a few approaches that can be taken to modelling continuous character traits, that is traits that vary along some continuous scale.

The most popular of these methods take the change in the trait over some time period to have a Gaussian distribution, which gives us Brownian Motion and Ornstein-Uhlenbeck models for the evolution of continuous character traits.

Now, Brownian Motion allows the trait to vary in an unrestricted way, evolution is essentially modelled as a random walk over the scale the trait is measured on. This is not necessarily the case in reality, as traits usually serve some purpose and so may have some optimal value which they vary around.

This is where The Ornstein-Uhlenbeck process for modelling stabilising selection comes into play. An Ornstein-Uhlenbeck process essentially behaves a Brownian Motion process with a tendency to some central value. This is more appropriate for modelling a trait with some optimal value which the trait then varies around, which is called stabilising selection.

And so what you see on this slide here is a simulation of an Ornstein Uhlenbeck Evolutionary Process over a given phylogenetic tree. The value at the root of the tree is 0 which is the value the trait varies around.

So these are the processes typically used to model the evolution of continuous character traits. However echolocation calls are not continuous character traits. So now I want to introduce the idea of a function-valued trait.

## Function-Valued Traits

A function-valued trait is a trait that is repeatedly measured, along some continuous scale, say time, where measurements can represent points on a curve.

Typical examples are growth curves, where the size of an individual is measured repeatedly over their lifetime.

For a function-valued trait, the mean for the trait over the space in which it is measured can vary, but this variation is both gradual and continuous. The same is true for the covariance of the trait.

# Modelling the Evolution of Function Valued Traits

What I have just described, in statistical terms, is a functional data object, and fortunately for me there has been a lot of work done in the area of Functional Data Analysis which informs our analysis of function valued traits.

In fact the machinery for applying Brownian Motion and Ornstein Uhlenbeck models for evolution to function-valued traits has already been developed.

## Phylogenetic Gaussian Process Regression

Firstly, I need to introduce Gaussian Processes.

Formally, a Gaussian process is a collection of random variables, any finite number of which have a joint Gaussian distribution.

A Gaussian Process places a prior distribution over functions, and so for a p dimensional input space, we can define a mean function, which is just the expected value of the functions, and a kernel which captures the covariance between points in the input space. The kernel. and the hyperparameters embeddid within it, defines the structure we place on the observed functions.

Nick Jones and John Moriarty, of Imperial and Manchester respectively extended Gaussian Processes to Phylogenies, which allows Evolutionary Inference on function-valued traits.

## Assumptions

For a phylogenetic Gaussian Process, the input space, \(\mathbf{x}\), becomes the Phylogenetic tree, and so the distance between observations becomes the distance between points in evolutionary time.

So, now we have observations which exist in the trait space and evolutionary time.

In order to make work in this context we need to make some simplifying assumptions.

Firstly, traits are conditionally independent given their common ancestors, which seems reasonable.

The second assumption states that the process governing the evolutionary dynamics are the same along every individual branch of the phylogenetic tree. This assumption is less reasonable and may be something you would look at relaxing as the analysis develops.

The third assumption just says that kernel of the marginal Gaussian process is space time separable. The marginal process being the process from an ancestor to its direct descendant.

With these assumptions in place Jones and Moriarty prove a few results, the most important of those for this analysis being this one here.

## Theoretical Results

What this result states is that a phylogenetic Gaussian process which is time-space separable is equivalent to the sum of univariate Gaussian Processes times their corresponding deterministic basis function, which we could also call a component.

Therefore, if we can define a set of basis functions our evolutionary inference over a function-valued trait reduces to the sum of \(n\) independent univariate Gaussian processes, which are the same processes that are used for modelling continuous character state evolution.

So now we can go ahead and model the evolution of our function-valued trait.

# Application to Bat Echolocation Calls

## The Data

The actual calls used in this analysis were recorded in north and central Mexico and they relate to 449 individuals from 22 species. Altogether we have just over 1800 individual call recordings and these are not necessarily spread uniformly across the sample with respect to individuals or species.

The dataset aimed for 100 individual calls per species, but some some species only had 4 idividual bats recorded so this means the number of calls per individual bat ranges from 1 to 40. This structure of the data was something we were wary of in the analysis.

We are not trying to infer the phylogenetic tree itself here, and I'm not convinced that using echolocation calls to do that would be a good idea, although maybe it's something we will revisit in the future. For now, we assume a phylogenetic tree for the recorded bats is the same as bat super tree published in 2012. that is what you can see on the slide.

## Echolocation Calls as Function-Valued Traits

Now, as regards the function-valued traits, the echolocation calls are repeated measurements of a trait the varies over a continuous scale, one of which is plotted here. However if someone has any ideas on how to compare acoustic waveforms directly for this sort of analysis I would be delighted to hear from them.

In our analysis we transformed the echolocation calls into the frequency domain with a Fourier transform. This gave us the energy spectral density of the call, which I will probably refer to as the spectral density from now on.

This gives us the energy intensity of the call along the frequency spectrum.

Finally, we smoothed each spectral density with smoothing splines and restricted each one to the 9-212 kHz frequency band, as this is the range of frequencies bats use for their echolocation calls.

These smoothed restricted spectral densities are the function valued traits we modelled.

## Independent Basis Functions

So now that we have a sample of function-valued traits, we needed to identify the deterministic basis functions that the independent Gaussian processes would act upon.

Now I dont have time to deal with the specifics of what we did but the details are covered in the conference paper.

Broadly speaking, we set up a resampling procedure to aviod bias in the basis functions identified. We then found Independent Principal Components from the resampled datasets. The basis functions identified can be seen here.

Our interpretation of these basis functions is that they reflect concentrations of energy at various bandwiths and identify the the frequency band used by different species for their echolocation calls.

Once the set of basis functions was identified we could then move on to inferring each of the independent univariate Gaussian processes.

# Results

## Kernel

We assumed that each phylogenetic Gaussian process was an Ornstein-Uhlenbeck process. We felt this was appropriate because bat echolocation calls seem to exhibit stabilising selection, as they can be thought of as serving a very specific purpose, which does in a sense have an optimal value.

The kernel for such a process is given here, where the hyperparameters arethe phylogenetic noise, the phylogenetic length scale, and the non-phylogenetic noise.

Again, without going into detail, a resampling procedure was set up and Type II maximum likelihood estimation used to estimate the hyperparameters of the phylogenetic Ornstein-Uhlenbeck Process. The results of which I have reported here.

## Model Hyperparameters

Now in order to interpret these results let me draw you attention to the third basis function.

If we refer back to our set of basis we can see that this represents a high concentration of energy at lower frequencies.

The next point I want to draw your attention to is the length scale hyperparameter \(\ell\). This can be interpreted as millions of years in evolutionary time over which the phylogenetic signal decays. Given the phylogenetic tree only goes back 50 million year, a lenght-scale of 70 indicates quite a strong phylogenetic signal.

Now I'd like to consider the noise hyperparameters, those for phylogenetic and non-phylogenetic noise. I thing what is important here is the ratio of these hyperparameters, which indicates what proportion of the variance in the sample is due to phylogenetic factors.

In this case more than 3/4 of the variation is attributable to phylogenetic factors.

So having inferred the model hyperparameters, we can now perform our Ancestral Reconstruction.

## Posterior Predictive Distributions

Ancestral resconstruction in this context reduces to finding the posterior predictive distribution for internal nodes of the tree.

What I'm showing you here is an ancestral reconstruction for the spectral density of the echolocation call for the most recent common ancestor of 3 extant species of mexican bat.

# Conclusions

So what can we conclude from this study.

Well we have identified a phylogenetic signal in the data, as one would hope.

More specifically, we have identified a strong phylogenetic signal at low frequencies, indicating that ancestral bats had lower frequency calls than their descendants.

To be honest, there is more work to do in interpreting the output of this model but it is a preliminary analysis, and I personally am more interested is seeing what the next iteration of this model tells us.

# Future Research Directions

Bat calls do in fact have a complex temporal structure and so spectral densities is not a good way of characterising them. A more usual method is to use a spectrogram, shown here.

The first extension to make to this model is to implement it for call spectrograms rather than spectral densities. We hope to complete this over the coming months.

Looking beyond this, we may examine other representations of the echolocation calls, such as Multi-component separation techniques.

Also the model for evolution used could also be improved, using Stable models for evolution seem like a promising route to consider.

That about covers everything I wanted to say. I'd just like thank Winton and Imperial for giving me the chance to present my work, I hope you enjoyed it and I'm happy to try answer any questions you may have.

Thank you.
